---
title: "Maladie à Virus Ebola — Situation Report Extraction"
subtitle: "Democratic Republic of Congo"
author: "Billy J Quilty (Charite Berlin, LSHTM & MSF Epicentre)"
date: today
date-format: "DD MMMM YYYY"
format:
  html:
    toc: true
    toc-depth: 2
    toc-title: "Contents"
    theme: flatly
    embed-resources: true
    fig-align: center
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")
pacman::p_load(tidyverse, lubridate, here, scales, ggrepel, flextable, ggh4x)

set_flextable_defaults(
  font.family = "Arial",
  font.size   = 10,
  border.color = "#CCCCCC"
)
```

```{r}
#| label: load-data
df_raw <- read_csv(
  here::here("outputs", "master_combined_counts.csv"),
  col_types       = cols(.default = "c"),
  show_col_types  = FALSE
)

df <- df_raw |>
  mutate(
    date        = dmy(count_end_date),
    suspect     = as.numeric(cases_suspect),
    probable    = as.numeric(cases_probable),
    confirmed   = as.numeric(cases_confirmed),
    deaths      = coalesce(as.numeric(deaths_suspected), as.numeric(deaths_probable), as.numeric(deaths_confirmed)),
    deaths_susp = as.numeric(deaths_suspected),
    is_agg      = (toupper(is_aggregate) == "TRUE"),
    # Only sum to zero when at least one component is present; otherwise NA
    # (avoids a false zero for zones that simply weren't reported in that sitrep)
    unconfirmed = if_else(
      !is.na(suspect) | !is.na(probable),
      coalesce(suspect, 0L) + coalesce(probable, 0L),
      NA_real_
    )
  )

ZONE_LEVELS <- c(
  "Mongbwalu", "Rwampara", "Bunia", "Nyankunde", "Bambu",
  "Butembo", "Katwa", "Goma"
)

# Colorblind-safe palette — Wong (2011)
PAL <- c(
  "Suspect/probable cases" = "#56B4E9",
  "Confirmed cases"        = "#D55E00",
  "Outbreak deaths"        = "#404040"
)

int_breaks <- function(x) unique(floor(pretty(x)))

theme_sitrep <- function(base_size = 11) {
  theme_minimal(base_size = base_size) %+replace%
    theme(
      plot.title       = element_text(face = "bold", size = base_size + 2, hjust = 0),
      plot.subtitle    = element_text(size = base_size - 1, colour = "grey45", hjust = 0,
                                      margin = margin(b = 8)),
      plot.caption     = element_text(size = base_size - 3, colour = "grey55", hjust = 1,
                                      margin = margin(t = 8)),
      plot.margin      = margin(12, 16, 8, 12),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(colour = "grey92"),
      legend.position  = "bottom",
      legend.key.size  = unit(0.75, "lines"),
      legend.text      = element_text(size = base_size - 1),
      strip.text       = element_text(face = "bold", size = base_size - 1,
                                      margin = margin(b = 4)),
      axis.title       = element_text(size = base_size - 1, colour = "grey35"),
      axis.text        = element_text(size = base_size - 2, colour = "grey25")
    )
}

# ── Derived subsets ────────────────────────────────────────────────────────────
totals <- df |>
  filter(count_type == "Cumules", is_agg, zone == "Total") |>
  group_by(date) |>
  slice_max(order_by = unconfirmed, n = 1, with_ties = FALSE) |>
  ungroup() |>
  arrange(date) |>
  mutate(deaths = cummax(replace_na(deaths, 0L)))

zone_ts <- df |>
  filter(count_type == "Cumules", !is_agg, zone %in% ZONE_LEVELS) |>
  group_by(date, zone, province) |>
  slice_max(order_by = unconfirmed, n = 1, with_ties = FALSE) |>
  ungroup() |>
  mutate(zone = factor(zone, levels = ZONE_LEVELS)) |>
  arrange(zone, date) |>
  group_by(zone) |>
  mutate(
    # Flag raw downward revisions before enforcing monotonicity
    confirmed_raw   = confirmed,
    unconfirmed_raw = unconfirmed,
    deaths_raw      = deaths,
    deaths_susp_raw = deaths_susp,
    rev_confirmed   = confirmed   < lag(cummax(replace_na(confirmed,   0L)), default = 0L),
    rev_unconfirmed = unconfirmed < lag(cummax(replace_na(unconfirmed, 0L)), default = 0L),
    # Apply cummax only over observed values; leave NAs before the first
    # real observation so geom_line breaks rather than drawing from zero
    unconfirmed = {x <- unconfirmed; i <- !is.na(x); if (any(i)) x[i] <- cummax(x[i]); x},
    confirmed   = {x <- confirmed;   i <- !is.na(x); if (any(i)) x[i] <- cummax(x[i]); x},
    deaths      = {x <- deaths;      i <- !is.na(x); if (any(i)) x[i] <- cummax(x[i]); x},
    deaths_susp = {x <- deaths_susp; i <- !is.na(x); if (any(i)) x[i] <- cummax(x[i]); x}
  ) |>
  ungroup()

# Warn about any downward revisions that cummax suppressed
revisions <- zone_ts |>
  filter(rev_confirmed | rev_unconfirmed) |>
  transmute(
    zone, date,
    confirmed_raw, confirmed,
    unconfirmed_raw, unconfirmed
  )
if (nrow(revisions) > 0) {
  message("NOTE: cummax suppressed ", nrow(revisions),
          " downward revision(s) in cumulative counts:\n",
          paste(capture.output(print(revisions)), collapse = "\n"))
}

zone_new <- df |>
  filter(count_type == "Nouveaux", !is_agg, zone %in% ZONE_LEVELS) |>
  group_by(date, zone, province) |>
  slice_max(order_by = unconfirmed, n = 1, with_ties = FALSE) |>
  ungroup() |>
  mutate(zone = factor(zone, levels = ZONE_LEVELS)) |>
  arrange(zone, date)

daily_new <- df |>
  filter(count_type == "Nouveaux", is_agg, zone == "Total") |>
  group_by(date) |>
  slice_max(order_by = unconfirmed, n = 1, with_ties = FALSE) |>
  ungroup() |>
  arrange(date)

make_zone_labels <- function(dat) {
  dat |>
    mutate(
      zone_label = paste0(zone, " (", province, ")"),
      zone_label = factor(zone_label,
        levels = dat |>
          distinct(zone, province) |>
          arrange(zone) |>
          mutate(lbl = paste0(zone, " (", province, ")")) |>
          pull(lbl)
      )
    )
}

# ── Inline summary statistics ──────────────────────────────────────────────────
latest_total <- totals |> slice_max(date, n = 1)
latest_date  <- format(max(totals$date, na.rm = TRUE), "%d %B %Y")
n_unconf     <- latest_total$unconfirmed
n_conf       <- latest_total$confirmed
n_deaths     <- coalesce(latest_total$deaths, 0L)
n_total      <- n_unconf + n_conf
cfr_pct      <- if (n_total > 0) round(n_deaths / n_total * 100, 1) else NA_real_
n_zones      <- n_distinct(zone_ts$zone)
n_provinces  <- n_distinct(zone_ts$province)
```

```{r}
#| label: load-kraemer
# Kraemer lab canonical name → this extraction zone name
KRAEMER_NAME_MAP <- c("Mongbalu" = "Mongbwalu", "Nyakunde" = "Nyankunde")

read_kraemer_long <- function(filename) {
  read_csv(
    here::here("Ebola_DRC_2026", "build", "long", filename),
    col_types      = cols(.default = "c"),
    show_col_types = FALSE
  )
}

kraemer <- read_kraemer_long("insp_sitrep__cumulative_confirmed_cases.csv") |>
  rename(confirmed = cumulative_confirmed_cases) |>
  full_join(
    read_kraemer_long("insp_sitrep__cumulative_suspected_cases.csv") |>
      rename(suspected = cumulative_suspected_cases),
    by = c("nom", "date")
  ) |>
  full_join(
    read_kraemer_long("insp_sitrep__cumulative_suspected_deaths.csv") |>
      rename(deaths_susp = cumulative_suspected_deaths),
    by = c("nom", "date")
  ) |>
  full_join(
    read_kraemer_long("insp_sitrep__cumulative_confirmed_deaths.csv") |>
      rename(deaths_conf = cumulative_confirmed_deaths),
    by = c("nom", "date")
  ) |>
  mutate(
    zone       = recode(nom, !!!KRAEMER_NAME_MAP),
    date       = as.Date(date),
    confirmed  = suppressWarnings(as.numeric(confirmed)),
    suspected  = suppressWarnings(as.numeric(suspected)),
    deaths_susp = suppressWarnings(as.numeric(deaths_susp)),
    deaths_conf = suppressWarnings(as.numeric(deaths_conf)),
    # Total deaths = suspected + confirmed (mirrors our coalesce logic)
    deaths     = coalesce(deaths_susp, 0) + coalesce(deaths_conf, 0),
    deaths     = if_else(is.na(deaths_susp) & is.na(deaths_conf), NA_real_, deaths),
    source     = "Kraemer lab"
  ) |>
  select(zone, date, confirmed, suspected, deaths_susp, deaths_conf, deaths, source) |>
  filter(zone %in% ZONE_LEVELS)

kraemer_hosp <- read_kraemer_long("insp_sitrep__hospitalised.csv") |>
  mutate(
    zone         = recode(nom, !!!KRAEMER_NAME_MAP),
    date         = as.Date(date),
    hospitalised = suppressWarnings(as.numeric(hospitalised))
  )

kraemer_contacts <- read_kraemer_long("insp_sitrep__cumulative_contacts_traced.csv") |>
  mutate(
    zone            = recode(nom, !!!KRAEMER_NAME_MAP),
    date            = as.Date(date),
    contacts_traced = suppressWarnings(as.numeric(cumulative_contacts_traced))
  ) |>
  filter(!is.na(contacts_traced))

# Response indicators (hospitalisations by zone)
kraemer_response <- read_kraemer_long("insp_sitrep__hospitalised.csv") |>
  full_join(read_kraemer_long("insp_sitrep__in_bed_previous_day.csv"), by = c("nom", "date")) |>
  full_join(
    read_kraemer_long("insp_sitrep__new_hosp_admissions.csv") |>
      rename(total_admissions = new_all_admissions),
    by = c("nom", "date")
  ) |>
  full_join(
    read_kraemer_long("insp_sitrep__new_hosp_detainees.csv") |>
      rename(new_admissions_ppl = new_hosp_detainees),
    by = c("nom", "date")
  ) |>
  full_join(
    read_kraemer_long("insp_sitrep__new_hosp_other.csv") |>
      rename(new_admissions_other = new_other),
    by = c("nom", "date")
  ) |>
  mutate(
    zone   = recode(nom, !!!KRAEMER_NAME_MAP),
    date   = as.Date(date),
    across(c(hospitalised, in_bed_previous_day, total_admissions,
             new_admissions_ppl, new_admissions_other),
           ~suppressWarnings(as.numeric(.x))),
    source = "Kraemer lab"
  ) |>
  filter(zone %in% ZONE_LEVELS)

# POE (checkpoint-level in Kraemer; summed to daily aggregate for comparison)
kraemer_poe <- read_kraemer_long("insp_sitrep__total_poe_passed.csv") |>
  rename(total_passed = total_poe_passed) |>
  full_join(
    read_kraemer_long("insp_sitrep__total_poe_screened.csv") |>
      rename(total_screened = total_poe_screened),
    by = c("nom", "date")
  ) |>
  full_join(
    read_kraemer_long("insp_sitrep__total_poe_hand_washing.csv") |>
      rename(total_handwashing = total_poe_hand_washing),
    by = c("nom", "date")
  ) |>
  full_join(
    read_kraemer_long("insp_sitrep__total_poe_sanitised.csv") |>
      rename(total_sensitised = total_poe_sanitised),
    by = c("nom", "date")
  ) |>
  mutate(
    date = as.Date(date),
    across(c(total_passed, total_screened, total_handwashing, total_sensitised),
           ~suppressWarnings(as.numeric(.x)))
  ) |>
  group_by(date) |>
  summarise(
    across(c(total_passed, total_screened, total_handwashing, total_sensitised),
           ~sum(.x, na.rm = TRUE)),
    .groups = "drop"
  ) |>
  mutate(source = "Kraemer lab")
```

## Introduction

This report summarises the 2026 Ebola (MVE) outbreak in the Democratic Republic of Congo, covering cases in Ituri and Nord-Kivu provinces. Data are drawn from official situation reports published by the Institut National de Santé Publique (INSP DRC) and are updated automatically as new reports become available.

## Methods

### Data sources

Case counts were taken from the INSP DRC situation report PDFs, downloaded automatically from <https://insp.cd/ebola/>. We verified that our copies of each PDF are identical to those used by the Kraemer lab (by comparing file checksums), so any differences between the two datasets come from how the tables were read, not which documents were used. The Kraemer lab series, produced by manual data entry, was used as an independent check on our automated extraction.

### Automated PDF extraction

Each PDF was read by Claude (Anthropic, `claude-sonnet-4-6`), which looks at the page visually — the same way a person would — rather than relying on the underlying text. This allows it to handle the complex table layouts in the situation reports accurately. From each PDF we extract four things: daily new case counts, cumulative case and death totals, patient movement through isolation units (available from SitRep 006 onwards), and Point of Entry screening activity. A full text transcript of each report is also saved for audit purposes.

### Data harmonisation

Health zone names were standardised to a consistent spelling across all reports. Suspected and probable case counts were added together into a single "unconfirmed" category, since different reports label these columns differently. Cumulative totals were not allowed to decrease over time (a count cannot go backwards), and where more than one row existed for the same zone and date, the higher value was kept.

### Statistical analysis

All analysis and figures were produced in R (version `r R.version$major`.`r R.version$minor`) using the `tidyverse` [@tidyverse], `ggplot2` [@ggplot2], and `flextable` [@flextable] packages. This report was rendered automatically with Quarto.

## Results

As of `r latest_date`, the DRC MVE outbreak had recorded a cumulative total of `r prettyNum(n_total, big.mark = ",")` cases (`r prettyNum(n_conf, big.mark = ",")` confirmed, `r prettyNum(n_unconf, big.mark = ",")` suspect or probable) and `r prettyNum(n_deaths, big.mark = ",")` deaths, corresponding to a case fatality ratio of `r cfr_pct`%. Cases have been reported across `r n_zones` health zones in `r n_provinces` provinces.

### Cumulative epidemic curve

```{r}
#| label: fig-cumulative
#| fig-cap: "Cumulative MVE cases and deaths, DRC. Suspect and probable cases are combined. End-of-line labels show the latest recorded value. Source: INSP DRC."
#| fig-width: 8
#| fig-height: 5
p1_dat <- totals |>
  select(date,
         `Suspect/probable cases` = unconfirmed,
         `Confirmed cases`        = confirmed,
         `Outbreak deaths`        = deaths) |>
  pivot_longer(-date, names_to = "category", values_to = "count") |>
  mutate(category = factor(category, levels = names(PAL)))

p1_last <- p1_dat |>
  filter(!is.na(count)) |>
  group_by(category) |>
  slice_max(date, n = 1) |>
  ungroup()

p <- ggplot(p1_dat, aes(x = date, y = count, colour = category, group = category)) +
  geom_line(linewidth = 1.1, na.rm = TRUE) +
  geom_label_repel(
    data           = p1_last,
    aes(label      = count),
    size           = 3.5,
    nudge_x        = 0.8,
    direction      = "y",
    hjust          = 0,
    segment.colour = "grey70",
    segment.size   = 0.3,
    show.legend    = FALSE,
    na.rm          = TRUE
  ) +
  scale_colour_manual(values = PAL) +
  scale_x_date(
    date_labels = "%d %b",
    date_breaks = "2 days",
    expand      = expansion(add = c(0, 1.2))
  ) +
  scale_y_continuous(breaks = int_breaks, expand = expansion(mult = c(0, 0.08))) +
  labs(
    title    = "MVE DRC \u2014 Cumulative epidemic curve",
    subtitle = "Total counts across all health zones",
    x = NULL, y = "Cumulative count", colour = NULL,
    caption  = "Source: INSP DRC situation reports"
  ) +
  theme_sitrep(base_size = 12)
ggsave(here("outputs", "plots", "fig-cumulative.png"), plot = p, width = 8, height = 5, dpi = 300)
p
```

### Daily new cases

```{r}
#| label: fig-new-cases
#| fig-cap: "New cases reported per situation report across all health zones, DRC. Source: INSP DRC."
#| fig-width: 7
#| fig-height: 4
p <- daily_new |>
  select(date,
         `Suspect/probable cases` = unconfirmed,
         `Confirmed cases`        = confirmed) |>
  pivot_longer(-date, names_to = "category", values_to = "count") |>
  mutate(category = factor(category, levels = c("Suspect/probable cases", "Confirmed cases"))) |>
  ggplot(aes(x = date, y = count, fill = category)) +
  geom_col(position = "dodge", width = 0.3, na.rm = TRUE) +
  scale_fill_manual(values = PAL) +
  scale_x_date(date_labels = "%d %b", date_breaks = "1 day") +
  scale_y_continuous(breaks = int_breaks, expand = expansion(mult = c(0, 0.1))) +
  labs(
    title    = "MVE DRC \u2014 Daily new cases",
    subtitle = "Total reported per situation report",
    x = NULL, y = "New cases", fill = NULL,
    caption  = "Source: INSP DRC situation reports"
  ) +
  theme_sitrep(base_size = 12)
ggsave(here("outputs", "plots", "fig-new-cases.png"), plot = p, width = 7, height = 4, dpi = 300)
p
```

### Geographic distribution — cumulative counts

```{r}
#| label: fig-zone-cumulative
#| fig-cap: "Cumulative cases by health zone. Y-axis is free across panels. Source: INSP DRC."
#| fig-width: 10
#| fig-height: 11
p <- zone_ts |>
  make_zone_labels() |>
  select(date, zone_label,
         `Suspect/probable cases` = unconfirmed,
         `Confirmed cases`        = confirmed,
         `Outbreak deaths`        = deaths) |>
  pivot_longer(
    cols      = c(`Suspect/probable cases`, `Confirmed cases`, `Outbreak deaths`),
    names_to  = "category",
    values_to = "count"
  ) |>
  mutate(category = factor(category, levels = names(PAL))) |>
  ggplot(aes(x = date, y = count, colour = category, group = category)) +
  geom_line(linewidth = 0.9, na.rm = TRUE) +
  scale_colour_manual(values = PAL) +
  scale_x_date(date_labels = "%d %b", date_breaks = "1 day") +
  scale_y_continuous(breaks = int_breaks, expand = expansion(mult = c(0, 0.15))) +
  facet_wrap(~ zone_label, scales = "free_y", ncol = 2) +
  labs(
    title    = "MVE DRC \u2014 Cumulative cases by health zone",
    subtitle = "Each panel shows one health zone; y-axis is free.",
    x = NULL, y = "Cumulative count", colour = NULL,
    caption  = "Source: INSP DRC situation reports"
  ) +
  theme_sitrep(base_size = 10) +
  theme(axis.text.x = element_text(size = 8, angle = 45, hjust = 1))
ggsave(here("outputs", "plots", "fig-zone-cumulative.png"), plot = p, width = 10, height = 11, dpi = 300)
p
```

### Geographic distribution — new cases

```{r}
#| label: fig-zone-new
#| fig-cap: "New cases per situation report by health zone. Y-axis is free across panels. Source: INSP DRC."
#| fig-width: 10
#| fig-height: 11
p <- zone_new |>
  make_zone_labels() |>
  select(date, zone_label,
         `Suspect/probable cases` = unconfirmed,
         `Confirmed cases`        = confirmed) |>
  pivot_longer(
    cols      = c(`Suspect/probable cases`, `Confirmed cases`),
    names_to  = "category",
    values_to = "count"
  ) |>
  mutate(category = factor(category, levels = c("Suspect/probable cases", "Confirmed cases"))) |>
  ggplot(aes(x = date, y = count, fill = category)) +
  geom_col(position = "dodge", width = 0.4, na.rm = TRUE) +
  scale_fill_manual(values = PAL) +
  scale_x_date(date_labels = "%d %b", date_breaks = "1 day") +
  scale_y_continuous(breaks = int_breaks, expand = expansion(mult = c(0, 0.15))) +
  facet_wrap(~ zone_label, scales = "free_y", ncol = 2) +
  labs(
    title    = "MVE DRC \u2014 Daily new cases by health zone",
    subtitle = "Per situation report; y-axis is free.",
    x = NULL, y = "New cases", fill = NULL,
    caption  = "Source: INSP DRC situation reports"
  ) +
  theme_sitrep(base_size = 10) +
  theme(axis.text.x = element_text(size = 8, angle = 45, hjust = 1))
ggsave(here("outputs", "plots", "fig-zone-new.png"), plot = p, width = 10, height = 11, dpi = 300)
p
```

### Current counts by health zone

Table 1 presents the most recent cumulative case counts disaggregated by health zone, drawn from the latest available situation report for each zone.

```{r}
#| label: tbl-zones
#| tbl-cap: "Table 1. Latest cumulative case counts by health zone."
zone_latest <- zone_ts |>
  group_by(zone, province) |>
  slice_max(date, n = 1) |>
  ungroup() |>
  arrange(zone) |>
  transmute(
    `Health zone`      = as.character(zone),
    Province           = province,
    `Latest sitrep`    = format(date, "%d %b %Y"),
    `Suspect/probable` = unconfirmed,
    `Confirmed`        = confirmed,
    `Deaths`           = deaths
  )

total_row <- tibble(
  `Health zone`      = "Total",
  Province           = "\u2014",
  `Latest sitrep`    = format(max(totals$date, na.rm = TRUE), "%d %b %Y"),
  `Suspect/probable` = sum(zone_latest$`Suspect/probable`, na.rm = TRUE),
  `Confirmed`        = sum(zone_latest$`Confirmed`,        na.rm = TRUE),
  `Deaths`           = sum(zone_latest$`Deaths`,           na.rm = TRUE)
)

bind_rows(zone_latest, total_row) |>
  flextable() |>
  bold(part = "header") |>
  bold(i = nrow(zone_latest) + 1L) |>
  bg(part = "header", bg = "#1A3A5C") |>
  color(part = "header", color = "white") |>
  bg(i = nrow(zone_latest) + 1L, bg = "#E8EFF7") |>
  hline(i = nrow(zone_latest), part = "body") |>
  align(j = 4:6, align = "right", part = "all") |>
  colformat_int(j = 4:6) |>
  autofit()
```

### Cross-source validation

We checked our automated extraction against the [Kraemer lab Ebola DRC 2026](https://github.com/kraemer-lab/Ebola_DRC_2026) manually coded dataset, which is built from the same situation reports. Because both datasets work from identical PDFs, any differences between them tell us something about how the tables were interpreted — which is useful for catching errors on either side and understanding ambiguous formatting in the source documents. Two zone name variants between the datasets (Mongbwalu/Mongbalu and Nyankunde/Nyakunde) were matched manually. Values marked "ND" (not available) were treated as missing.

### Data quality notes — case counts

Two issues in the Mongbwalu data are worth noting. First, the early Kraemer data shows 246 suspected cases at Mongbwalu on 14 May. Our earliest report (SitRep 001) records the same 246 figure, but only as a total for the whole of Ituri province — it does not break the number down by health zone. The first report with a Mongbwalu-specific count in our data is SitRep 004 (18 May, 302 suspected cases). As a result, the Mongbwalu line in the figures below starts later than in the Kraemer data; this is a gap in zone-level reporting, not a sign that cases only began in May.

Second, both our data and the Kraemer data show a fall in cumulative confirmed cases at Mongbwalu between 21 May (14 cases) and 22 May (10 cases). Cumulative case counts should not decrease, so this is likely a correction or reclassification in the 22 May situation report rather than an error in how the data were extracted. In our figures the value is held at 14 on 22 May (since we do not allow counts to fall); the Kraemer series shows the raw figure.

```{r}
#| label: fig-comparison
#| fig-cap: "Cross-source comparison of cumulative confirmed cases, suspect/probable cases, and outbreak deaths by health zone. Solid lines: automated extraction (raw); dashed lines: Kraemer lab (manual extraction). Both automated and Kraemer sources use identical PDFs; divergence reflects extraction method differences."
#| fig-width: 12
#| fig-height: 14
PAL_COMP <- c(
  "Automated extraction" = "#0072B2",
  "Kraemer lab"          = "#E69F00"
)
LTY_COMP <- c(
  "Automated extraction" = "solid",
  "Kraemer lab"          = "dashed"
)

zone_province_lut <- zone_ts |>
  distinct(zone, province) |>
  mutate(zone = as.character(zone))

our_for_comp <- zone_ts |>
  transmute(zone = as.character(zone), province, date,
            confirmed   = confirmed_raw,
            suspected   = unconfirmed_raw,
            deaths_susp = deaths_susp_raw,
            deaths      = deaths_raw,
            source      = "Automated extraction")

# Kraemer deaths_susp = suspected only; deaths = suspected + confirmed (total)
kraemer_for_comp <- kraemer |>
  left_join(zone_province_lut, by = "zone") |>
  transmute(zone, province, date, confirmed, suspected,
            deaths_susp, deaths,
            source)

comp_long <- bind_rows(our_for_comp, kraemer_for_comp) |>
  pivot_longer(c(confirmed, suspected, deaths_susp, deaths),
               names_to = "metric", values_to = "count") |>
  mutate(
    metric = recode(metric,
      "confirmed"   = "Confirmed cases",
      "suspected"   = "Suspect/probable cases",
      "deaths_susp" = "Suspected deaths",
      "deaths"      = "Total deaths (coalesced)"
    ),
    metric   = factor(metric, levels = c("Confirmed cases", "Suspect/probable cases",
                                          "Suspected deaths", "Total deaths (coalesced)")),
    source   = factor(source, levels = c("Automated extraction", "Kraemer lab")),
    province = factor(province, levels = c("Ituri", "Nord-Kivu")),
    zone     = factor(zone, levels = ZONE_LEVELS)
  ) |>
  filter(!is.na(count))

p <- ggplot(comp_long, aes(x = date, y = count, colour = source, linetype = source)) +
  geom_line(linewidth = 0.9, na.rm = TRUE) +
  geom_point(data = ~filter(.x, source == "Kraemer lab"), size = 2.0, na.rm = TRUE) +
  facet_nested(metric ~ province + zone, scales = "free_y") +
  scale_colour_manual(values = PAL_COMP) +
  scale_linetype_manual(values = LTY_COMP) +
  scale_x_date(date_labels = "%d %b", date_breaks = "2 days") +
  scale_y_continuous(breaks = int_breaks, limits = c(0, NA), expand = expansion(mult = c(0, 0.15))) +
  labs(
    title    = "MVE DRC \u2014 Cross-source validation",
    subtitle = "Cumulative counts: automated extraction (raw) vs Kraemer lab (manual extraction, same source PDFs)",
    x = NULL, y = "Cumulative count", colour = NULL, linetype = NULL,
    caption  = "Sources: INSP DRC situation reports; github.com/kraemer-lab/Ebola_DRC_2026"
  ) +
  theme_sitrep(base_size = 9) +
  theme(
    axis.text.x     = element_text(angle = 45, hjust = 1, size = 7),
    panel.border    = element_rect(colour = "grey70", fill = NA, linewidth = 0.4),
    panel.spacing   = unit(0.6, "lines")
  )
ggsave(here("outputs", "plots", "fig-comparison.png"), plot = p, width = 12, height = 14, dpi = 300)
p
```

Table 2 presents numerical differences at the most recent date available in each source for each zone.

```{r}
#| label: tbl-comparison
#| tbl-cap: "Table 2. Cumulative case counts at the latest available date per zone: automated extraction vs Kraemer lab. Delta = ours minus Kraemer. Dashes indicate no data in that source."
ours_latest <- zone_ts |>
  group_by(zone) |>
  slice_max(date, n = 1) |>
  ungroup() |>
  transmute(zone, conf_ours = confirmed, susp_ours = unconfirmed, dead_ours = deaths)

kraemer_latest <- kraemer |>
  group_by(zone) |>
  slice_max(date, n = 1) |>
  ungroup() |>
  transmute(zone, conf_them = confirmed, susp_them = suspected, dead_them = deaths)

full_join(ours_latest, kraemer_latest, by = "zone") |>
  mutate(
    d_conf = conf_ours - conf_them,
    d_susp = susp_ours - susp_them,
    d_dead = dead_ours - dead_them
  ) |>
  arrange(zone) |>
  transmute(
    zone,
    conf_ours, conf_them, d_conf,
    susp_ours, susp_them, d_susp,
    dead_ours, dead_them, d_dead
  ) |>
  setNames(c(
    "Health zone",
    "Confirmed (ours)", "Confirmed (Kraemer)", "\u0394 confirmed",
    "Suspect/prob. (ours)", "Suspect/prob. (Kraemer)", "\u0394 suspect/prob.",
    "Deaths (ours)", "Deaths (Kraemer)", "\u0394 deaths"
  )) |>
  flextable() |>
  bold(part = "header") |>
  bg(part = "header", bg = "#1A3A5C") |>
  color(part = "header", color = "white") |>
  align(j = 2:10, align = "right", part = "all") |>
  colformat_int(j = 2:10, na_str = "\u2014") |>
  autofit()
```

### Response indicators

```{r}
#| label: load-response-ours
resp_ours <- read_csv(
  here::here("outputs", "master_response_counts.csv"),
  col_types      = cols(.default = "c"),
  show_col_types = FALSE
) |>
  mutate(
    date   = dmy(date),
    zone   = factor(zone, levels = ZONE_LEVELS),
    across(c(in_bed_previous_day, new_admissions_ppl, new_admissions_other,
             total_admissions, total_isolated, cumulative_admitted),
           ~suppressWarnings(as.numeric(.x))),
    source = "Automated extraction"
  ) |>
  filter(!is.na(zone), !zone %in% c("Total"))

poe_ours <- read_csv(
  here::here("outputs", "master_poe_counts.csv"),
  col_types      = cols(.default = "c"),
  show_col_types = FALSE
) |>
  mutate(
    date = dmy(date),
    across(c(total_passed, total_screened, total_handwashing, total_sensitised),
           ~suppressWarnings(as.numeric(.x))),
    source = "Automated extraction"
  ) |>
  select(date, total_passed, total_screened, total_handwashing, total_sensitised, source)
```

### Data quality notes — response indicators

Two points are worth flagging when comparing the response indicator figures. For Rwampara on 21 May, SitRep 007 splits the data across three sub-columns without explaining what each represents. Both datasets handle this differently, so the 21 May Rwampara figures should be treated with caution. For 20–21 May more broadly, SitReps 005 and 006 used an earlier table layout that made the in-bed count and new admissions columns harder to distinguish; the two datasets differ slightly on those dates as a result. From 22 May onwards, when the table format was standardised, the datasets are in close agreement.

```{r}
#| label: fig-comparison-response
#| fig-cap: "Cross-source comparison of zone-level response indicators. Solid lines: automated extraction; dashed lines: Kraemer lab. Both series treat the ambiguous 21 May Rwampara sub-columns differently; values for that date should be interpreted cautiously. The two datasets differ slightly on 20–21 May due to an ambiguous table layout in SitReps 005–006, and are in close agreement from 22 May onwards."
#| fig-width: 12
#| fig-height: 9
PAL_RESP <- c("Automated extraction" = "#0072B2", "Kraemer lab" = "#E69F00")
LTY_RESP <- c("Automated extraction" = "solid",   "Kraemer lab" = "dashed")

resp_comp <- bind_rows(
  resp_ours |>
    transmute(date, zone,
              hospitalised        = total_isolated,
              in_bed_previous_day,
              total_admissions,
              source),
  kraemer_response |>
    transmute(date, zone = factor(zone, levels = ZONE_LEVELS),
              hospitalised,
              in_bed_previous_day,
              total_admissions,
              source)
) |>
  pivot_longer(c(hospitalised, in_bed_previous_day, total_admissions),
               names_to = "metric", values_to = "count") |>
  mutate(
    metric = recode(metric,
      "hospitalised"        = "Currently hospitalised",
      "in_bed_previous_day" = "In bed (prev. day)",
      "total_admissions"    = "New admissions (total)"
    ),
    metric = factor(metric, levels = c("Currently hospitalised",
                                        "In bed (prev. day)",
                                        "New admissions (total)")),
    source = factor(source, levels = c("Automated extraction", "Kraemer lab"))
  ) |>
  filter(!is.na(count), !is.na(zone))

p <- ggplot(resp_comp, aes(x = date, y = count, colour = source, linetype = source)) +
  geom_line(linewidth = 0.9, na.rm = TRUE) +
  geom_point(size = 2.0, na.rm = TRUE) +
  facet_nested(metric ~ zone, scales = "free_y") +
  scale_colour_manual(values = PAL_RESP) +
  scale_linetype_manual(values = LTY_RESP) +
  scale_x_date(date_labels = "%d %b", date_breaks = "2 days") +
  scale_y_continuous(breaks = int_breaks, limits = c(0, NA),
                     expand = expansion(mult = c(0, 0.15))) +
  labs(
    title    = "MVE DRC \u2014 Response indicators — cross-source validation",
    subtitle = "Automated extraction vs Kraemer lab (manual, same source PDFs)",
    x = NULL, y = "Count", colour = NULL, linetype = NULL,
    caption  = "Sources: INSP DRC situation reports; github.com/kraemer-lab/Ebola_DRC_2026"
  ) +
  theme_sitrep(base_size = 9) +
  theme(
    axis.text.x  = element_text(angle = 45, hjust = 1, size = 7),
    panel.border = element_rect(colour = "grey70", fill = NA, linewidth = 0.4),
    panel.spacing = unit(0.5, "lines")
  )
ggsave(here("outputs", "plots", "fig-comparison-response.png"), plot = p,
       width = 12, height = 9, dpi = 300)
p
```

```{r}
#| label: fig-comparison-poe
#| fig-cap: "Cross-source comparison of daily Point of Entry (POE) screening totals. Automated extraction reports a single aggregate per sitrep date; Kraemer lab data are summed across individual checkpoints to produce a daily total."
#| fig-width: 9
#| fig-height: 6
poe_comp <- bind_rows(poe_ours, kraemer_poe) |>
  pivot_longer(c(total_passed, total_screened, total_handwashing, total_sensitised),
               names_to = "metric", values_to = "count") |>
  mutate(
    metric = recode(metric,
      "total_passed"      = "Total passed",
      "total_screened"    = "Screened",
      "total_handwashing" = "Handwashing stations",
      "total_sensitised"  = "Sensitised"
    ),
    metric = factor(metric, levels = c("Total passed", "Screened",
                                        "Handwashing stations", "Sensitised")),
    source = factor(source, levels = c("Automated extraction", "Kraemer lab"))
  ) |>
  filter(!is.na(count))

p <- ggplot(poe_comp, aes(x = date, y = count, colour = source, linetype = source)) +
  geom_line(linewidth = 0.9, na.rm = TRUE) +
  geom_point(size = 2.5, na.rm = TRUE) +
  facet_wrap(~ metric, scales = "free_y", ncol = 2) +
  scale_colour_manual(values = PAL_RESP) +
  scale_linetype_manual(values = LTY_RESP) +
  scale_x_date(date_labels = "%d %b", date_breaks = "2 days") +
  scale_y_continuous(breaks = int_breaks, limits = c(0, NA),
                     expand = expansion(mult = c(0, 0.12))) +
  labs(
    title    = "MVE DRC \u2014 POE screening — cross-source validation",
    subtitle = "Daily aggregate totals: automated extraction vs Kraemer lab (sum across checkpoints)",
    x = NULL, y = "Count", colour = NULL, linetype = NULL,
    caption  = "Sources: INSP DRC situation reports; github.com/kraemer-lab/Ebola_DRC_2026"
  ) +
  theme_sitrep(base_size = 11) +
  theme(
    axis.text.x  = element_text(angle = 45, hjust = 1, size = 8),
    panel.border = element_rect(colour = "grey70", fill = NA, linewidth = 0.4),
    panel.spacing = unit(0.6, "lines")
  )
ggsave(here("outputs", "plots", "fig-comparison-poe.png"), plot = p,
       width = 9, height = 6, dpi = 300)
p
```

```{r}
#| label: fig-comparison-contacts
#| fig-cap: "Cross-source comparison of cumulative contacts traced by health zone. Solid lines: automated extraction (raw); dashed lines: Kraemer lab (manual extraction). ND values excluded."
#| fig-width: 11
#| fig-height: 7
contacts_ours <- df |>
  filter(
    count_type == "Cumules",
    zone %in% ZONE_LEVELS,
    !is.na(contacts), contacts != ""
  ) |>
  mutate(contacts = suppressWarnings(as.numeric(contacts)),
         zone = factor(zone, levels = ZONE_LEVELS)) |>
  filter(!is.na(contacts)) |>
  arrange(zone, date) |>
  group_by(zone) |>
  ungroup() |>
  transmute(date, zone, contacts, source = "Automated extraction")

contacts_kraemer <- kraemer_contacts |>
  filter(zone %in% ZONE_LEVELS) |>
  transmute(date, zone = factor(zone, levels = ZONE_LEVELS),
            contacts = contacts_traced, source = "Kraemer lab")

contacts_comp <- bind_rows(contacts_ours, contacts_kraemer) |>
  filter(!is.na(contacts)) |>
  left_join(zone_province_lut, by = "zone") |>
  mutate(
    source   = factor(source, levels = c("Automated extraction", "Kraemer lab")),
    province = factor(province, levels = c("Ituri", "Nord-Kivu")),
    zone     = factor(zone, levels = ZONE_LEVELS)
  )

p <- ggplot(contacts_comp,
            aes(x = date, y = contacts, colour = source, linetype = source)) +
  geom_line(linewidth = 0.9, na.rm = TRUE) +
  geom_point(data = ~filter(.x, source == "Kraemer lab"), size = 2.0, na.rm = TRUE) +
  facet_nested(~ province + zone, scales = "free_y") +
  scale_colour_manual(values = PAL_COMP[c("Automated extraction","Kraemer lab")]) +
  scale_linetype_manual(values = LTY_COMP[c("Automated extraction","Kraemer lab")]) +
  scale_x_date(date_labels = "%d %b", date_breaks = "2 days") +
  scale_y_continuous(breaks = int_breaks, limits = c(0, NA),
                     expand = expansion(mult = c(0, 0.15))) +
  labs(
    title    = "MVE DRC \u2014 Cumulative contacts traced — cross-source validation",
    subtitle = "Automated extraction (raw) vs Kraemer lab (manual extraction, same source PDFs)",
    x = NULL, y = "Cumulative contacts traced", colour = NULL, linetype = NULL,
    caption  = "Sources: INSP DRC situation reports; github.com/kraemer-lab/Ebola_DRC_2026"
  ) +
  theme_sitrep(base_size = 9) +
  theme(
    axis.text.x  = element_text(angle = 45, hjust = 1, size = 7),
    panel.border = element_rect(colour = "grey70", fill = NA, linewidth = 0.4),
    panel.spacing = unit(0.5, "lines")
  )
ggsave(here("outputs", "plots", "fig-comparison-contacts.png"), plot = p,
       width = 11, height = 7, dpi = 300)
p
```

## Discussion

Case counts have risen steadily over the reporting period. Confirmed cases remain a small fraction of all reported cases, which is typical in field settings where laboratory testing capacity is limited. The outbreak spans health zones in both Ituri and Nord-Kivu, pointing to the need for a coordinated response across provincial boundaries.

A few limitations are worth keeping in mind. Earlier situation reports did not break case counts down by health zone, so the zone-level data only covers part of the outbreak period. Response indicators such as hospitalisations and contacts traced are not formatted consistently across reports, which makes trend comparisons harder. Comparing our automated extraction with the Kraemer lab manual series provides a useful cross-check: where the two agree we can be more confident in the figures; where they differ, it usually points to genuine ambiguity in the source document that is worth flagging to the data providers.
