---
title: "Maladie à Virus Ebola — Situation Report"
subtitle: "Democratic Republic of Congo"
date: today
date-format: "DD MMMM YYYY"
format:
  html:
    toc: true
    toc-depth: 2
    toc-title: "Contents"
    theme: flatly
    embed-resources: true
    fig-align: center
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")
pacman::p_load(tidyverse, lubridate, here, scales, ggrepel, flextable, ggh4x)

set_flextable_defaults(
  font.family = "Arial",
  font.size   = 10,
  border.color = "#CCCCCC"
)
```

```{r}
#| label: load-data
df_raw <- read_csv(
  here::here("outputs", "master_combined_counts.csv"),
  col_types       = cols(.default = "c"),
  show_col_types  = FALSE
)

df <- df_raw |>
  mutate(
    date        = dmy(count_end_date),
    suspect     = as.numeric(cases_suspect),
    probable    = as.numeric(cases_probable),
    confirmed   = as.numeric(cases_confirmed),
    deaths      = coalesce(as.numeric(deaths_suspected), as.numeric(deaths_probable), as.numeric(deaths_confirmed)),
    deaths_susp = as.numeric(deaths_suspected),
    is_agg      = (is_aggregate == "TRUE"),
    # Only sum to zero when at least one component is present; otherwise NA
    # (avoids a false zero for zones that simply weren't reported in that sitrep)
    unconfirmed = if_else(
      !is.na(suspect) | !is.na(probable),
      coalesce(suspect, 0L) + coalesce(probable, 0L),
      NA_real_
    )
  )

ZONE_LEVELS <- c(
  "Mongbwalu", "Rwampara", "Bunia", "Nyankunde", "Bambu",
  "Butembo", "Katwa", "Goma"
)

# Colorblind-safe palette — Wong (2011)
PAL <- c(
  "Suspect/probable cases" = "#56B4E9",
  "Confirmed cases"        = "#D55E00",
  "Outbreak deaths"        = "#404040"
)

int_breaks <- function(x) unique(floor(pretty(x)))

theme_sitrep <- function(base_size = 11) {
  theme_minimal(base_size = base_size) %+replace%
    theme(
      plot.title       = element_text(face = "bold", size = base_size + 2, hjust = 0),
      plot.subtitle    = element_text(size = base_size - 1, colour = "grey45", hjust = 0,
                                      margin = margin(b = 8)),
      plot.caption     = element_text(size = base_size - 3, colour = "grey55", hjust = 1,
                                      margin = margin(t = 8)),
      plot.margin      = margin(12, 16, 8, 12),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(colour = "grey92"),
      legend.position  = "bottom",
      legend.key.size  = unit(0.75, "lines"),
      legend.text      = element_text(size = base_size - 1),
      strip.text       = element_text(face = "bold", size = base_size - 1,
                                      margin = margin(b = 4)),
      axis.title       = element_text(size = base_size - 1, colour = "grey35"),
      axis.text        = element_text(size = base_size - 2, colour = "grey25")
    )
}

# ── Derived subsets ────────────────────────────────────────────────────────────
totals <- df |>
  filter(count_type == "Cumules", is_agg, zone == "Total") |>
  group_by(date) |>
  slice_max(order_by = unconfirmed, n = 1, with_ties = FALSE) |>
  ungroup() |>
  arrange(date) |>
  mutate(deaths = cummax(replace_na(deaths, 0L)))

zone_ts <- df |>
  filter(count_type == "Cumules", !is_agg, zone %in% ZONE_LEVELS) |>
  group_by(date, zone, province) |>
  slice_max(order_by = unconfirmed, n = 1, with_ties = FALSE) |>
  ungroup() |>
  mutate(zone = factor(zone, levels = ZONE_LEVELS)) |>
  arrange(zone, date) |>
  group_by(zone) |>
  mutate(
    # Flag raw downward revisions before enforcing monotonicity
    confirmed_raw   = confirmed,
    unconfirmed_raw = unconfirmed,
    deaths_raw      = deaths,
    deaths_susp_raw = deaths_susp,
    rev_confirmed   = confirmed   < lag(cummax(replace_na(confirmed,   0L)), default = 0L),
    rev_unconfirmed = unconfirmed < lag(cummax(replace_na(unconfirmed, 0L)), default = 0L),
    # Apply cummax only over observed values; leave NAs before the first
    # real observation so geom_line breaks rather than drawing from zero
    unconfirmed = {x <- unconfirmed; i <- !is.na(x); if (any(i)) x[i] <- cummax(x[i]); x},
    confirmed   = {x <- confirmed;   i <- !is.na(x); if (any(i)) x[i] <- cummax(x[i]); x},
    deaths      = {x <- deaths;      i <- !is.na(x); if (any(i)) x[i] <- cummax(x[i]); x},
    deaths_susp = {x <- deaths_susp; i <- !is.na(x); if (any(i)) x[i] <- cummax(x[i]); x}
  ) |>
  ungroup()

# Warn about any downward revisions that cummax suppressed
revisions <- zone_ts |>
  filter(rev_confirmed | rev_unconfirmed) |>
  transmute(
    zone, date,
    confirmed_raw, confirmed,
    unconfirmed_raw, unconfirmed
  )
if (nrow(revisions) > 0) {
  message("NOTE: cummax suppressed ", nrow(revisions),
          " downward revision(s) in cumulative counts:\n",
          paste(capture.output(print(revisions)), collapse = "\n"))
}

zone_new <- df |>
  filter(count_type == "Nouveaux", !is_agg, zone %in% ZONE_LEVELS) |>
  group_by(date, zone, province) |>
  slice_max(order_by = unconfirmed, n = 1, with_ties = FALSE) |>
  ungroup() |>
  mutate(zone = factor(zone, levels = ZONE_LEVELS)) |>
  arrange(zone, date)

daily_new <- df |>
  filter(count_type == "Nouveaux", is_agg, zone == "Total") |>
  group_by(date) |>
  slice_max(order_by = unconfirmed, n = 1, with_ties = FALSE) |>
  ungroup() |>
  arrange(date)

make_zone_labels <- function(dat) {
  dat |>
    mutate(
      zone_label = paste0(zone, " (", province, ")"),
      zone_label = factor(zone_label,
        levels = dat |>
          distinct(zone, province) |>
          arrange(zone) |>
          mutate(lbl = paste0(zone, " (", province, ")")) |>
          pull(lbl)
      )
    )
}

# ── Inline summary statistics ──────────────────────────────────────────────────
latest_total <- totals |> slice_max(date, n = 1)
latest_date  <- format(max(totals$date, na.rm = TRUE), "%d %B %Y")
n_unconf     <- latest_total$unconfirmed
n_conf       <- latest_total$confirmed
n_deaths     <- coalesce(latest_total$deaths, 0L)
n_total      <- n_unconf + n_conf
cfr_pct      <- if (n_total > 0) round(n_deaths / n_total * 100, 1) else NA_real_
n_zones      <- n_distinct(zone_ts$zone)
n_provinces  <- n_distinct(zone_ts$province)
```

```{r}
#| label: load-kraemer
# Kraemer lab canonical name → this extraction zone name
KRAEMER_NAME_MAP <- c("Mongbalu" = "Mongbwalu", "Nyakunde" = "Nyankunde")

read_kraemer_long <- function(filename) {
  read_csv(
    here::here("Ebola_DRC_2026", "build", "long", filename),
    col_types      = cols(.default = "c"),
    show_col_types = FALSE
  )
}

kraemer <- read_kraemer_long("insp_sitrep__cumulative_confirmed_cases.csv") |>
  rename(confirmed = cumulative_confirmed_cases) |>
  full_join(
    read_kraemer_long("insp_sitrep__cumulative_suspected_cases.csv") |>
      rename(suspected = cumulative_suspected_cases),
    by = c("nom", "date")
  ) |>
  full_join(
    read_kraemer_long("insp_sitrep__cumulative_suspected_deaths.csv") |>
      rename(deaths_susp = cumulative_suspected_deaths),
    by = c("nom", "date")
  ) |>
  full_join(
    read_kraemer_long("insp_sitrep__cumulative_confirmed_deaths.csv") |>
      rename(deaths_conf = cumulative_confirmed_deaths),
    by = c("nom", "date")
  ) |>
  mutate(
    zone       = recode(nom, !!!KRAEMER_NAME_MAP),
    date       = as.Date(date),
    confirmed  = suppressWarnings(as.numeric(confirmed)),
    suspected  = suppressWarnings(as.numeric(suspected)),
    deaths_susp = suppressWarnings(as.numeric(deaths_susp)),
    deaths_conf = suppressWarnings(as.numeric(deaths_conf)),
    # Total deaths = suspected + confirmed (mirrors our coalesce logic)
    deaths     = coalesce(deaths_susp, 0) + coalesce(deaths_conf, 0),
    deaths     = if_else(is.na(deaths_susp) & is.na(deaths_conf), NA_real_, deaths),
    source     = "Kraemer lab"
  ) |>
  select(zone, date, confirmed, suspected, deaths_susp, deaths_conf, deaths, source) |>
  filter(zone %in% ZONE_LEVELS)

kraemer_hosp <- read_kraemer_long("insp_sitrep__hospitalised.csv") |>
  mutate(
    zone         = recode(nom, !!!KRAEMER_NAME_MAP),
    date         = as.Date(date),
    hospitalised = suppressWarnings(as.numeric(hospitalised))
  )

kraemer_contacts <- read_kraemer_long("insp_sitrep__cumulative_contacts_traced.csv") |>
  mutate(
    zone            = recode(nom, !!!KRAEMER_NAME_MAP),
    date            = as.Date(date),
    contacts_traced = suppressWarnings(as.numeric(cumulative_contacts_traced))
  ) |>
  filter(!is.na(contacts_traced))

# Response indicators (hospitalisations by zone)
kraemer_response <- read_kraemer_long("insp_sitrep__hospitalised.csv") |>
  full_join(read_kraemer_long("insp_sitrep__in_bed_previous_day.csv"), by = c("nom", "date")) |>
  full_join(
    read_kraemer_long("insp_sitrep__new_hosp_admissions.csv") |>
      rename(total_admissions = new_all_admissions),
    by = c("nom", "date")
  ) |>
  full_join(
    read_kraemer_long("insp_sitrep__new_hosp_detainees.csv") |>
      rename(new_admissions_ppl = new_hosp_detainees),
    by = c("nom", "date")
  ) |>
  full_join(
    read_kraemer_long("insp_sitrep__new_hosp_other.csv") |>
      rename(new_admissions_other = new_other),
    by = c("nom", "date")
  ) |>
  mutate(
    zone   = recode(nom, !!!KRAEMER_NAME_MAP),
    date   = as.Date(date),
    across(c(hospitalised, in_bed_previous_day, total_admissions,
             new_admissions_ppl, new_admissions_other),
           ~suppressWarnings(as.numeric(.x))),
    source = "Kraemer lab"
  ) |>
  filter(zone %in% ZONE_LEVELS)

# POE (checkpoint-level in Kraemer; summed to daily aggregate for comparison)
kraemer_poe <- read_kraemer_long("insp_sitrep__total_poe_passed.csv") |>
  rename(total_passed = total_poe_passed) |>
  full_join(
    read_kraemer_long("insp_sitrep__total_poe_screened.csv") |>
      rename(total_screened = total_poe_screened),
    by = c("nom", "date")
  ) |>
  full_join(
    read_kraemer_long("insp_sitrep__total_poe_hand_washing.csv") |>
      rename(total_handwashing = total_poe_hand_washing),
    by = c("nom", "date")
  ) |>
  full_join(
    read_kraemer_long("insp_sitrep__total_poe_sanitised.csv") |>
      rename(total_sensitised = total_poe_sanitised),
    by = c("nom", "date")
  ) |>
  mutate(
    date = as.Date(date),
    across(c(total_passed, total_screened, total_handwashing, total_sensitised),
           ~suppressWarnings(as.numeric(.x)))
  ) |>
  group_by(date) |>
  summarise(
    across(c(total_passed, total_screened, total_handwashing, total_sensitised),
           ~sum(.x, na.rm = TRUE)),
    .groups = "drop"
  ) |>
  mutate(source = "Kraemer lab")
```

## Introduction

The Democratic Republic of Congo is experiencing an outbreak of Maladie à Virus Ebola (MVE) with cases reported across health zones in Ituri and Nord-Kivu provinces. This report summarises the epidemiological situation based on data extracted from official INSP DRC situation reports. It is generated automatically and reflects the most recent available data.

## Methods

### Data sources

Epidemiological data were obtained from eleven PDF situation reports published by the Institut National de Santé Publique (INSP) DRC during the 2026 MVE outbreak in Ituri and Nord-Kivu provinces. PDFs were downloaded automatically from the INSP DRC website (<https://insp.cd/ebola/>), with a manifest ensuring idempotent incremental updates. MD5 checksums confirmed that the downloaded PDFs are byte-for-byte identical to those archived in the Kraemer lab repository, establishing that both sources operate on the same documents. For cross-source validation, pre-harmonised long-format case-count data were obtained from the Kraemer lab repository, which represents independent manual extraction of the same INSP DRC situation reports.

### Automated PDF extraction

Each PDF was passed to the Anthropic Claude claude-sonnet-4-6 large language model via the Messages API as a base64-encoded document, enabling pixel-level reading without reliance on the text layer. A temperature of 0 was used to maximise reproducibility. The model returned a structured JSON object containing four tables. The new-cases and cumulative case-count tables each provide rows normalised to province, health zone, and counts of suspected, probable, and confirmed cases and deaths; validation checks discarded tables identified as alert-investigation summaries rather than case counts. Where available (SitReps 006 onwards), the patient movement table (*Mouvement des malades*, Tableau VI) was also extracted, providing zone-level counts of patients in bed at the start of the day, new admissions, total isolated, and cumulative admissions. In addition, the Points d'Entrée/Points de Contrôle (PoE/PoC) summary table was extracted, recording aggregate counts of travellers screened, contacts listed, and alerts raised per reporting period. A second API call produced a verbatim full-text transcription stored as an audit trail.

### Data harmonisation

Zone names were normalised to canonical spellings. Suspect and probable case counts were combined into a single unconfirmed category to accommodate inconsistent column usage across situation reports from different issuing bodies. Cumulative counts were constrained to be monotonically non-decreasing within each health zone, and where multiple documents covered the same zone and date the row with the highest unconfirmed count was retained.

### Statistical analysis

All analyses and visualisations were produced in R (version `r R.version$major`.`r R.version$minor`) using the `tidyverse` suite [@tidyverse], with figures produced using `ggplot2` [@ggplot2] and tables formatted with `flextable` [@flextable]. The report was rendered with Quarto. The extraction pipeline and this report were developed with the assistance of GitHub Copilot (Microsoft) and the Claude API (Anthropic, model: claude-sonnet-4-6).

## Results

As of `r latest_date`, the DRC MVE outbreak had recorded a cumulative total of `r prettyNum(n_total, big.mark = ",")` cases (`r prettyNum(n_conf, big.mark = ",")` confirmed, `r prettyNum(n_unconf, big.mark = ",")` suspect or probable) and `r prettyNum(n_deaths, big.mark = ",")` deaths, corresponding to a case fatality ratio of `r cfr_pct`%. Cases have been reported across `r n_zones` health zones in `r n_provinces` provinces.

### Cumulative epidemic curve

```{r}
#| label: fig-cumulative
#| fig-cap: "Cumulative MVE cases and deaths, DRC. Suspect and probable cases are combined. End-of-line labels show the latest recorded value. Source: INSP DRC."
#| fig-width: 8
#| fig-height: 5
p1_dat <- totals |>
  select(date,
         `Suspect/probable cases` = unconfirmed,
         `Confirmed cases`        = confirmed,
         `Outbreak deaths`        = deaths) |>
  pivot_longer(-date, names_to = "category", values_to = "count") |>
  mutate(category = factor(category, levels = names(PAL)))

p1_last <- p1_dat |>
  filter(!is.na(count)) |>
  group_by(category) |>
  slice_max(date, n = 1) |>
  ungroup()

p <- ggplot(p1_dat, aes(x = date, y = count, colour = category, group = category)) +
  geom_line(linewidth = 1.1, na.rm = TRUE) +
  geom_point(size = 2.8, na.rm = TRUE) +
  geom_label_repel(
    data           = p1_last,
    aes(label      = count),
    size           = 3.5,
    nudge_x        = 1.5,
    direction      = "y",
    hjust          = 0,
    segment.colour = "grey70",
    segment.size   = 0.3,
    show.legend    = FALSE,
    na.rm          = TRUE
  ) +
  scale_colour_manual(values = PAL) +
  scale_x_date(
    date_labels = "%d %b",
    date_breaks = "2 days",
    expand      = expansion(add = c(0, 4))
  ) +
  scale_y_continuous(breaks = int_breaks, expand = expansion(mult = c(0, 0.08))) +
  labs(
    title    = "MVE DRC \u2014 Cumulative epidemic curve",
    subtitle = "Total counts across all health zones",
    x = NULL, y = "Cumulative count", colour = NULL,
    caption  = "Source: INSP DRC situation reports"
  ) +
  theme_sitrep(base_size = 12)
ggsave(here("outputs", "plots", "fig-cumulative.png"), plot = p, width = 8, height = 5, dpi = 300)
p
```

### Daily new cases

```{r}
#| label: fig-new-cases
#| fig-cap: "New cases reported per situation report across all health zones, DRC. Source: INSP DRC."
#| fig-width: 7
#| fig-height: 4
p <- daily_new |>
  select(date,
         `Suspect/probable cases` = unconfirmed,
         `Confirmed cases`        = confirmed) |>
  pivot_longer(-date, names_to = "category", values_to = "count") |>
  mutate(category = factor(category, levels = c("Suspect/probable cases", "Confirmed cases"))) |>
  ggplot(aes(x = date, y = count, fill = category)) +
  geom_col(position = "dodge", width = 0.3, na.rm = TRUE) +
  scale_fill_manual(values = PAL) +
  scale_x_date(date_labels = "%d %b", date_breaks = "1 day") +
  scale_y_continuous(breaks = int_breaks, expand = expansion(mult = c(0, 0.1))) +
  labs(
    title    = "MVE DRC \u2014 Daily new cases",
    subtitle = "Total reported per situation report",
    x = NULL, y = "New cases", fill = NULL,
    caption  = "Source: INSP DRC situation reports"
  ) +
  theme_sitrep(base_size = 12)
ggsave(here("outputs", "plots", "fig-new-cases.png"), plot = p, width = 7, height = 4, dpi = 300)
p
```

### Geographic distribution — cumulative counts

```{r}
#| label: fig-zone-cumulative
#| fig-cap: "Cumulative cases by health zone. Y-axis is free across panels. Source: INSP DRC."
#| fig-width: 10
#| fig-height: 11
p <- zone_ts |>
  make_zone_labels() |>
  select(date, zone_label,
         `Suspect/probable cases` = unconfirmed,
         `Confirmed cases`        = confirmed,
         `Outbreak deaths`        = deaths) |>
  pivot_longer(
    cols      = c(`Suspect/probable cases`, `Confirmed cases`, `Outbreak deaths`),
    names_to  = "category",
    values_to = "count"
  ) |>
  mutate(category = factor(category, levels = names(PAL))) |>
  ggplot(aes(x = date, y = count, colour = category, group = category)) +
  geom_line(linewidth = 0.9, na.rm = TRUE) +
  geom_point(size = 2.2, na.rm = TRUE) +
  scale_colour_manual(values = PAL) +
  scale_x_date(date_labels = "%d %b", date_breaks = "1 day") +
  scale_y_continuous(breaks = int_breaks, expand = expansion(mult = c(0, 0.15))) +
  facet_wrap(~ zone_label, scales = "free_y", ncol = 2) +
  labs(
    title    = "MVE DRC \u2014 Cumulative cases by health zone",
    subtitle = "Each panel shows one health zone; y-axis is free.",
    x = NULL, y = "Cumulative count", colour = NULL,
    caption  = "Source: INSP DRC situation reports"
  ) +
  theme_sitrep(base_size = 10) +
  theme(axis.text.x = element_text(size = 8, angle = 45, hjust = 1))
ggsave(here("outputs", "plots", "fig-zone-cumulative.png"), plot = p, width = 10, height = 11, dpi = 300)
p
```

### Geographic distribution — new cases

```{r}
#| label: fig-zone-new
#| fig-cap: "New cases per situation report by health zone. Y-axis is free across panels. Source: INSP DRC."
#| fig-width: 10
#| fig-height: 11
p <- zone_new |>
  make_zone_labels() |>
  select(date, zone_label,
         `Suspect/probable cases` = unconfirmed,
         `Confirmed cases`        = confirmed) |>
  pivot_longer(
    cols      = c(`Suspect/probable cases`, `Confirmed cases`),
    names_to  = "category",
    values_to = "count"
  ) |>
  mutate(category = factor(category, levels = c("Suspect/probable cases", "Confirmed cases"))) |>
  ggplot(aes(x = date, y = count, fill = category)) +
  geom_col(position = "dodge", width = 0.4, na.rm = TRUE) +
  scale_fill_manual(values = PAL) +
  scale_x_date(date_labels = "%d %b", date_breaks = "1 day") +
  scale_y_continuous(breaks = int_breaks, expand = expansion(mult = c(0, 0.15))) +
  facet_wrap(~ zone_label, scales = "free_y", ncol = 2) +
  labs(
    title    = "MVE DRC \u2014 Daily new cases by health zone",
    subtitle = "Per situation report; y-axis is free.",
    x = NULL, y = "New cases", fill = NULL,
    caption  = "Source: INSP DRC situation reports"
  ) +
  theme_sitrep(base_size = 10) +
  theme(axis.text.x = element_text(size = 8, angle = 45, hjust = 1))
ggsave(here("outputs", "plots", "fig-zone-new.png"), plot = p, width = 10, height = 11, dpi = 300)
p
```

### Current counts by health zone

Table 1 presents the most recent cumulative case counts disaggregated by health zone, drawn from the latest available situation report for each zone.

```{r}
#| label: tbl-zones
#| tbl-cap: "Table 1. Latest cumulative case counts by health zone."
zone_latest <- zone_ts |>
  group_by(zone, province) |>
  slice_max(date, n = 1) |>
  ungroup() |>
  arrange(zone) |>
  transmute(
    `Health zone`      = as.character(zone),
    Province           = province,
    `Latest sitrep`    = format(date, "%d %b %Y"),
    `Suspect/probable` = unconfirmed,
    `Confirmed`        = confirmed,
    `Deaths`           = deaths
  )

total_row <- tibble(
  `Health zone`      = "Total",
  Province           = "\u2014",
  `Latest sitrep`    = format(max(totals$date, na.rm = TRUE), "%d %b %Y"),
  `Suspect/probable` = sum(zone_latest$`Suspect/probable`, na.rm = TRUE),
  `Confirmed`        = sum(zone_latest$`Confirmed`,        na.rm = TRUE),
  `Deaths`           = sum(zone_latest$`Deaths`,           na.rm = TRUE)
)

bind_rows(zone_latest, total_row) |>
  flextable() |>
  bold(part = "header") |>
  bold(i = nrow(zone_latest) + 1L) |>
  bg(part = "header", bg = "#1A3A5C") |>
  color(part = "header", color = "white") |>
  bg(i = nrow(zone_latest) + 1L, bg = "#E8EFF7") |>
  hline(i = nrow(zone_latest), part = "body") |>
  align(j = 4:6, align = "right", part = "all") |>
  colformat_int(j = 4:6) |>
  autofit()
```

### Cross-source validation

Cumulative zone-level counts from this pipeline are compared against the [Kraemer lab Ebola DRC 2026](https://github.com/kraemer-lab/Ebola_DRC_2026) repository, which codes INSP DRC situation reports into a standardised long-format by manual human extraction. Both pipelines operate on the same source PDFs (MD5-verified). Zone names differ between sources at two sites (Mongbwalu/Mongbalu and Nyankunde/Nyakunde); counts are matched on the harmonised name. `ND` values in the Kraemer lab series are treated as missing. Any divergence between the two series therefore reflects differences in extraction method — automated LLM-based visual extraction versus human manual coding — rather than differences in the source documents.

### Data quality notes — case counts

Two discrepancies in the Mongbwalu series require attention. First, the Kraemer lab series records 246 cumulative suspected cases at Mongbwalu as early as 14 May 2026, whereas the earliest situation report captured by this pipeline dates to 18 May (at which point the cumulative suspected count is 302). This pipeline therefore does not represent the full epidemic history at Mongbwalu: a substantial burden of suspected cases predates the extraction window, and the apparent early trajectory visible in the automated series is an artefact of incomplete coverage rather than a true epidemic onset.

Second, both this pipeline (before cummax) and the Kraemer lab series record a decrease in cumulative confirmed cases for Mongbwalu between 21 May (14 cases) and 22 May (10 cases). As cumulative counts cannot decrease, this reflects an error in the 22 May situation report itself — either a reclassification of previously confirmed cases or a data entry error — rather than an artefact of the extraction method. The monotonic cummax constraint applied to the automated extraction series suppresses this dip (retaining 14 on 22 May), but the Kraemer lab series, which does not apply cummax, shows the raw decrease. Users should treat the confirmed case count for Mongbwalu around 22 May with caution.

```{r}
#| label: fig-comparison
#| fig-cap: "Cross-source comparison of cumulative confirmed cases, suspect/probable cases, and outbreak deaths by health zone. Solid lines: automated extraction (cummax applied); dotted lines: automated extraction (raw, no cummax); dashed lines: Kraemer lab (manual extraction). Both automated and Kraemer sources use identical PDFs; divergence reflects extraction method differences."
#| fig-width: 12
#| fig-height: 14
PAL_COMP <- c(
  "Automated extraction" = "#0072B2",
  "Automated (no cummax)" = "#56B4E9",
  "Kraemer lab"          = "#E69F00"
)
LTY_COMP <- c(
  "Automated extraction" = "solid",
  "Automated (no cummax)" = "dotted",
  "Kraemer lab"          = "dashed"
)

zone_province_lut <- zone_ts |>
  distinct(zone, province) |>
  mutate(zone = as.character(zone))

our_for_comp <- zone_ts |>
  transmute(zone = as.character(zone), province, date,
            confirmed, suspected = unconfirmed,
            deaths_susp, deaths,
            source = "Automated extraction")

our_raw_for_comp <- zone_ts |>
  transmute(zone = as.character(zone), province, date,
            confirmed   = confirmed_raw,
            suspected   = unconfirmed_raw,
            deaths_susp = deaths_susp_raw,
            deaths      = deaths_raw,
            source      = "Automated (no cummax)")

# Kraemer deaths_susp = suspected only; deaths = suspected + confirmed (total)
kraemer_for_comp <- kraemer |>
  left_join(zone_province_lut, by = "zone") |>
  transmute(zone, province, date, confirmed, suspected,
            deaths_susp, deaths,
            source)

comp_long <- bind_rows(our_for_comp, our_raw_for_comp, kraemer_for_comp) |>
  pivot_longer(c(confirmed, suspected, deaths_susp, deaths),
               names_to = "metric", values_to = "count") |>
  mutate(
    metric = recode(metric,
      "confirmed"   = "Confirmed cases",
      "suspected"   = "Suspect/probable cases",
      "deaths_susp" = "Suspected deaths",
      "deaths"      = "Total deaths (coalesced)"
    ),
    metric   = factor(metric, levels = c("Confirmed cases", "Suspect/probable cases",
                                          "Suspected deaths", "Total deaths (coalesced)")),
    source   = factor(source, levels = c("Automated extraction", "Automated (no cummax)", "Kraemer lab")),
    province = factor(province, levels = c("Ituri", "Nord-Kivu")),
    zone     = factor(zone, levels = ZONE_LEVELS)
  ) |>
  filter(!is.na(count))

p <- ggplot(comp_long, aes(x = date, y = count, colour = source, linetype = source)) +
  geom_line(linewidth = 0.9, na.rm = TRUE) +
  geom_point(size = 2.0, na.rm = TRUE) +
  facet_nested(metric ~ province + zone, scales = "free_y") +
  scale_colour_manual(values = PAL_COMP) +
  scale_linetype_manual(values = LTY_COMP) +
  scale_x_date(date_labels = "%d %b", date_breaks = "2 days") +
  scale_y_continuous(breaks = int_breaks, limits = c(0, NA), expand = expansion(mult = c(0, 0.15))) +
  labs(
    title    = "MVE DRC \u2014 Cross-source validation",
    subtitle = "Cumulative counts: automated extraction (with and without cummax) vs Kraemer lab (manual extraction, same source PDFs)",
    x = NULL, y = "Cumulative count", colour = NULL, linetype = NULL,
    caption  = "Sources: INSP DRC situation reports; github.com/kraemer-lab/Ebola_DRC_2026"
  ) +
  theme_sitrep(base_size = 9) +
  theme(
    axis.text.x     = element_text(angle = 45, hjust = 1, size = 7),
    panel.border    = element_rect(colour = "grey70", fill = NA, linewidth = 0.4),
    panel.spacing   = unit(0.6, "lines")
  )
ggsave(here("outputs", "plots", "fig-comparison.png"), plot = p, width = 12, height = 14, dpi = 300)
p
```

Table 2 presents numerical differences at the most recent date available in each source for each zone.

```{r}
#| label: tbl-comparison
#| tbl-cap: "Table 2. Cumulative case counts at the latest available date per zone: automated extraction vs Kraemer lab. Delta = ours minus Kraemer. Dashes indicate no data in that source."
ours_latest <- zone_ts |>
  group_by(zone) |>
  slice_max(date, n = 1) |>
  ungroup() |>
  transmute(zone, conf_ours = confirmed, susp_ours = unconfirmed, dead_ours = deaths)

kraemer_latest <- kraemer |>
  group_by(zone) |>
  slice_max(date, n = 1) |>
  ungroup() |>
  transmute(zone, conf_them = confirmed, susp_them = suspected, dead_them = deaths)

full_join(ours_latest, kraemer_latest, by = "zone") |>
  mutate(
    d_conf = conf_ours - conf_them,
    d_susp = susp_ours - susp_them,
    d_dead = dead_ours - dead_them
  ) |>
  arrange(zone) |>
  transmute(
    zone,
    conf_ours, conf_them, d_conf,
    susp_ours, susp_them, d_susp,
    dead_ours, dead_them, d_dead
  ) |>
  setNames(c(
    "Health zone",
    "Confirmed (ours)", "Confirmed (Kraemer)", "\u0394 confirmed",
    "Suspect/prob. (ours)", "Suspect/prob. (Kraemer)", "\u0394 suspect/prob.",
    "Deaths (ours)", "Deaths (Kraemer)", "\u0394 deaths"
  )) |>
  flextable() |>
  bold(part = "header") |>
  bg(part = "header", bg = "#1A3A5C") |>
  color(part = "header", color = "white") |>
  align(j = 2:10, align = "right", part = "all") |>
  colformat_int(j = 2:10, na_str = "\u2014") |>
  autofit()
```

### Response indicators

```{r}
#| label: load-response-ours
resp_ours <- read_csv(
  here::here("outputs", "master_response_counts.csv"),
  col_types      = cols(.default = "c"),
  show_col_types = FALSE
) |>
  mutate(
    date   = dmy(date),
    zone   = factor(zone, levels = ZONE_LEVELS),
    across(c(in_bed_previous_day, new_admissions_ppl, new_admissions_other,
             total_admissions, total_isolated, cumulative_admitted),
           ~suppressWarnings(as.numeric(.x))),
    source = "Automated extraction"
  ) |>
  filter(!is.na(zone), !zone %in% c("Total"))

poe_ours <- read_csv(
  here::here("outputs", "master_poe_counts.csv"),
  col_types      = cols(.default = "c"),
  show_col_types = FALSE
) |>
  mutate(
    date = dmy(date),
    across(c(total_passed, total_screened, total_handwashing, total_sensitised),
           ~suppressWarnings(as.numeric(.x))),
    source = "Automated extraction"
  ) |>
  select(date, total_passed, total_screened, total_handwashing, total_sensitised, source)
```

### Data quality notes — response indicators

Two discrepancies were identified through cross-source validation.

**Rwampara ambiguous table structure (21 May).** SitRep 007 presents Rwampara across three columns (in-bed: 21 | 15 | 35; total isolated: 23 | 19 | 46) whose relationship is not explained in the document — they may represent different isolation units, patient categories, or time sub-periods. The automated extraction summed the first two sub-rows (total isolated = 42); the Kraemer lab summed all three (88). Neither value can be verified against the source document. The 21 May Rwampara data point should be treated with caution in both series.

**New admissions column mapping (20–21 May).** SitReps 005 and 006 used an earlier table format that did not include an explicit "in-bed-previous-day" column. Kraemer's extraction assigned the `new_all_admissions` label to what was in fact the in-bed-previous-day value for those two dates, with the consequence that their new-admissions series for 20–21 May reflects the wrong metric. Values from 22 May onwards, when the table format was standardised, are concordant between the two sources.

```{r}
#| label: fig-comparison-response
#| fig-cap: "Cross-source comparison of zone-level response indicators. Solid lines: automated extraction; dashed lines: Kraemer lab. The 21 May Rwampara values are unreliable in both series: the sitrep presents three unexplained sub-columns whose relationship is ambiguous. For 20–21 May, Kraemer's new-admissions values reflect a column-mapping error in the earlier table format."
#| fig-width: 12
#| fig-height: 9
PAL_RESP <- c("Automated extraction" = "#0072B2", "Kraemer lab" = "#E69F00")
LTY_RESP <- c("Automated extraction" = "solid",   "Kraemer lab" = "dashed")

resp_comp <- bind_rows(
  resp_ours |>
    transmute(date, zone,
              hospitalised        = total_isolated,
              in_bed_previous_day,
              total_admissions,
              source),
  kraemer_response |>
    transmute(date, zone = factor(zone, levels = ZONE_LEVELS),
              hospitalised,
              in_bed_previous_day,
              total_admissions,
              source)
) |>
  pivot_longer(c(hospitalised, in_bed_previous_day, total_admissions),
               names_to = "metric", values_to = "count") |>
  mutate(
    metric = recode(metric,
      "hospitalised"        = "Currently hospitalised",
      "in_bed_previous_day" = "In bed (prev. day)",
      "total_admissions"    = "New admissions (total)"
    ),
    metric = factor(metric, levels = c("Currently hospitalised",
                                        "In bed (prev. day)",
                                        "New admissions (total)")),
    source = factor(source, levels = c("Automated extraction", "Kraemer lab"))
  ) |>
  filter(!is.na(count), !is.na(zone))

p <- ggplot(resp_comp, aes(x = date, y = count, colour = source, linetype = source)) +
  geom_line(linewidth = 0.9, na.rm = TRUE) +
  geom_point(size = 2.0, na.rm = TRUE) +
  facet_nested(metric ~ zone, scales = "free_y") +
  scale_colour_manual(values = PAL_RESP) +
  scale_linetype_manual(values = LTY_RESP) +
  scale_x_date(date_labels = "%d %b", date_breaks = "2 days") +
  scale_y_continuous(breaks = int_breaks, limits = c(0, NA),
                     expand = expansion(mult = c(0, 0.15))) +
  labs(
    title    = "MVE DRC \u2014 Response indicators — cross-source validation",
    subtitle = "Automated extraction vs Kraemer lab (manual, same source PDFs)",
    x = NULL, y = "Count", colour = NULL, linetype = NULL,
    caption  = "Sources: INSP DRC situation reports; github.com/kraemer-lab/Ebola_DRC_2026"
  ) +
  theme_sitrep(base_size = 9) +
  theme(
    axis.text.x  = element_text(angle = 45, hjust = 1, size = 7),
    panel.border = element_rect(colour = "grey70", fill = NA, linewidth = 0.4),
    panel.spacing = unit(0.5, "lines")
  )
ggsave(here("outputs", "plots", "fig-comparison-response.png"), plot = p,
       width = 12, height = 9, dpi = 300)
p
```

```{r}
#| label: fig-comparison-poe
#| fig-cap: "Cross-source comparison of daily Point of Entry (POE) screening totals. Automated extraction reports a single aggregate per sitrep date; Kraemer lab data are summed across individual checkpoints to produce a daily total."
#| fig-width: 9
#| fig-height: 6
poe_comp <- bind_rows(poe_ours, kraemer_poe) |>
  pivot_longer(c(total_passed, total_screened, total_handwashing, total_sensitised),
               names_to = "metric", values_to = "count") |>
  mutate(
    metric = recode(metric,
      "total_passed"      = "Total passed",
      "total_screened"    = "Screened",
      "total_handwashing" = "Handwashing stations",
      "total_sensitised"  = "Sensitised"
    ),
    metric = factor(metric, levels = c("Total passed", "Screened",
                                        "Handwashing stations", "Sensitised")),
    source = factor(source, levels = c("Automated extraction", "Kraemer lab"))
  ) |>
  filter(!is.na(count))

p <- ggplot(poe_comp, aes(x = date, y = count, colour = source, linetype = source)) +
  geom_line(linewidth = 0.9, na.rm = TRUE) +
  geom_point(size = 2.5, na.rm = TRUE) +
  facet_wrap(~ metric, scales = "free_y", ncol = 2) +
  scale_colour_manual(values = PAL_RESP) +
  scale_linetype_manual(values = LTY_RESP) +
  scale_x_date(date_labels = "%d %b", date_breaks = "2 days") +
  scale_y_continuous(breaks = int_breaks, limits = c(0, NA),
                     expand = expansion(mult = c(0, 0.12))) +
  labs(
    title    = "MVE DRC \u2014 POE screening — cross-source validation",
    subtitle = "Daily aggregate totals: automated extraction vs Kraemer lab (sum across checkpoints)",
    x = NULL, y = "Count", colour = NULL, linetype = NULL,
    caption  = "Sources: INSP DRC situation reports; github.com/kraemer-lab/Ebola_DRC_2026"
  ) +
  theme_sitrep(base_size = 11) +
  theme(
    axis.text.x  = element_text(angle = 45, hjust = 1, size = 8),
    panel.border = element_rect(colour = "grey70", fill = NA, linewidth = 0.4),
    panel.spacing = unit(0.6, "lines")
  )
ggsave(here("outputs", "plots", "fig-comparison-poe.png"), plot = p,
       width = 9, height = 6, dpi = 300)
p
```

```{r}
#| label: fig-comparison-contacts
#| fig-cap: "Cross-source comparison of cumulative contacts traced by health zone. Solid lines: automated extraction (cummax applied); dashed lines: Kraemer lab (manual extraction). ND values excluded."
#| fig-width: 11
#| fig-height: 7
contacts_ours <- df |>
  filter(
    count_type == "Cumules",
    zone %in% ZONE_LEVELS,
    !is.na(contacts), contacts != ""
  ) |>
  mutate(contacts = suppressWarnings(as.numeric(contacts)),
         zone = factor(zone, levels = ZONE_LEVELS)) |>
  filter(!is.na(contacts)) |>
  arrange(zone, date) |>
  group_by(zone) |>
  mutate(contacts = cummax(contacts)) |>
  ungroup() |>
  transmute(date, zone, contacts, source = "Automated extraction")

contacts_kraemer <- kraemer_contacts |>
  filter(zone %in% ZONE_LEVELS) |>
  transmute(date, zone = factor(zone, levels = ZONE_LEVELS),
            contacts = contacts_traced, source = "Kraemer lab")

contacts_comp <- bind_rows(contacts_ours, contacts_kraemer) |>
  filter(!is.na(contacts)) |>
  left_join(zone_province_lut, by = "zone") |>
  mutate(
    source   = factor(source, levels = c("Automated extraction", "Kraemer lab")),
    province = factor(province, levels = c("Ituri", "Nord-Kivu")),
    zone     = factor(zone, levels = ZONE_LEVELS)
  )

p <- ggplot(contacts_comp,
            aes(x = date, y = contacts, colour = source, linetype = source)) +
  geom_line(linewidth = 0.9, na.rm = TRUE) +
  geom_point(size = 2.0, na.rm = TRUE) +
  facet_nested(~ province + zone, scales = "free_y") +
  scale_colour_manual(values = PAL_COMP[c("Automated extraction","Kraemer lab")]) +
  scale_linetype_manual(values = LTY_COMP[c("Automated extraction","Kraemer lab")]) +
  scale_x_date(date_labels = "%d %b", date_breaks = "2 days") +
  scale_y_continuous(breaks = int_breaks, limits = c(0, NA),
                     expand = expansion(mult = c(0, 0.15))) +
  labs(
    title    = "MVE DRC \u2014 Cumulative contacts traced — cross-source validation",
    subtitle = "Automated extraction (cummax) vs Kraemer lab (manual extraction, same source PDFs)",
    x = NULL, y = "Cumulative contacts traced", colour = NULL, linetype = NULL,
    caption  = "Sources: INSP DRC situation reports; github.com/kraemer-lab/Ebola_DRC_2026"
  ) +
  theme_sitrep(base_size = 9) +
  theme(
    axis.text.x  = element_text(angle = 45, hjust = 1, size = 7),
    panel.border = element_rect(colour = "grey70", fill = NA, linewidth = 0.4),
    panel.spacing = unit(0.5, "lines")
  )
ggsave(here("outputs", "plots", "fig-comparison-contacts.png"), plot = p,
       width = 11, height = 7, dpi = 300)
p
```

## Discussion

The outbreak trajectory shows a consistent increase in suspect and probable case counts across the reporting period. Confirmed cases remain a minority of the total, reflecting the challenges of laboratory confirmation in the field setting. The geographic spread across both Ituri and Nord-Kivu provinces underscores the need for multi-province coordination of the response.

Per-zone new case data are only available from 20 May 2026 onwards, as earlier situation reports did not disaggregate new cases by health zone. Response indicator data (hospitalisation, contacts traced) are sourced from the Kraemer lab repository as these metrics are not captured in the automated extraction pipeline. Results should be interpreted in the context of reporting delays and the possibility of incomplete situation report coverage for some health zones and dates. The case fatality ratio reported here should be treated as a provisional estimate, as delays in death reporting may affect the numerator, and the deaths column captures the first available of deaths-among-suspects, deaths-among-probables, and deaths-among-confirmed, which are not consistently labelled across situation reports.

MD5 checksum verification confirmed that the PDFs processed by this pipeline are byte-for-byte identical to those archived in the Kraemer lab repository. Divergence in the cross-source comparison therefore reflects either differences in how each extraction method interprets ambiguous table layouts, or outright reading errors in one of the two series.

Four sources of divergence were identified. First, for SitReps N°001 and N°002 (14–15 May), the INSP documents present only an outbreak-wide aggregate without a standardised zone-level cumulative tableau. The N°001 PDF text explicitly states that its 246-case suspect total covers "Zones de Santé (ZS) de Mongbwalu, Rwampara et Bunia" collectively; Kraemer incorrectly attributed this aggregate to Mongbwalu alone. This pipeline extracted the zone-level probable case figures from Tableau 1 in N°002 (57 for Mongbwalu, 27 for Rwampara, 3 for Bunia) and has no zone-specific entry for 14 May, where no zone-level breakdown exists. Second, direct comparison of extracted values against the PDF text reveals reading errors in the Kraemer series for two zone–date combinations: Katwa confirmed cases are recorded as 1 and 0 by Kraemer for 21 and 22 May respectively, whereas both the published PDF text and this pipeline's extraction agree on 2 and 3; Butembo confirmed cases are recorded as 2 by Kraemer for 21 May, whereas the PDF states 1 and this pipeline extracted 1. A persistent one-case difference in Goma suspected cases (0 in this pipeline versus 1 in Kraemer across 20–23 May) could not be attributed to a specific textual feature and may reflect a cell-border ambiguity. Third, this pipeline enforces monotonic non-decrease via cumulative maximum, masking genuine downward revisions visible in the Kraemer series: Mongbwalu confirmed cases fall from 14 (21 May) to 10 (22 May) in the Kraemer data, most likely reflecting reclassification of four previously confirmed cases; this pipeline retains 14 on 22 May before both series converge at 19 on 23 May. Nyankunde confirmed cases similarly fall from 11 to 9 between 21 and 22 May in the Kraemer data; the cummax retains 11 across 22–23 May. Fourth, three of the PDFs processed here carry explicit draft labels in their filenames (`Draft-MVE-Ituri-ActuelB-1`, `Draft_SitRep_MVE_RDC_20260520_vf1`, `Draft_SitRep_MVE_RDC_2022vf2_IM`); the Kraemer lab may hold the corresponding final published versions, and provisional counts in drafts may differ from figures in the final document. This pipeline has additionally processed SitReps N°009 and N°010 (through 25 May), which extend beyond the current Kraemer series, and several health zones present in the Kraemer series (Aru, Kilo, Nizi, Miti-Murhesa) are excluded from this pipeline's zone filter.
