---
title: "Maladie à Virus Ebola — Situation Report Extraction"
subtitle: "Democratic Republic of Congo"
author: "Billy J Quilty (Charité Berlin, LSHTM & MSF Epicentre)"
date: today
date-format: "DD MMMM YYYY"
format:
  html:
    toc: true
    toc-depth: 2
    toc-title: "Contents"
    theme: flatly
    embed-resources: true
    fig-align: center
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")
pacman::p_load(tidyverse, lubridate, here, scales, ggrepel, flextable, ggh4x, patchwork)

set_flextable_defaults(
  font.family = "Arial",
  font.size   = 10,
  border.color = "#CCCCCC"
)
```

```{r}
#| label: load-data
df_raw <- read_csv(
  here::here("data", "processed", "master_combined_counts.csv"),
  col_types       = cols(.default = "c"),
  show_col_types  = FALSE
)

df <- df_raw |>
  mutate(
    # Normalise zone spellings that vary across SitRep versions
    zone = case_match(
      zone,
      "ARU"          ~ "Aru",
      "Karissibi"    ~ "Karisimbi",
      "Miti Murhesa" ~ "Miti-Murhesa",
      "TOTAL"        ~ "Total",
      .default = zone
    )
  ) |>
  mutate(
    date        = dmy(count_end_date),
    suspect     = as.numeric(cases_suspect),
    probable    = as.numeric(cases_probable),
    confirmed   = as.numeric(cases_confirmed),
    deaths      = coalesce(as.numeric(deaths_suspected), as.numeric(deaths_probable), as.numeric(deaths_confirmed)),
    deaths_susp = as.numeric(deaths_suspected),
    deaths_conf = as.numeric(deaths_confirmed),
    # Some rows are aggregate/unknown but were not flagged as such at extraction.
    # Mark them here so they are excluded from zone-level plots.
    is_agg      = (toupper(is_aggregate) == "TRUE") |
                  grepl("^Autres ZS", zone, ignore.case = TRUE) |
                  zone %in% c("Sans fiche", "Echantillons sans fiche",
                              "ZS non identifiée", "TOTAL"),
    # Only sum to zero when at least one component is present; otherwise NA
    # (avoids a false zero for zones that simply weren't reported in that sitrep)
    unconfirmed = if_else(
      !is.na(suspect) | !is.na(probable),
      coalesce(suspect, 0L) + coalesce(probable, 0L),
      NA_real_
    )
  )

# Backfill province for rows where the PDF did not repeat it on every data row.
# Aggregate/total rows are left as-is (their province is intentionally blank or
# a province name used to identify the aggregate level).
ZONE_PROVINCE_LUT <- tribble(
  ~zone,                     ~province_lut,
  "Aru",                     "Ituri",
  "Aungba",                  "Ituri",
  "Bambu",                   "Ituri",
  "Bunia",                   "Ituri",
  "Damas",                   "Ituri",
  "Gety",                    "Ituri",
  "Kilo",                    "Ituri",
  "Kilo Mission",            "Ituri",
  "Logo",                    "Ituri",
  "Mambasa",                 "Ituri",
  "Mangala",                 "Ituri",
  "Manguripa",               "Ituri",
  "Mongbwalu",               "Ituri",
  "Nizi",                    "Ituri",
  "Nyankunde",               "Ituri",
  "Rimba",                   "Ituri",
  "Rwampara",                "Ituri",
  "Butembo",                 "Nord-Kivu",
  "Goma",                    "Nord-Kivu",
  "Kalunguta",               "Nord-Kivu",
  "Karisimbi",               "Nord-Kivu",
  "Katwa",                   "Nord-Kivu",
  "Kyondo",                  "Nord-Kivu",
  "Lita",                    "Nord-Kivu",
  "Oicha",                   "Nord-Kivu",
  "Miti-Murhesa",            "Sud-Kivu",
)

df <- df |>
  left_join(ZONE_PROVINCE_LUT, by = "zone") |>
  mutate(province = if_else(!is_agg & is.na(province), province_lut, province)) |>
  select(-province_lut)

# Restrict the main comparison plot to the 8 highest-volume health zones
# in the latest processed data, so the zone panels stay readable.
TOP_N_ZONES <- 8

ZONE_LEVELS <- df |>
  filter(count_type == "Cumules", !is_agg, !is.na(zone), zone != "Total") |>
  group_by(zone) |>
  slice_max(order_by = date, n = 1, with_ties = FALSE) |>
  ungroup() |>
  mutate(total_cases = coalesce(unconfirmed, 0) + coalesce(confirmed, 0)) |>
  arrange(desc(total_cases), zone) |>
  slice_head(n = TOP_N_ZONES) |>
  pull(zone)

# Colorblind-safe palette — Wong (2011)
PAL <- c(
  "Suspect/probable cases" = "#56B4E9",
  "Confirmed cases"        = "#D55E00",
  "Outbreak deaths"        = "#404040"
)

int_breaks <- function(x) unique(floor(pretty(x)))

theme_sitrep <- function(base_size = 11) {
  theme_minimal(base_size = base_size) %+replace%
    theme(
      plot.title       = element_text(face = "bold", size = base_size + 2, hjust = 0),
      plot.subtitle    = element_text(size = base_size - 1, colour = "grey45", hjust = 0,
                                      margin = margin(b = 8)),
      plot.caption     = element_text(size = base_size - 3, colour = "grey55", hjust = 1,
                                      margin = margin(t = 8)),
      plot.margin      = margin(12, 16, 8, 12),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(colour = "grey92"),
      legend.position  = "bottom",
      legend.key.size  = unit(0.75, "lines"),
      legend.text      = element_text(size = base_size - 1),
      strip.text       = element_text(face = "bold", size = base_size - 1,
                                      margin = margin(b = 4)),
      axis.title       = element_text(size = base_size - 1, colour = "grey35"),
      axis.text        = element_text(size = base_size - 2, colour = "grey25")
    )
}

# ── Derived subsets ────────────────────────────────────────────────────────────
totals <- df |>
  filter(count_type == "Cumules", is_agg, zone == "Total") |>
  group_by(date) |>
  slice_max(order_by = unconfirmed, n = 1, with_ties = FALSE) |>
  ungroup() |>
  arrange(date) |>
  mutate(deaths = cummax(replace_na(deaths, 0L)))

zone_ts <- df |>
  filter(count_type == "Cumules", !is_agg, zone %in% ZONE_LEVELS) |>
  group_by(date, zone, province) |>
  slice_max(order_by = unconfirmed, n = 1, with_ties = FALSE) |>
  ungroup() |>
  mutate(zone = factor(zone, levels = ZONE_LEVELS)) |>
  arrange(zone, date) |>
  group_by(zone) |>
  mutate(
    # Flag raw downward revisions before enforcing monotonicity
    confirmed_raw   = confirmed,
    unconfirmed_raw = unconfirmed,
    deaths_raw      = deaths,
    deaths_susp_raw = deaths_susp,
    deaths_conf_raw = deaths_conf,
    rev_confirmed   = confirmed   < lag(cummax(replace_na(confirmed,   0L)), default = 0L),
    rev_unconfirmed = unconfirmed < lag(cummax(replace_na(unconfirmed, 0L)), default = 0L),
    # Apply cummax only over observed values; leave NAs before the first
    # real observation so geom_line breaks rather than drawing from zero
    unconfirmed = {x <- unconfirmed; i <- !is.na(x); if (any(i)) x[i] <- cummax(x[i]); x},
    confirmed   = {x <- confirmed;   i <- !is.na(x); if (any(i)) x[i] <- cummax(x[i]); x},
    deaths      = {x <- deaths;      i <- !is.na(x); if (any(i)) x[i] <- cummax(x[i]); x},
    deaths_susp = {x <- deaths_susp; i <- !is.na(x); if (any(i)) x[i] <- cummax(x[i]); x},
    deaths_conf = {x <- deaths_conf; i <- !is.na(x); if (any(i)) x[i] <- cummax(x[i]); x}
  ) |>
  ungroup()

# Warn about any downward revisions that cummax suppressed
revisions <- zone_ts |>
  filter(rev_confirmed | rev_unconfirmed) |>
  transmute(
    zone, date,
    confirmed_raw, confirmed,
    unconfirmed_raw, unconfirmed
  )
if (nrow(revisions) > 0) {
  message("NOTE: cummax suppressed ", nrow(revisions),
          " downward revision(s) in cumulative counts:\n",
          paste(capture.output(print(revisions)), collapse = "\n"))
}

zone_new <- df |>
  filter(count_type == "Nouveaux", !is_agg, zone %in% ZONE_LEVELS) |>
  group_by(date, zone, province) |>
  slice_max(order_by = unconfirmed, n = 1, with_ties = FALSE) |>
  ungroup() |>
  mutate(zone = factor(zone, levels = ZONE_LEVELS)) |>
  arrange(zone, date)

daily_new <- df |>
  filter(count_type == "Nouveaux", is_agg, zone == "Total") |>
  group_by(date) |>
  slice_max(order_by = unconfirmed, n = 1, with_ties = FALSE) |>
  ungroup() |>
  arrange(date)

make_zone_labels <- function(dat) {
  dat |>
    mutate(
      zone_label = paste0(zone, " (", province, ")"),
      zone_label = factor(zone_label,
        levels = dat |>
          distinct(zone, province) |>
          arrange(zone) |>
          mutate(lbl = paste0(zone, " (", province, ")")) |>
          pull(lbl)
      )
    )
}

# ── Inline summary statistics ──────────────────────────────────────────────────
latest_total <- totals |> slice_max(date, n = 1)
latest_date  <- format(max(totals$date, na.rm = TRUE), "%d %B %Y")
n_unconf     <- coalesce(latest_total$unconfirmed, 0)
n_conf       <- coalesce(latest_total$confirmed, 0)
n_deaths     <- coalesce(latest_total$deaths, 0L)
n_total      <- n_unconf + n_conf
cfr_pct      <- if (n_total > 0) round(n_deaths / n_total * 100, 1) else NA_real_
n_zones      <- n_distinct(zone_ts$zone)
n_provinces  <- n_distinct(zone_ts$province)
n_sitreps    <- n_distinct(df$sitrep_source)
```

```{r}
#| label: load-kraemer
# INRB-UMIE canonical name → this extraction zone name
KRAEMER_NAME_MAP <- c("Mongbalu" = "Mongbwalu", "Nyakunde" = "Nyankunde")

read_kraemer_long <- function(filename) {
  read_csv(
    here::here(
      "Ebola_DRC_2026", "data", "insp_sitrep", "processed",
      paste0(filename, "__daily.csv")
    ),
    col_types      = cols(.default = "c"),
    show_col_types = FALSE
  )
}

parse_kraemer_date <- function(x) {
  coalesce(dmy(x, quiet = TRUE), ymd(x, quiet = TRUE), mdy(x, quiet = TRUE))
}

kraemer <- read_kraemer_long("insp_sitrep__cumulative_confirmed_cases") |>
  rename(confirmed = cumulative_confirmed_cases) |>
  full_join(
    read_kraemer_long("insp_sitrep__cumulative_suspected_cases") |>
      rename(suspected = cumulative_suspected_cases),
    by = c("nom", "date")
  ) |>
  full_join(
    read_kraemer_long("insp_sitrep__cumulative_suspected_deaths") |>
      rename(deaths_susp = cumulative_suspected_deaths),
    by = c("nom", "date")
  ) |>
  full_join(
    read_kraemer_long("insp_sitrep__cumulative_confirmed_deaths") |>
      rename(deaths_conf = cumulative_confirmed_deaths),
    by = c("nom", "date")
  ) |>
  mutate(
    zone       = recode(nom, !!!KRAEMER_NAME_MAP),
    date       = parse_kraemer_date(date),
    confirmed  = suppressWarnings(as.numeric(confirmed)),
    suspected  = suppressWarnings(as.numeric(suspected)),
    deaths_susp = suppressWarnings(as.numeric(deaths_susp)),
    deaths_conf = suppressWarnings(as.numeric(deaths_conf)),
    # Total deaths = suspected + confirmed (mirrors our coalesce logic)
    deaths     = coalesce(deaths_susp, 0) + coalesce(deaths_conf, 0),
    deaths     = if_else(is.na(deaths_susp) & is.na(deaths_conf), NA_real_, deaths),
    source     = "INRB-UMIE"
  ) |>
  select(zone, date, confirmed, suspected, deaths_susp, deaths_conf, deaths, source) |>
  filter(zone %in% ZONE_LEVELS)

kraemer_hosp <- read_kraemer_long("insp_sitrep__hospitalised") |>
  mutate(
    zone         = recode(nom, !!!KRAEMER_NAME_MAP),
    date         = parse_kraemer_date(date),
    hospitalised = suppressWarnings(as.numeric(hospitalised))
  )

kraemer_contacts <- read_kraemer_long("insp_sitrep__cumulative_contacts_traced") |>
  mutate(
    zone            = recode(nom, !!!KRAEMER_NAME_MAP),
    date            = parse_kraemer_date(date),
    contacts_traced = suppressWarnings(as.numeric(cumulative_contacts_traced))
  ) |>
  filter(!is.na(contacts_traced))

# Response indicators (hospitalisations by zone)
kraemer_response <- read_kraemer_long("insp_sitrep__hospitalised") |>
  full_join(read_kraemer_long("insp_sitrep__in_bed_previous_day"), by = c("nom", "date")) |>
  full_join(
    read_kraemer_long("insp_sitrep__new_hosp_admissions") |>
      rename(total_admissions = new_all_admissions),
    by = c("nom", "date")
  ) |>
  full_join(
    read_kraemer_long("insp_sitrep__new_hosp_detainees") |>
      rename(new_admissions_ppl = new_hosp_detainees),
    by = c("nom", "date")
  ) |>
  full_join(
    read_kraemer_long("insp_sitrep__new_hosp_other") |>
      rename(new_admissions_other = new_other),
    by = c("nom", "date")
  ) |>
  mutate(
    zone   = recode(nom, !!!KRAEMER_NAME_MAP),
    date   = parse_kraemer_date(date),
    across(c(hospitalised, in_bed_previous_day, total_admissions,
             new_admissions_ppl, new_admissions_other),
           ~suppressWarnings(as.numeric(.x))),
    source = "INRB-UMIE"
  ) |>
  filter(zone %in% ZONE_LEVELS)

# POE (checkpoint-level in INRB-UMIE; summed to daily aggregate for comparison)
kraemer_poe <- read_kraemer_long("insp_sitrep__total_poe_passed") |>
  rename(total_passed = total_poe_passed) |>
  full_join(
    read_kraemer_long("insp_sitrep__total_poe_screened") |>
      rename(total_screened = total_poe_screened),
    by = c("nom", "date")
  ) |>
  full_join(
    read_kraemer_long("insp_sitrep__total_poe_hand_washing") |>
      rename(total_handwashing = total_poe_hand_washing),
    by = c("nom", "date")
  ) |>
  full_join(
    read_kraemer_long("insp_sitrep__total_poe_sanitised") |>
      rename(total_sensitised = total_poe_sanitised),
    by = c("nom", "date")
  ) |>
  mutate(
    date = parse_kraemer_date(date),
    across(c(total_passed, total_screened, total_handwashing, total_sensitised),
           ~suppressWarnings(as.numeric(.x)))
  ) |>
  group_by(date) |>
  summarise(
    across(c(total_passed, total_screened, total_handwashing, total_sensitised),
           ~sum(.x, na.rm = TRUE)),
    .groups = "drop"
  ) |>
  mutate(source = "INRB-UMIE")
```

```{r}
#| label: load-bvd-reference
# INRB-UMIE national totals (single snapshot: 27 May 2026)
read_kraemer_national_snapshot <- function(filename, value_col) {
  read_kraemer_long(filename) |>
    rename(value = all_of(value_col)) |>
    mutate(
      date  = parse_kraemer_date(date),
      value = suppressWarnings(as.numeric(value))
    ) |>
    filter(!is.na(date)) |>
    group_by(date) |>
    summarise(value = first(value), .groups = "drop")
}

parsed_kraemer_nat <-
  read_kraemer_national_snapshot(
    "insp_sitrep__national_cumulative_suspected_cases",
    "national_cumulative_suspected_cases"
  ) |> rename(kraemer_suspected = value) |>
  full_join(
    read_kraemer_national_snapshot(
      "insp_sitrep__national_cumulative_confirmed_cases",
      "national_cumulative_confirmed_cases"
    ) |> rename(kraemer_confirmed = value),
    by = "date"
  ) |>
  full_join(
    read_kraemer_national_snapshot(
      "insp_sitrep__national_cumulative_suspected_deaths",
      "national_cumulative_suspected_deaths"
    ) |> rename(kraemer_susp_deaths = value),
    by = "date"
  ) |>
  full_join(
    read_kraemer_national_snapshot(
      "insp_sitrep__national_cumulative_confirmed_deaths",
      "national_cumulative_confirmed_deaths"
    ) |> rename(kraemer_conf_deaths = value),
    by = "date"
  )

# Pull our national aggregate Cumules rows
# Priority: province=="Total" > zone=="Total" > both NA
ous_national <- df |>
  filter(
    count_type == "Cumules",
    toupper(is_aggregate) == "TRUE",
    (zone == "Total" | province == "Total" | (is.na(zone) & is.na(province)))
  ) |>
  mutate(
    row_priority = case_when(
      province == "Total" ~ 3L,
      zone == "Total"     ~ 2L,
      TRUE                ~ 1L
    )
  ) |>
  group_by(date) |>
  slice_max(order_by = row_priority, n = 1, with_ties = FALSE) |>
  ungroup() |>
  transmute(date, sitrep_source,
            ours_suspected   = suspect,
            ours_confirmed   = confirmed,
            ours_deaths      = deaths,
            ours_deaths_susp = deaths_susp,
            ours_deaths_conf = deaths_conf)


# Join our extraction with INRB-UMIE national totals
nat_comparison <- ous_national |>
  full_join(parsed_kraemer_nat, by = "date") |>
  arrange(date)
```

::: {.callout-warning}
**Automated extraction — verification required.**
Case counts and figures in this report are extracted from INSP DRC PDF situation reports by an AI vision model (Anthropic Claude `claude-sonnet-4-6`) without systematic manual verification of every value. Extraction errors are possible, particularly where PDF table layouts are complex or inconsistent. All values should be verified against the [original INSP DRC situation reports](https://insp.cd/ebola/) before any operational, clinical, or policy use. This report is intended for research and situational awareness only.
:::

## Introduction

This report documents automated extraction of epidemiological data from `r n_sitreps` situation reports (SitReps) published by the Institut National de Santé Publique (INSP DRC) during the 2026 Ebola (MVE) outbreak in the Democratic Republic of Congo, covering the period through `r latest_date`. The focus is on data completeness, extraction fidelity, and cross-source consistency rather than epidemiological interpretation.

## Methods

### Data sources

Case counts were taken from the INSP DRC situation report PDFs, downloaded automatically from <https://insp.cd/ebola/>. Where our PDF copies and the INRB-UMIE copies overlap, file checksums confirm they are identical, so any differences between the two datasets come from how the tables were read rather than which documents were used. The INRB-UMIE series, produced by manual data entry, was used as an independent check on our automated extraction.

### Automated PDF extraction

Each PDF is read by Claude (Anthropic, `claude-sonnet-4-6`) using visual understanding of the page layout, allowing it to handle the complex table structures accurately. Four datasets are extracted from each report: daily new case counts, cumulative case and death totals, patient movement through isolation units (available from SitRep 006 onwards), and Point of Entry screening activity. A full text transcript is also saved for audit purposes.

### Data harmonisation

Health zone names were standardised to a consistent spelling across all reports. Suspected and probable case counts were summed into a single "unconfirmed" category, since the column labels vary across reports. Cumulative totals were not permitted to decrease over time, and where more than one row existed for the same zone and date the higher value was retained.

### Statistical analysis

All analysis and figures were produced in R (version `r R.version$major`.`r R.version$minor`) using the `tidyverse` [@tidyverse], `ggplot2` [@ggplot2], and `flextable` [@flextable] packages. This report was rendered automatically with Quarto.

### Cross-source validation

We checked our automated extraction against the [INRB-UMIE Ebola DRC 2026](https://github.com/INRB-UMIE/Ebola_DRC_2026) manually coded dataset, which extends through SitRep 22 (6 June 2026) in the current build. For SitReps 1–11 both datasets use the same PDF files, so differences reflect extraction method rather than source documents. From SitRep 12 onwards INRB-UMIE note that health-zone data is occasionally sent directly from INSP to INRB and may not feature in the published PDFs; this explains divergence at zone level from SitRep 12 onwards. Two zone name variants (Mongbwalu/Mongbalu and Nyankunde/Nyakunde) were matched manually; values marked "ND" were treated as missing.

## Results

The following sections compare our automated extraction against the INRB-UMIE manually coded dataset and document known data quality issues in the source PDFs.

### National-level comparison — this repo and INRB-UMIE

- **Cases agree across sources.** Cumulative suspected and confirmed case counts match exactly for every date in the overlap period; divergences are confined to the deaths series.
- **Banner vs sub-table discrepancy (SitRep 010 onwards).** INSP's front-page banner and the per-zone sub-tables frequently disagree on national totals. This repo and INRB-UMIE (`national_*` files) both read from the banner. The disagreement reflects a structural feature of the PDFs.
- **SitRep 012 (26 May) — two PDF versions.** The original extraction records 238 deaths; the revised version (SitRep\_012\_v2) records 246 deaths. No per-zone case table was present in either version to cross-check against the banner.
- **INRB-UMIE national coverage.** The current build (`build-2026-06-07`) includes national banner figures from SitRep 004 (14 May) through SitRep 022 (6 June 2026), and is consistent with this repo across all overlapping dates.

```{r}
#| label: tbl-national-comparison
#| tbl-cap: "National cumulative totals — cross-source comparison (this repo vs INRB-UMIE). Shading indicates values that differ. INRB-UMIE: INRB-UMIE/Ebola\_DRC\_2026. A dash indicates no data available from that source for that date."
nat_comparison |>
  transmute(
    Date                         = format(date, "%d %b"),
    `Suspect. (ours)`            = ours_suspected,
    `Suspect. (INRB-UMIE)`      = kraemer_suspected,
    `Confirmed (ours)`           = ours_confirmed,
    `Confirmed (INRB-UMIE)`     = kraemer_confirmed,
    `Deaths (ours)`              = ours_deaths,
    `Deaths susp. (INRB-UMIE)`  = kraemer_susp_deaths,
    `Deaths conf. (INRB-UMIE)`  = kraemer_conf_deaths
  ) |>
  flextable() |>
  set_header_labels(Date = "Date") |>
  add_header_row(
    values    = c("", "Suspected cases", "Confirmed cases", "Deaths"),
    colwidths = c(1, 2, 2, 3)
  ) |>
  merge_h(part = "header") |>
  align(align = "center", part = "header") |>
  align(j = -1, align = "right", part = "body") |>
  bg(
    i = ~ !is.na(`Deaths (ours)`) & !is.na(`Deaths susp. (INRB-UMIE)`) &
         `Deaths (ours)` != `Deaths susp. (INRB-UMIE)`,
    j = c("Deaths (ours)", "Deaths susp. (INRB-UMIE)"),
    bg = "#FFF3CD"
  ) |>
  colformat_num(na_str = "—") |>
  colformat_int(j = 2:8, na_str = "—") |>
  bold(part = "header") |>
  hline(part = "body") |>
  autofit()
```

### Data quality notes — case counts

- **Mongbwalu zone-level data starts late.** SitRep 001 records 246 suspected cases for all of Ituri province combined; no Mongbwalu-specific count appears until SitRep 004 (18 May, 302 suspected). The Mongbwalu line therefore starts later than in INRB-UMIE — this is a reporting gap, not a case-onset gap.
- **Mongbwalu confirmed-case revision (22 May).** Both datasets show cumulative confirmed cases falling from 14 (21 May) to 10 (22 May). Our series holds the value at 14 (counts cannot decrease); INRB-UMIE shows the raw figure. This is likely a downward revision in the source PDF.
- **SitRep 014 (28 May) — no zone breakdown.** Only a national aggregate was published during an official case-count revision. The dotted vertical line in zone panels marks this gap.
- **SitRep 016 (30 May) — province-level only.** Zone-level breakdowns return in SitRep 015 (29 May) but are absent again in SitRep 016, which reports only Ituri / Nord-Kivu / Sud-Kivu provincial aggregates.
- **SitRep 009 (23 May) — deaths dip.** Cumulative suspected deaths fell from 160 (20 May) to 119 before recovering to 223 in SitRep 010. This is reproduced faithfully from the PDF and likely reflects an interim partial recount.

**Suspect and probable case reporting discontinued.** From SitRep 015 (29 May) onwards, INSP stopped publishing cumulative suspect/probable breakdowns in the zone tables. From SitRep 016 (30 May) onwards they are absent from both cumulative and daily new-cases tables at all geographic levels. SitRep 014 stated explicitly that suspected deaths were temporarily excluded pending investigation results (*"Les décès suspects ont été temporairement exclus du comptage dans l'attente des résultats des investigations en cours"*). The orange-shaded region in the national "Suspected deaths" panel marks this ongoing series break.

**INRB-UMIE suspected cases after SitRep 016 are not cumulative.** From SitRep 016 onwards, INSP's banner split suspect reporting into 'Cas suspects en cours d'investigation' (under investigation) and 'Cas suspects en isolement' (in isolation). INRB-UMIE sum these two active categories as their national suspected case figure. This means the INRB-UMIE suspected case series from SitRep 016 onwards reflects **active suspects at a point in time**, not cumulative ever-suspected — explaining the declining trend.

### Daily new cases (national)

```{r}
#| label: fig-daily-new
#| fig-cap: "Daily new cases reported nationally by case classification (from 'Nouveaux' rows in each SitRep). Suspect/probable cases absent from SitRep 016 onwards."
#| fig-width: 9
#| fig-height: 4
daily_new |>
  transmute(date,
            `Confirmed`        = confirmed,
            `Suspect/probable` = unconfirmed) |>
  pivot_longer(-date, names_to = "type", values_to = "n") |>
  filter(!is.na(n)) |>
  mutate(type = factor(type, levels = c("Suspect/probable", "Confirmed"))) |>
  ggplot(aes(x = date, y = n, fill = type)) +
  geom_col(position = "stack", width = 0.9) +
  scale_fill_manual(values = c("Suspect/probable" = "#56B4E9",
                               "Confirmed"        = "#D55E00")) +
  scale_x_date(date_labels = "%d %b", date_breaks = "3 days") +
  scale_y_continuous(breaks = int_breaks, expand = expansion(mult = c(0, 0.1))) +
  labs(x = NULL, y = "Daily new cases", fill = NULL,
       title = "MVE DRC — Daily new cases (national aggregate)") +
  theme_sitrep() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Cumulative counts by zone and source

```{r}
#| label: fig-comparison
#| fig-cap: "Cross-source comparison of cumulative confirmed cases, suspect/probable cases, and outbreak deaths by health zone. Solid lines: automated extraction (raw); dashed lines: INRB-UMIE (manual extraction). For SitReps 1–11 both sources use the same PDFs; from SitRep 12 onwards INRB-UMIE zone data may include figures sent directly from INSP to INRB that do not appear in the published PDFs. From SitRep 16 onwards INRB-UMIE national suspected cases reflect active suspects (under investigation + in isolation), not cumulative totals."
#| fig-width: 12
#| fig-height: 14
PAL_COMP <- c(
  "Automated extraction" = "#0072B2",
  "INRB-UMIE"            = "#E69F00"
)
LTY_COMP <- c(
  "Automated extraction" = "solid",
  "INRB-UMIE"            = "dashed"
)
POINT_SIZE_OURS <- 2.2
POINT_SIZE_KRAEMER <- 1.7

zone_province_lut <- zone_ts |>
  distinct(zone, province) |>
  mutate(zone = as.character(zone))

our_for_comp <- zone_ts |>
  transmute(zone = as.character(zone), province, date,
            confirmed   = confirmed_raw,
            suspected   = unconfirmed_raw,
            deaths_susp = deaths_susp_raw,
            deaths_conf = deaths_conf,
            source      = "Automated extraction")

# National totals — ours (raw extracted values, no cummax enforcement)
our_nat_for_comp <- ous_national |>
  transmute(zone     = "National",
            province = "National",
            date,
            confirmed   = ours_confirmed,
            suspected   = ours_suspected,
            deaths_susp = ours_deaths_susp,
            deaths_conf = ours_deaths_conf,
            source      = "Automated extraction")

# INRB-UMIE deaths_susp = suspected only; deaths = suspected + confirmed (total)
kraemer_for_comp <- kraemer |>
  left_join(zone_province_lut, by = "zone") |>
  transmute(zone, province, date, confirmed, suspected,
            deaths_susp, deaths_conf,
            source)

# INRB-UMIE national totals
kraemer_nat_for_comp <- parsed_kraemer_nat |>
  transmute(zone     = "National",
            province = "National",
            date,
            confirmed   = kraemer_confirmed,
            suspected   = kraemer_suspected,
            deaths_susp = kraemer_susp_deaths,
            deaths_conf = kraemer_conf_deaths,
            source      = "INRB-UMIE")

comp_long <- bind_rows(our_nat_for_comp, kraemer_nat_for_comp,
                       our_for_comp, kraemer_for_comp) |>
  pivot_longer(c(confirmed, suspected, deaths_susp, deaths_conf),
               names_to = "metric", values_to = "count") |>
  mutate(
    metric = recode(metric,
      "confirmed"   = "Confirmed cases",
      "suspected"   = "Suspect/probable cases",
      "deaths_susp" = "Suspected deaths",
      "deaths_conf" = "Confirmed deaths"
    ),
    metric   = factor(metric, levels = c("Confirmed cases", "Suspect/probable cases",
                                          "Suspected deaths", "Confirmed deaths")),
    source   = factor(source, levels = c("Automated extraction", "INRB-UMIE")),
    province = factor(province, levels = c("National", "Ituri", "Nord-Kivu")),
    zone     = factor(zone, levels = c("National", ZONE_LEVELS))
  ) |>
  filter(!is.na(count))

comp_long_nat   <- filter(comp_long, zone == "National")
comp_long_zones <- filter(comp_long, zone != "National")

base_theme <- list(
  scale_colour_manual(values = PAL_COMP),
  scale_linetype_manual(values = LTY_COMP),
  scale_x_date(date_labels = "%d %b", date_breaks = "2 days"),
  scale_y_continuous(breaks = int_breaks, limits = c(0, NA),
                     expand = expansion(mult = c(0, 0.15))),
  theme_sitrep(base_size = 9),
  theme(
    axis.text.x  = element_text(angle = 45, hjust = 1, size = 7),
    panel.border = element_rect(colour = "grey70", fill = NA, linewidth = 0.4),
    panel.spacing = unit(0.6, "lines")
  )
)

p_nat <- ggplot(comp_long_nat,
                aes(x = date, y = count, colour = source, linetype = source)) +
  # Shaded region: suspected deaths absent from INSP PDFs (SitRep 014+)
  geom_rect(
    data = tibble(metric = factor("Suspected deaths",
                                  levels = levels(comp_long_nat$metric)),
                  xmin = as.Date("2026-05-28"), xmax = Inf,
                  ymin = -Inf, ymax = Inf),
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
    fill = "orange", alpha = 0.12, inherit.aes = FALSE
  ) +
  geom_text(
    data = tibble(metric = factor("Suspected deaths",
                                  levels = levels(comp_long_nat$metric)),
                  x = as.Date("2026-05-28"), y = Inf,
                  label = "Suspected deaths\nabsent SitRep 014+"),
    aes(x = x, y = y, label = label),
    hjust = -0.04, vjust = 1.1, size = 1.8, colour = "darkorange3",
    inherit.aes = FALSE
  ) +
  # Shaded region: INRB-UMIE switches to active suspects (SitRep 016+)
  geom_rect(
    data = tibble(metric = factor("Suspect/probable cases",
                                  levels = levels(comp_long_nat$metric)),
                  xmin = as.Date("2026-05-30"), xmax = Inf,
                  ymin = -Inf, ymax = Inf),
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
    fill = "purple", alpha = 0.08, inherit.aes = FALSE
  ) +
  geom_text(
    data = tibble(metric = factor("Suspect/probable cases",
                                  levels = levels(comp_long_nat$metric)),
                  x = as.Date("2026-05-30"), y = Inf,
                  label = "INRB-UMIE: active\nsuspects only\n(SitRep 016+)"),
    aes(x = x, y = y, label = label),
    hjust = -0.04, vjust = 1.1, size = 1.8, colour = "purple4",
    inherit.aes = FALSE
  ) +
  geom_line(linewidth = 0.9, na.rm = TRUE) +
  geom_point(data = ~filter(.x, source == "Automated extraction"),
             size = POINT_SIZE_OURS,    na.rm = TRUE, show.legend = FALSE) +
  geom_point(data = ~filter(.x, source == "INRB-UMIE"),
             size = POINT_SIZE_KRAEMER, na.rm = TRUE, show.legend = FALSE) +
  facet_wrap(~ metric, ncol = 1, scales = "free_y") +
  labs(x = NULL, y = "Cumulative count", colour = NULL, linetype = NULL,
       title = "National") +
  base_theme

zones_label_pt <- comp_long_zones |>
  filter(date == as.Date("2026-05-28"),
         metric == "Confirmed cases",
         zone   == "Mongbwalu",
         source == "INRB-UMIE") |>
  slice(1)

p_zones <- ggplot(comp_long_zones,
                  aes(x = date, y = count, colour = source, linetype = source)) +
  geom_line(linewidth = 0.9, na.rm = TRUE) +
  geom_point(data = ~filter(.x, source == "Automated extraction"),
             size = POINT_SIZE_OURS,    na.rm = TRUE, show.legend = FALSE) +
  geom_point(data = ~filter(.x, source == "INRB-UMIE"),
             size = POINT_SIZE_KRAEMER, na.rm = TRUE, show.legend = FALSE) +
  ggrepel::geom_text_repel(
    data            = zones_label_pt,
    aes(x = date, y = count, label = "SitRep 014:\nno PDF zone data"),
    colour          = "red3",
    size            = 1.9,
    fontface        = "plain",
    inherit.aes     = FALSE,
    nudge_x         = -2,
    nudge_y         = 8,
    segment.colour  = "red3",
    segment.size    = 0.3,
    min.segment.len = 0,
    box.padding     = 0.2,
    show.legend     = FALSE
  ) +
  facet_nested(metric ~ province + zone, scales = "free_y") +
  labs(x = NULL, y = NULL, colour = NULL, linetype = NULL,
       title = "By health zone") +
  base_theme +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())

p <- (p_nat | p_zones) +
  plot_layout(widths = c(1, 5), guides = "collect") +
  plot_annotation(
    title    = "MVE DRC \u2014 Cross-source validation",
    subtitle = "Cumulative counts: automated extraction (raw) vs INRB-UMIE (manual extraction; NB: INRB-UMIE zone data diverges from published PDFs from SitRep 012 onwards)",
    caption  = "Sources: INSP DRC situation reports; github.com/INRB-UMIE/Ebola_DRC_2026",
    theme    = theme_sitrep(base_size = 9)
  ) &
  theme(legend.position = "bottom")
ggsave(here("outputs", "plots", "fig-comparison.png"), plot = p, width = 12, height = 14, dpi = 300)
p
```

@tbl-comparison summarises numerical differences at the most recent date available in each source per zone.

```{r}
#| label: tbl-comparison
#| tbl-cap: "Cumulative case counts at the latest available date per zone: automated extraction vs INRB-UMIE. \u0394 = ours minus INRB-UMIE. Dashes indicate no data in that source. Note: INRB-UMIE zone data from SitRep 12 onwards diverges from the published PDFs (exact supplementary source not documented in their public repository)."
ours_latest <- zone_ts |>
  group_by(zone) |>
  slice_max(date, n = 1) |>
  ungroup() |>
  transmute(zone, conf_ours = confirmed, susp_ours = unconfirmed, dead_ours = deaths)

kraemer_latest <- kraemer |>
  group_by(zone) |>
  slice_max(date, n = 1) |>
  ungroup() |>
  transmute(zone, conf_them = confirmed, susp_them = suspected, dead_them = deaths)

full_join(ours_latest, kraemer_latest, by = "zone") |>
  mutate(
    d_conf = conf_ours - conf_them,
    d_susp = susp_ours - susp_them,
    d_dead = dead_ours - dead_them
  ) |>
  arrange(zone) |>
  transmute(
    zone,
    conf_ours, conf_them, d_conf,
    susp_ours, susp_them, d_susp,
    dead_ours, dead_them, d_dead
  ) |>
  setNames(c(
    "Health zone",
    "Confirmed (ours)", "Confirmed (INRB-UMIE)", "\u0394 confirmed",
    "Suspect/prob. (ours)", "Suspect/prob. (INRB-UMIE)", "\u0394 suspect/prob.",
    "Deaths (ours)", "Deaths (INRB-UMIE)", "\u0394 deaths"
  )) |>
  flextable() |>
  bold(part = "header") |>
  bg(part = "header", bg = "#1A3A5C") |>
  color(part = "header", color = "white") |>
  align(j = 2:10, align = "right", part = "all") |>
  colformat_int(j = 2:10, na_str = "\u2014") |>
  autofit()
```

### Response indicators

```{r}
#| label: load-response-ours
resp_ours <- read_csv(
  here::here("data", "processed", "master_response_counts.csv"),
  col_types      = cols(.default = "c"),
  show_col_types = FALSE
) |>
  mutate(
    date   = dmy(date),
    zone   = factor(zone, levels = ZONE_LEVELS),
    across(c(in_bed_previous_day, new_admissions_ppl, new_admissions_other,
             total_admissions, total_isolated, cumulative_admitted),
           ~suppressWarnings(as.numeric(.x))),
    source = "Automated extraction"
  ) |>
  filter(!is.na(zone), !zone %in% c("Total"))

poe_ours <- read_csv(
  here::here("data", "processed", "master_poe_counts.csv"),
  col_types      = cols(.default = "c"),
  show_col_types = FALSE
) |>
  mutate(
    date = dmy(date),
    across(c(total_passed, total_screened, total_handwashing, total_sensitised),
           ~suppressWarnings(as.numeric(.x))),
    source = "Automated extraction"
  ) |>
  select(date, total_passed, total_screened, total_handwashing, total_sensitised, source)
```

### Data quality notes — response indicators

- **Rwampara 21 May (SitRep 007).** Data are split across three unlabelled sub-columns; both datasets handle these differently. Values for that date should be treated with caution.
- **20–21 May (SitReps 005–006).** An earlier table layout made in-bed count and new admissions columns ambiguous; minor differences between datasets on those dates result from this. Agreement is close from 22 May onwards when the format was standardised.
- **Coverage.** INRB-UMIE (`build-2026-06-07`) extends through 6 June 2026; our extraction covers through SitRep 023 (7 June 2026). Validation panels overlap through 6 June 2026.

```{r}
#| label: fig-comparison-response
#| fig-cap: "Cross-source comparison of zone-level response indicators. Solid lines: automated extraction; dashed lines: INRB-UMIE. Both series treat the ambiguous 21 May Rwampara sub-columns differently; values for that date should be interpreted cautiously. The two datasets differ slightly on 20–21 May due to an ambiguous table layout in SitReps 005–006, and are in close agreement from 22 May onwards."
#| fig-width: 12
#| fig-height: 9
PAL_RESP <- c("Automated extraction" = "#0072B2", "INRB-UMIE" = "#E69F00")
LTY_RESP <- c("Automated extraction" = "solid",   "INRB-UMIE" = "dashed")

resp_comp <- bind_rows(
  resp_ours |>
    transmute(date, zone,
              hospitalised        = total_isolated,
              in_bed_previous_day,
              total_admissions,
              source),
  kraemer_response |>
    transmute(date, zone = factor(zone, levels = ZONE_LEVELS),
              hospitalised,
              in_bed_previous_day,
              total_admissions,
              source)
) |>
  pivot_longer(c(hospitalised, in_bed_previous_day, total_admissions),
               names_to = "metric", values_to = "count") |>
  mutate(
    metric = recode(metric,
      "hospitalised"        = "Currently hospitalised",
      "in_bed_previous_day" = "In bed (prev. day)",
      "total_admissions"    = "New admissions (total)"
    ),
    metric = factor(metric, levels = c("Currently hospitalised",
                                        "In bed (prev. day)",
                                        "New admissions (total)")),
    source = factor(source, levels = c("Automated extraction", "INRB-UMIE"))
  ) |>
  filter(!is.na(count), !is.na(zone))

p <- ggplot(resp_comp, aes(x = date, y = count, colour = source, linetype = source)) +
  geom_line(linewidth = 0.9, na.rm = TRUE) +
  geom_point(data = ~filter(.x, source == "Automated extraction"), size = POINT_SIZE_OURS, na.rm = TRUE, show.legend = FALSE) +
  geom_point(data = ~filter(.x, source == "INRB-UMIE"), size = POINT_SIZE_KRAEMER, na.rm = TRUE, show.legend = FALSE) +
  facet_nested(metric ~ zone, scales = "free_y") +
  scale_colour_manual(values = PAL_RESP) +
  scale_linetype_manual(values = LTY_RESP) +
  scale_x_date(date_labels = "%d %b", date_breaks = "2 days") +
  scale_y_continuous(breaks = int_breaks, limits = c(0, NA),
                     expand = expansion(mult = c(0, 0.15))) +
  labs(
    title    = "MVE DRC \u2014 Response indicators — cross-source validation",
    subtitle = "Automated extraction vs INRB-UMIE (manual, same source PDFs)",
    x = NULL, y = "Count", colour = NULL, linetype = NULL,
    caption  = "Sources: INSP DRC situation reports; github.com/INRB-UMIE/Ebola_DRC_2026"
  ) +
  theme_sitrep(base_size = 9) +
  theme(
    axis.text.x  = element_text(angle = 45, hjust = 1, size = 7),
    panel.border = element_rect(colour = "grey70", fill = NA, linewidth = 0.4),
    panel.spacing = unit(0.5, "lines")
  )
ggsave(here("outputs", "plots", "fig-comparison-response.png"), plot = p,
       width = 12, height = 9, dpi = 300)
p
```

```{r}
#| label: fig-comparison-poe
#| fig-cap: "Cross-source comparison of daily Point of Entry (POE) screening totals. Automated extraction reports a single aggregate per sitrep date; INRB-UMIE data are summed across individual checkpoints to produce a daily total."
#| fig-width: 9
#| fig-height: 6
poe_comp <- bind_rows(poe_ours, kraemer_poe) |>
  pivot_longer(c(total_passed, total_screened, total_handwashing, total_sensitised),
               names_to = "metric", values_to = "count") |>
  mutate(
    metric = recode(metric,
      "total_passed"      = "Total passed",
      "total_screened"    = "Screened",
      "total_handwashing" = "Handwashing stations",
      "total_sensitised"  = "Sensitised"
    ),
    metric = factor(metric, levels = c("Total passed", "Screened",
                                        "Handwashing stations", "Sensitised")),
    source = factor(source, levels = c("Automated extraction", "INRB-UMIE"))
  ) |>
  filter(!is.na(count))

p <- ggplot(poe_comp, aes(x = date, y = count, colour = source, linetype = source)) +
  geom_line(linewidth = 0.9, na.rm = TRUE) +
  geom_point(data = ~filter(.x, source == "Automated extraction"), size = POINT_SIZE_OURS + 0.3, na.rm = TRUE, show.legend = FALSE) +
  geom_point(data = ~filter(.x, source == "INRB-UMIE"), size = POINT_SIZE_KRAEMER + 0.3, na.rm = TRUE, show.legend = FALSE) +
  facet_wrap(~ metric, scales = "free_y", ncol = 2) +
  scale_colour_manual(values = PAL_RESP) +
  scale_linetype_manual(values = LTY_RESP) +
  scale_x_date(date_labels = "%d %b", date_breaks = "2 days") +
  scale_y_continuous(breaks = int_breaks, limits = c(0, NA),
                     expand = expansion(mult = c(0, 0.12))) +
  labs(
    title    = "MVE DRC \u2014 POE screening — cross-source validation",
    subtitle = paste0("Daily aggregate totals: automated extraction vs INRB-UMIE (INRB-UMIE sums across individual checkpoints; extends to ",
                       format(max(kraemer_poe$date, na.rm = TRUE), "%d %b %Y"), ")"),
    x = NULL, y = "Count", colour = NULL, linetype = NULL,
    caption  = "Sources: INSP DRC situation reports; github.com/INRB-UMIE/Ebola_DRC_2026"
  ) +
  theme_sitrep(base_size = 11) +
  theme(
    axis.text.x  = element_text(angle = 45, hjust = 1, size = 8),
    panel.border = element_rect(colour = "grey70", fill = NA, linewidth = 0.4),
    panel.spacing = unit(0.6, "lines")
  )
ggsave(here("outputs", "plots", "fig-comparison-poe.png"), plot = p,
       width = 9, height = 6, dpi = 300)
p
```

```{r}
#| label: fig-comparison-contacts
#| fig-cap: "Cross-source comparison of cumulative contacts traced by health zone. Solid lines: automated extraction (raw); dashed lines: INRB-UMIE (manual extraction). ND values excluded."
#| fig-width: 11
#| fig-height: 7
contacts_ours <- df |>
  filter(
    count_type == "Cumules",
    zone %in% ZONE_LEVELS,
    !is.na(contacts), contacts != ""
  ) |>
  mutate(contacts = suppressWarnings(as.numeric(contacts)),
         zone = factor(zone, levels = ZONE_LEVELS)) |>
  filter(!is.na(contacts)) |>
  arrange(zone, date) |>
  group_by(zone) |>
  ungroup() |>
  transmute(date, zone, contacts, source = "Automated extraction")

contacts_kraemer <- kraemer_contacts |>
  filter(zone %in% ZONE_LEVELS) |>
  transmute(date, zone = factor(zone, levels = ZONE_LEVELS),
            contacts = contacts_traced, source = "INRB-UMIE")

contacts_comp <- bind_rows(contacts_ours, contacts_kraemer) |>
  filter(!is.na(contacts)) |>
  left_join(zone_province_lut, by = "zone") |>
  mutate(
    source   = factor(source, levels = c("Automated extraction", "INRB-UMIE")),
    province = factor(province, levels = c("Ituri", "Nord-Kivu")),
    zone     = factor(zone, levels = ZONE_LEVELS)
  )

p <- ggplot(contacts_comp,
            aes(x = date, y = contacts, colour = source, linetype = source)) +
  geom_line(linewidth = 0.9, na.rm = TRUE) +
  geom_point(data = ~filter(.x, source == "Automated extraction"), size = POINT_SIZE_OURS, na.rm = TRUE, show.legend = FALSE) +
  geom_point(data = ~filter(.x, source == "INRB-UMIE"), size = POINT_SIZE_KRAEMER, na.rm = TRUE, show.legend = FALSE) +
  facet_nested(~ province + zone, scales = "free_y") +
  scale_colour_manual(values = PAL_COMP[c("Automated extraction","INRB-UMIE")]) +
  scale_linetype_manual(values = LTY_COMP[c("Automated extraction","INRB-UMIE")]) +
  scale_x_date(date_labels = "%d %b", date_breaks = "2 days") +
  scale_y_continuous(breaks = int_breaks, limits = c(0, NA),
                     expand = expansion(mult = c(0, 0.15))) +
  labs(
    title    = "MVE DRC \u2014 Cumulative contacts traced — cross-source validation",
    subtitle = "Automated extraction (raw) vs INRB-UMIE (manual extraction, same source PDFs)",
    x = NULL, y = "Cumulative contacts traced", colour = NULL, linetype = NULL,
    caption  = "Sources: INSP DRC situation reports; github.com/INRB-UMIE/Ebola_DRC_2026"
  ) +
  theme_sitrep(base_size = 9) +
  theme(
    axis.text.x  = element_text(angle = 45, hjust = 1, size = 7),
    panel.border = element_rect(colour = "grey70", fill = NA, linewidth = 0.4),
    panel.spacing = unit(0.5, "lines")
  )
ggsave(here("outputs", "plots", "fig-comparison-contacts.png"), plot = p,
       width = 11, height = 7, dpi = 300)
p
```

## Discussion

**Outbreak trajectory**

- `r n_conf` confirmed cases have been reported nationally as of `r latest_date`.
- The epidemic is concentrated in Ituri province (Bunia, Rwampara, Mongbwalu), with a smaller number of cases in Nord-Kivu and Sud-Kivu.
- INSP discontinued cumulative suspect/probable reporting from SitRep 015 onwards, limiting assessment of total burden and complicating comparisons with earlier reports.

**Limitations**

- Zone-level breakdowns are absent or incomplete in SitReps 001–013 and in SitReps 014 and 016; the zone-level series is therefore not fully continuous.
- Response indicator table layouts varied across early reports, introducing minor inconsistencies in trend data.
- Automated extraction and INRB-UMIE manual coding provide complementary checks: agreement increases confidence in a figure; disagreement typically reflects genuine ambiguity in the source document.

## Appendix — Internal consistency check

#### Zone sum vs national aggregate

Cumulative zone-row sums are compared with the national banner figure for each date. Highlighted rows (non-zero Δ) arise from three causes:

- Early reports (SitReps 001–003) provide only a province-level Ituri total, not individual zone rows.
- Some reports omit certain zones from the breakdown table.
- The banner vs sub-table discrepancy present from SitRep 010 onwards (described above).

```{r}
#| label: tbl-zone-sum-check
#| tbl-cap: "Zone sum vs national aggregate. Values show the cumulative total across all extracted health zones versus the national figure from the front-page banner. Δ = zone sum minus national banner. Highlighted rows have a non-zero difference in at least one metric."
zone_sum_vs_nat <- zone_ts |>
  group_by(date) |>
  summarise(
    sum_susp = if (any(!is.na(unconfirmed_raw))) sum(unconfirmed_raw, na.rm = TRUE) else NA_real_,
    sum_conf = if (any(!is.na(confirmed_raw)))   sum(confirmed_raw,   na.rm = TRUE) else NA_real_,
    sum_dths = if (any(!is.na(deaths_raw)))       sum(deaths_raw,      na.rm = TRUE) else NA_real_,
    .groups  = "drop"
  ) |>
  left_join(
    ous_national |> select(date,
                            nat_susp = ours_suspected,
                            nat_conf = ours_confirmed,
                            nat_dths = ours_deaths),
    by = "date"
  ) |>
  mutate(
    d_susp = sum_susp - nat_susp,
    d_conf = sum_conf - nat_conf,
    d_dths = sum_dths - nat_dths
  ) |>
  arrange(date)

discrepant_rows <- which(
  (!is.na(zone_sum_vs_nat$d_susp) & zone_sum_vs_nat$d_susp != 0) |
  (!is.na(zone_sum_vs_nat$d_conf) & zone_sum_vs_nat$d_conf != 0) |
  (!is.na(zone_sum_vs_nat$d_dths) & zone_sum_vs_nat$d_dths != 0)
)

zone_sum_vs_nat |>
  transmute(
    Date             = format(date, "%d %b"),
    `Zones (susp.)`  = sum_susp,
    `Nat. (susp.)`   = nat_susp,
    `Δ susp.`        = d_susp,
    `Zones (conf.)`  = sum_conf,
    `Nat. (conf.)`   = nat_conf,
    `Δ conf.`        = d_conf,
    `Zones (deaths)` = sum_dths,
    `Nat. (deaths)`  = nat_dths,
    `Δ deaths`       = d_dths
  ) |>
  flextable() |>
  add_header_row(
    values    = c("", "Suspected", "Confirmed", "Deaths"),
    colwidths = c(1, 3, 3, 3)
  ) |>
  merge_h(part = "header") |>
  align(align = "center", part = "header") |>
  align(j = -1, align = "right", part = "body") |>
  bg(i = discrepant_rows, bg = "#FFF3CD") |>
  colformat_int(j = 2:10, na_str = "—") |>
  bold(part = "header") |>
  hline(part = "body") |>
  autofit()
```
