---
title: "Maladie à Virus Ebola — Situation Report"
subtitle: "Democratic Republic of Congo"
date: today
date-format: "DD MMMM YYYY"
format:
  html:
    toc: true
    toc-depth: 2
    toc-title: "Contents"
    theme: flatly
    embed-resources: true
    fig-align: center
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")
pacman::p_load(tidyverse, lubridate, here, scales, ggrepel, flextable, ggh4x)

set_flextable_defaults(
  font.family = "Arial",
  font.size   = 10,
  border.color = "#CCCCCC"
)
```

```{r}
#| label: load-data
df_raw <- read_csv(
  here::here("outputs", "master_combined_counts.csv"),
  col_types       = cols(.default = "c"),
  show_col_types  = FALSE
)

df <- df_raw |>
  mutate(
    date        = dmy(count_end_date),
    suspect     = as.numeric(cases_suspect),
    probable    = as.numeric(cases_probable),
    confirmed   = as.numeric(cases_confirmed),
    deaths      = coalesce(as.numeric(deaths_suspected), as.numeric(deaths_probable), as.numeric(deaths_confirmed)),
    deaths_susp = as.numeric(deaths_suspected),
    is_agg      = (toupper(is_aggregate) == "TRUE"),
    # Only sum to zero when at least one component is present; otherwise NA
    # (avoids a false zero for zones that simply weren't reported in that sitrep)
    unconfirmed = if_else(
      !is.na(suspect) | !is.na(probable),
      coalesce(suspect, 0L) + coalesce(probable, 0L),
      NA_real_
    )
  )

ZONE_LEVELS <- c(
  "Mongbwalu", "Rwampara", "Bunia", "Nyankunde", "Bambu",
  "Butembo", "Katwa", "Goma"
)

# Colorblind-safe palette — Wong (2011)
PAL <- c(
  "Suspect/probable cases" = "#56B4E9",
  "Confirmed cases"        = "#D55E00",
  "Outbreak deaths"        = "#404040"
)

int_breaks <- function(x) unique(floor(pretty(x)))

theme_sitrep <- function(base_size = 11) {
  theme_minimal(base_size = base_size) %+replace%
    theme(
      plot.title       = element_text(face = "bold", size = base_size + 2, hjust = 0),
      plot.subtitle    = element_text(size = base_size - 1, colour = "grey45", hjust = 0,
                                      margin = margin(b = 8)),
      plot.caption     = element_text(size = base_size - 3, colour = "grey55", hjust = 1,
                                      margin = margin(t = 8)),
      plot.margin      = margin(12, 16, 8, 12),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(colour = "grey92"),
      legend.position  = "bottom",
      legend.key.size  = unit(0.75, "lines"),
      legend.text      = element_text(size = base_size - 1),
      strip.text       = element_text(face = "bold", size = base_size - 1,
                                      margin = margin(b = 4)),
      axis.title       = element_text(size = base_size - 1, colour = "grey35"),
      axis.text        = element_text(size = base_size - 2, colour = "grey25")
    )
}

# ── Derived subsets ────────────────────────────────────────────────────────────
totals <- df |>
  filter(count_type == "Cumules", is_agg, zone == "Total") |>
  group_by(date) |>
  slice_max(order_by = unconfirmed, n = 1, with_ties = FALSE) |>
  ungroup() |>
  arrange(date) |>
  mutate(deaths = cummax(replace_na(deaths, 0L)))

zone_ts <- df |>
  filter(count_type == "Cumules", !is_agg, zone %in% ZONE_LEVELS) |>
  group_by(date, zone, province) |>
  slice_max(order_by = unconfirmed, n = 1, with_ties = FALSE) |>
  ungroup() |>
  mutate(zone = factor(zone, levels = ZONE_LEVELS)) |>
  arrange(zone, date) |>
  group_by(zone) |>
  mutate(
    # Flag raw downward revisions before enforcing monotonicity
    confirmed_raw   = confirmed,
    unconfirmed_raw = unconfirmed,
    deaths_raw      = deaths,
    deaths_susp_raw = deaths_susp,
    rev_confirmed   = confirmed   < lag(cummax(replace_na(confirmed,   0L)), default = 0L),
    rev_unconfirmed = unconfirmed < lag(cummax(replace_na(unconfirmed, 0L)), default = 0L),
    # Apply cummax only over observed values; leave NAs before the first
    # real observation so geom_line breaks rather than drawing from zero
    unconfirmed = {x <- unconfirmed; i <- !is.na(x); if (any(i)) x[i] <- cummax(x[i]); x},
    confirmed   = {x <- confirmed;   i <- !is.na(x); if (any(i)) x[i] <- cummax(x[i]); x},
    deaths      = {x <- deaths;      i <- !is.na(x); if (any(i)) x[i] <- cummax(x[i]); x},
    deaths_susp = {x <- deaths_susp; i <- !is.na(x); if (any(i)) x[i] <- cummax(x[i]); x}
  ) |>
  ungroup()

# Warn about any downward revisions that cummax suppressed
revisions <- zone_ts |>
  filter(rev_confirmed | rev_unconfirmed) |>
  transmute(
    zone, date,
    confirmed_raw, confirmed,
    unconfirmed_raw, unconfirmed
  )
if (nrow(revisions) > 0) {
  message("NOTE: cummax suppressed ", nrow(revisions),
          " downward revision(s) in cumulative counts:\n",
          paste(capture.output(print(revisions)), collapse = "\n"))
}

zone_new <- df |>
  filter(count_type == "Nouveaux", !is_agg, zone %in% ZONE_LEVELS) |>
  group_by(date, zone, province) |>
  slice_max(order_by = unconfirmed, n = 1, with_ties = FALSE) |>
  ungroup() |>
  mutate(zone = factor(zone, levels = ZONE_LEVELS)) |>
  arrange(zone, date)

daily_new <- df |>
  filter(count_type == "Nouveaux", is_agg, zone == "Total") |>
  group_by(date) |>
  slice_max(order_by = unconfirmed, n = 1, with_ties = FALSE) |>
  ungroup() |>
  arrange(date)

make_zone_labels <- function(dat) {
  dat |>
    mutate(
      zone_label = paste0(zone, " (", province, ")"),
      zone_label = factor(zone_label,
        levels = dat |>
          distinct(zone, province) |>
          arrange(zone) |>
          mutate(lbl = paste0(zone, " (", province, ")")) |>
          pull(lbl)
      )
    )
}

# ── Inline summary statistics ──────────────────────────────────────────────────
latest_total <- totals |> slice_max(date, n = 1)
latest_date  <- format(max(totals$date, na.rm = TRUE), "%d %B %Y")
n_unconf     <- latest_total$unconfirmed
n_conf       <- latest_total$confirmed
n_deaths     <- coalesce(latest_total$deaths, 0L)
n_total      <- n_unconf + n_conf
cfr_pct      <- if (n_total > 0) round(n_deaths / n_total * 100, 1) else NA_real_
n_zones      <- n_distinct(zone_ts$zone)
n_provinces  <- n_distinct(zone_ts$province)
```

```{r}
#| label: load-kraemer
# Kraemer lab canonical name → this extraction zone name
KRAEMER_NAME_MAP <- c("Mongbalu" = "Mongbwalu", "Nyakunde" = "Nyankunde")

read_kraemer_long <- function(filename) {
  read_csv(
    here::here("Ebola_DRC_2026", "build", "long", filename),
    col_types      = cols(.default = "c"),
    show_col_types = FALSE
  )
}

kraemer <- read_kraemer_long("insp_sitrep__cumulative_confirmed_cases.csv") |>
  rename(confirmed = cumulative_confirmed_cases) |>
  full_join(
    read_kraemer_long("insp_sitrep__cumulative_suspected_cases.csv") |>
      rename(suspected = cumulative_suspected_cases),
    by = c("nom", "date")
  ) |>
  full_join(
    read_kraemer_long("insp_sitrep__cumulative_suspected_deaths.csv") |>
      rename(deaths_susp = cumulative_suspected_deaths),
    by = c("nom", "date")
  ) |>
  full_join(
    read_kraemer_long("insp_sitrep__cumulative_confirmed_deaths.csv") |>
      rename(deaths_conf = cumulative_confirmed_deaths),
    by = c("nom", "date")
  ) |>
  mutate(
    zone       = recode(nom, !!!KRAEMER_NAME_MAP),
    date       = as.Date(date),
    confirmed  = suppressWarnings(as.numeric(confirmed)),
    suspected  = suppressWarnings(as.numeric(suspected)),
    deaths_susp = suppressWarnings(as.numeric(deaths_susp)),
    deaths_conf = suppressWarnings(as.numeric(deaths_conf)),
    # Total deaths = suspected + confirmed (mirrors our coalesce logic)
    deaths     = coalesce(deaths_susp, 0) + coalesce(deaths_conf, 0),
    deaths     = if_else(is.na(deaths_susp) & is.na(deaths_conf), NA_real_, deaths),
    source     = "Kraemer lab"
  ) |>
  select(zone, date, confirmed, suspected, deaths_susp, deaths_conf, deaths, source) |>
  filter(zone %in% ZONE_LEVELS)

kraemer_hosp <- read_kraemer_long("insp_sitrep__hospitalised.csv") |>
  mutate(
    zone         = recode(nom, !!!KRAEMER_NAME_MAP),
    date         = as.Date(date),
    hospitalised = suppressWarnings(as.numeric(hospitalised))
  )

kraemer_contacts <- read_kraemer_long("insp_sitrep__cumulative_contacts_traced.csv") |>
  mutate(
    zone            = recode(nom, !!!KRAEMER_NAME_MAP),
    date            = as.Date(date),
    contacts_traced = suppressWarnings(as.numeric(cumulative_contacts_traced))
  ) |>
  filter(!is.na(contacts_traced))

# Response indicators (hospitalisations by zone)
kraemer_response <- read_kraemer_long("insp_sitrep__hospitalised.csv") |>
  full_join(read_kraemer_long("insp_sitrep__in_bed_previous_day.csv"), by = c("nom", "date")) |>
  full_join(
    read_kraemer_long("insp_sitrep__new_hosp_admissions.csv") |>
      rename(total_admissions = new_all_admissions),
    by = c("nom", "date")
  ) |>
  full_join(
    read_kraemer_long("insp_sitrep__new_hosp_detainees.csv") |>
      rename(new_admissions_ppl = new_hosp_detainees),
    by = c("nom", "date")
  ) |>
  full_join(
    read_kraemer_long("insp_sitrep__new_hosp_other.csv") |>
      rename(new_admissions_other = new_other),
    by = c("nom", "date")
  ) |>
  mutate(
    zone   = recode(nom, !!!KRAEMER_NAME_MAP),
    date   = as.Date(date),
    across(c(hospitalised, in_bed_previous_day, total_admissions,
             new_admissions_ppl, new_admissions_other),
           ~suppressWarnings(as.numeric(.x))),
    source = "Kraemer lab"
  ) |>
  filter(zone %in% ZONE_LEVELS)

# POE (checkpoint-level in Kraemer; summed to daily aggregate for comparison)
kraemer_poe <- read_kraemer_long("insp_sitrep__total_poe_passed.csv") |>
  rename(total_passed = total_poe_passed) |>
  full_join(
    read_kraemer_long("insp_sitrep__total_poe_screened.csv") |>
      rename(total_screened = total_poe_screened),
    by = c("nom", "date")
  ) |>
  full_join(
    read_kraemer_long("insp_sitrep__total_poe_hand_washing.csv") |>
      rename(total_handwashing = total_poe_hand_washing),
    by = c("nom", "date")
  ) |>
  full_join(
    read_kraemer_long("insp_sitrep__total_poe_sanitised.csv") |>
      rename(total_sensitised = total_poe_sanitised),
    by = c("nom", "date")
  ) |>
  mutate(
    date = as.Date(date),
    across(c(total_passed, total_screened, total_handwashing, total_sensitised),
           ~suppressWarnings(as.numeric(.x)))
  ) |>
  group_by(date) |>
  summarise(
    across(c(total_passed, total_screened, total_handwashing, total_sensitised),
           ~sum(.x, na.rm = TRUE)),
    .groups = "drop"
  ) |>
  mutate(source = "Kraemer lab")
```

## Introduction

This report provides an automated summary of the 2026 DRC Maladie à Virus Ebola (MVE) outbreak in Ituri and Nord-Kivu, using data extracted from official INSP situation reports.

## Methods

### Data sources

Data were extracted from INSP DRC PDF situation reports for the 2026 outbreak, downloaded from <https://insp.cd/ebola/> with manifest-based incremental tracking. MD5 checksums showed these PDFs are byte-identical to those used by the Kraemer lab. Cross-source validation used Kraemer's manually coded long-format time series from the same source documents.

### Automated PDF extraction

PDFs were sent to Claude (`claude-sonnet-4-6`, temperature 0) as base64 documents for visual table extraction. The pipeline extracts four outputs: new cases, cumulative cases/deaths, patient movement (from SitRep 006 onwards), and PoE/PoC activity, plus a verbatim text transcript for audit. Output is validated and normalised to a common schema by province and health zone.

### Data harmonisation

Zone names were standardised, and suspect plus probable counts were combined as an unconfirmed category to handle naming differences between reports. Cumulative series were constrained to be non-decreasing within zone, and when multiple rows existed for the same zone-date the row with the highest unconfirmed count was retained.

### Statistical analysis

Analyses were run in R (version `r R.version$major`.`r R.version$minor`) using `tidyverse` [@tidyverse], with `ggplot2` figures [@ggplot2], `flextable` tables [@flextable], and Quarto rendering.

## Results

As of `r latest_date`, the DRC MVE outbreak had recorded a cumulative total of `r prettyNum(n_total, big.mark = ",")` cases (`r prettyNum(n_conf, big.mark = ",")` confirmed, `r prettyNum(n_unconf, big.mark = ",")` suspect or probable) and `r prettyNum(n_deaths, big.mark = ",")` deaths, corresponding to a case fatality ratio of `r cfr_pct`%. Cases have been reported across `r n_zones` health zones in `r n_provinces` provinces.

### Cumulative epidemic curve

```{r}
#| label: fig-cumulative
#| fig-cap: "Cumulative MVE cases and deaths, DRC. Suspect and probable cases are combined. End-of-line labels show the latest recorded value. Source: INSP DRC."
#| fig-width: 8
#| fig-height: 5
p1_dat <- totals |>
  select(date,
         `Suspect/probable cases` = unconfirmed,
         `Confirmed cases`        = confirmed,
         `Outbreak deaths`        = deaths) |>
  pivot_longer(-date, names_to = "category", values_to = "count") |>
  mutate(category = factor(category, levels = names(PAL)))

p1_last <- p1_dat |>
  filter(!is.na(count)) |>
  group_by(category) |>
  slice_max(date, n = 1) |>
  ungroup()

p <- ggplot(p1_dat, aes(x = date, y = count, colour = category, group = category)) +
  geom_line(linewidth = 1.1, na.rm = TRUE) +
  geom_point(size = 2.8, na.rm = TRUE) +
  geom_label_repel(
    data           = p1_last,
    aes(label      = count),
    size           = 3.5,
    nudge_x        = 1.5,
    direction      = "y",
    hjust          = 0,
    segment.colour = "grey70",
    segment.size   = 0.3,
    show.legend    = FALSE,
    na.rm          = TRUE
  ) +
  scale_colour_manual(values = PAL) +
  scale_x_date(
    date_labels = "%d %b",
    date_breaks = "2 days",
    expand      = expansion(add = c(0, 4))
  ) +
  scale_y_continuous(breaks = int_breaks, expand = expansion(mult = c(0, 0.08))) +
  labs(
    title    = "MVE DRC \u2014 Cumulative epidemic curve",
    subtitle = "Total counts across all health zones",
    x = NULL, y = "Cumulative count", colour = NULL,
    caption  = "Source: INSP DRC situation reports"
  ) +
  theme_sitrep(base_size = 12)
ggsave(here("outputs", "plots", "fig-cumulative.png"), plot = p, width = 8, height = 5, dpi = 300)
p
```

### Daily new cases

```{r}
#| label: fig-new-cases
#| fig-cap: "New cases reported per situation report across all health zones, DRC. Source: INSP DRC."
#| fig-width: 7
#| fig-height: 4
p <- daily_new |>
  select(date,
         `Suspect/probable cases` = unconfirmed,
         `Confirmed cases`        = confirmed) |>
  pivot_longer(-date, names_to = "category", values_to = "count") |>
  mutate(category = factor(category, levels = c("Suspect/probable cases", "Confirmed cases"))) |>
  ggplot(aes(x = date, y = count, fill = category)) +
  geom_col(position = "dodge", width = 0.3, na.rm = TRUE) +
  scale_fill_manual(values = PAL) +
  scale_x_date(date_labels = "%d %b", date_breaks = "1 day") +
  scale_y_continuous(breaks = int_breaks, expand = expansion(mult = c(0, 0.1))) +
  labs(
    title    = "MVE DRC \u2014 Daily new cases",
    subtitle = "Total reported per situation report",
    x = NULL, y = "New cases", fill = NULL,
    caption  = "Source: INSP DRC situation reports"
  ) +
  theme_sitrep(base_size = 12)
ggsave(here("outputs", "plots", "fig-new-cases.png"), plot = p, width = 7, height = 4, dpi = 300)
p
```

### Geographic distribution — cumulative counts

```{r}
#| label: fig-zone-cumulative
#| fig-cap: "Cumulative cases by health zone. Y-axis is free across panels. Source: INSP DRC."
#| fig-width: 10
#| fig-height: 11
p <- zone_ts |>
  make_zone_labels() |>
  select(date, zone_label,
         `Suspect/probable cases` = unconfirmed,
         `Confirmed cases`        = confirmed,
         `Outbreak deaths`        = deaths) |>
  pivot_longer(
    cols      = c(`Suspect/probable cases`, `Confirmed cases`, `Outbreak deaths`),
    names_to  = "category",
    values_to = "count"
  ) |>
  mutate(category = factor(category, levels = names(PAL))) |>
  ggplot(aes(x = date, y = count, colour = category, group = category)) +
  geom_line(linewidth = 0.9, na.rm = TRUE) +
  geom_point(size = 2.2, na.rm = TRUE) +
  scale_colour_manual(values = PAL) +
  scale_x_date(date_labels = "%d %b", date_breaks = "1 day") +
  scale_y_continuous(breaks = int_breaks, expand = expansion(mult = c(0, 0.15))) +
  facet_wrap(~ zone_label, scales = "free_y", ncol = 2) +
  labs(
    title    = "MVE DRC \u2014 Cumulative cases by health zone",
    subtitle = "Each panel shows one health zone; y-axis is free.",
    x = NULL, y = "Cumulative count", colour = NULL,
    caption  = "Source: INSP DRC situation reports"
  ) +
  theme_sitrep(base_size = 10) +
  theme(axis.text.x = element_text(size = 8, angle = 45, hjust = 1))
ggsave(here("outputs", "plots", "fig-zone-cumulative.png"), plot = p, width = 10, height = 11, dpi = 300)
p
```

### Geographic distribution — new cases

```{r}
#| label: fig-zone-new
#| fig-cap: "New cases per situation report by health zone. Y-axis is free across panels. Source: INSP DRC."
#| fig-width: 10
#| fig-height: 11
p <- zone_new |>
  make_zone_labels() |>
  select(date, zone_label,
         `Suspect/probable cases` = unconfirmed,
         `Confirmed cases`        = confirmed) |>
  pivot_longer(
    cols      = c(`Suspect/probable cases`, `Confirmed cases`),
    names_to  = "category",
    values_to = "count"
  ) |>
  mutate(category = factor(category, levels = c("Suspect/probable cases", "Confirmed cases"))) |>
  ggplot(aes(x = date, y = count, fill = category)) +
  geom_col(position = "dodge", width = 0.4, na.rm = TRUE) +
  scale_fill_manual(values = PAL) +
  scale_x_date(date_labels = "%d %b", date_breaks = "1 day") +
  scale_y_continuous(breaks = int_breaks, expand = expansion(mult = c(0, 0.15))) +
  facet_wrap(~ zone_label, scales = "free_y", ncol = 2) +
  labs(
    title    = "MVE DRC \u2014 Daily new cases by health zone",
    subtitle = "Per situation report; y-axis is free.",
    x = NULL, y = "New cases", fill = NULL,
    caption  = "Source: INSP DRC situation reports"
  ) +
  theme_sitrep(base_size = 10) +
  theme(axis.text.x = element_text(size = 8, angle = 45, hjust = 1))
ggsave(here("outputs", "plots", "fig-zone-new.png"), plot = p, width = 10, height = 11, dpi = 300)
p
```

### Current counts by health zone

Table 1 presents the most recent cumulative case counts disaggregated by health zone, drawn from the latest available situation report for each zone.

```{r}
#| label: tbl-zones
#| tbl-cap: "Table 1. Latest cumulative case counts by health zone."
zone_latest <- zone_ts |>
  group_by(zone, province) |>
  slice_max(date, n = 1) |>
  ungroup() |>
  arrange(zone) |>
  transmute(
    `Health zone`      = as.character(zone),
    Province           = province,
    `Latest sitrep`    = format(date, "%d %b %Y"),
    `Suspect/probable` = unconfirmed,
    `Confirmed`        = confirmed,
    `Deaths`           = deaths
  )

total_row <- tibble(
  `Health zone`      = "Total",
  Province           = "\u2014",
  `Latest sitrep`    = format(max(totals$date, na.rm = TRUE), "%d %b %Y"),
  `Suspect/probable` = sum(zone_latest$`Suspect/probable`, na.rm = TRUE),
  `Confirmed`        = sum(zone_latest$`Confirmed`,        na.rm = TRUE),
  `Deaths`           = sum(zone_latest$`Deaths`,           na.rm = TRUE)
)

bind_rows(zone_latest, total_row) |>
  flextable() |>
  bold(part = "header") |>
  bold(i = nrow(zone_latest) + 1L) |>
  bg(part = "header", bg = "#1A3A5C") |>
  color(part = "header", color = "white") |>
  bg(i = nrow(zone_latest) + 1L, bg = "#E8EFF7") |>
  hline(i = nrow(zone_latest), part = "body") |>
  align(j = 4:6, align = "right", part = "all") |>
  colformat_int(j = 4:6) |>
  autofit()
```

### Cross-source validation

Zone-level cumulative counts from this pipeline were compared with the [Kraemer lab Ebola DRC 2026](https://github.com/kraemer-lab/Ebola_DRC_2026) manual extraction. As both use the same MD5-verified PDFs, remaining differences reflect extraction and interpretation choices rather than source-document differences. Name variants (Mongbwalu/Mongbalu, Nyankunde/Nyakunde) were harmonised, and `ND` values were treated as missing.

### Data quality notes — case counts

Two Mongbwalu issues are important for interpretation. First, the 14 May value of 246 suspected cases appears in SitRep 001 only as an Ituri-wide aggregate, not as a zone-level row. The first report with zone-level Mongbwalu data in this pipeline is SitRep 004 (18 May, 302 suspected), so the apparent early zone trajectory reflects incomplete zone-level coverage.

Second, both this pipeline (pre-cummax) and the Kraemer series show a drop in Mongbwalu cumulative confirmed cases from 14 (21 May) to 10 (22 May). Because cumulative counts should not decrease, this likely reflects a sitrep reporting/reclassification issue rather than extraction artefact. The cummax-constrained automated series suppresses this dip; the Kraemer series shows it directly.

```{r}
#| label: fig-comparison
#| fig-cap: "Cross-source comparison of cumulative confirmed cases, suspect/probable cases, and outbreak deaths by health zone. Solid lines: automated extraction (cummax applied); dotted lines: automated extraction (raw, no cummax); dashed lines: Kraemer lab (manual extraction). Both automated and Kraemer sources use identical PDFs; divergence reflects extraction method differences."
#| fig-width: 12
#| fig-height: 14
PAL_COMP <- c(
  "Automated extraction" = "#0072B2",
  "Automated (no cummax)" = "#56B4E9",
  "Kraemer lab"          = "#E69F00"
)
LTY_COMP <- c(
  "Automated extraction" = "solid",
  "Automated (no cummax)" = "dotted",
  "Kraemer lab"          = "dashed"
)

zone_province_lut <- zone_ts |>
  distinct(zone, province) |>
  mutate(zone = as.character(zone))

our_for_comp <- zone_ts |>
  transmute(zone = as.character(zone), province, date,
            confirmed, suspected = unconfirmed,
            deaths_susp, deaths,
            source = "Automated extraction")

our_raw_for_comp <- zone_ts |>
  transmute(zone = as.character(zone), province, date,
            confirmed   = confirmed_raw,
            suspected   = unconfirmed_raw,
            deaths_susp = deaths_susp_raw,
            deaths      = deaths_raw,
            source      = "Automated (no cummax)")

# Kraemer deaths_susp = suspected only; deaths = suspected + confirmed (total)
kraemer_for_comp <- kraemer |>
  left_join(zone_province_lut, by = "zone") |>
  transmute(zone, province, date, confirmed, suspected,
            deaths_susp, deaths,
            source)

comp_long <- bind_rows(our_for_comp, our_raw_for_comp, kraemer_for_comp) |>
  pivot_longer(c(confirmed, suspected, deaths_susp, deaths),
               names_to = "metric", values_to = "count") |>
  mutate(
    metric = recode(metric,
      "confirmed"   = "Confirmed cases",
      "suspected"   = "Suspect/probable cases",
      "deaths_susp" = "Suspected deaths",
      "deaths"      = "Total deaths (coalesced)"
    ),
    metric   = factor(metric, levels = c("Confirmed cases", "Suspect/probable cases",
                                          "Suspected deaths", "Total deaths (coalesced)")),
    source   = factor(source, levels = c("Automated extraction", "Automated (no cummax)", "Kraemer lab")),
    province = factor(province, levels = c("Ituri", "Nord-Kivu")),
    zone     = factor(zone, levels = ZONE_LEVELS)
  ) |>
  filter(!is.na(count))

p <- ggplot(comp_long, aes(x = date, y = count, colour = source, linetype = source)) +
  geom_line(linewidth = 0.9, na.rm = TRUE) +
  geom_point(size = 2.0, na.rm = TRUE) +
  facet_nested(metric ~ province + zone, scales = "free_y") +
  scale_colour_manual(values = PAL_COMP) +
  scale_linetype_manual(values = LTY_COMP) +
  scale_x_date(date_labels = "%d %b", date_breaks = "2 days") +
  scale_y_continuous(breaks = int_breaks, limits = c(0, NA), expand = expansion(mult = c(0, 0.15))) +
  labs(
    title    = "MVE DRC \u2014 Cross-source validation",
    subtitle = "Cumulative counts: automated extraction (with and without cummax) vs Kraemer lab (manual extraction, same source PDFs)",
    x = NULL, y = "Cumulative count", colour = NULL, linetype = NULL,
    caption  = "Sources: INSP DRC situation reports; github.com/kraemer-lab/Ebola_DRC_2026"
  ) +
  theme_sitrep(base_size = 9) +
  theme(
    axis.text.x     = element_text(angle = 45, hjust = 1, size = 7),
    panel.border    = element_rect(colour = "grey70", fill = NA, linewidth = 0.4),
    panel.spacing   = unit(0.6, "lines")
  )
ggsave(here("outputs", "plots", "fig-comparison.png"), plot = p, width = 12, height = 14, dpi = 300)
p
```

Table 2 presents numerical differences at the most recent date available in each source for each zone.

```{r}
#| label: tbl-comparison
#| tbl-cap: "Table 2. Cumulative case counts at the latest available date per zone: automated extraction vs Kraemer lab. Delta = ours minus Kraemer. Dashes indicate no data in that source."
ours_latest <- zone_ts |>
  group_by(zone) |>
  slice_max(date, n = 1) |>
  ungroup() |>
  transmute(zone, conf_ours = confirmed, susp_ours = unconfirmed, dead_ours = deaths)

kraemer_latest <- kraemer |>
  group_by(zone) |>
  slice_max(date, n = 1) |>
  ungroup() |>
  transmute(zone, conf_them = confirmed, susp_them = suspected, dead_them = deaths)

full_join(ours_latest, kraemer_latest, by = "zone") |>
  mutate(
    d_conf = conf_ours - conf_them,
    d_susp = susp_ours - susp_them,
    d_dead = dead_ours - dead_them
  ) |>
  arrange(zone) |>
  transmute(
    zone,
    conf_ours, conf_them, d_conf,
    susp_ours, susp_them, d_susp,
    dead_ours, dead_them, d_dead
  ) |>
  setNames(c(
    "Health zone",
    "Confirmed (ours)", "Confirmed (Kraemer)", "\u0394 confirmed",
    "Suspect/prob. (ours)", "Suspect/prob. (Kraemer)", "\u0394 suspect/prob.",
    "Deaths (ours)", "Deaths (Kraemer)", "\u0394 deaths"
  )) |>
  flextable() |>
  bold(part = "header") |>
  bg(part = "header", bg = "#1A3A5C") |>
  color(part = "header", color = "white") |>
  align(j = 2:10, align = "right", part = "all") |>
  colformat_int(j = 2:10, na_str = "\u2014") |>
  autofit()
```

### Response indicators

```{r}
#| label: load-response-ours
resp_ours <- read_csv(
  here::here("outputs", "master_response_counts.csv"),
  col_types      = cols(.default = "c"),
  show_col_types = FALSE
) |>
  mutate(
    date   = dmy(date),
    zone   = factor(zone, levels = ZONE_LEVELS),
    across(c(in_bed_previous_day, new_admissions_ppl, new_admissions_other,
             total_admissions, total_isolated, cumulative_admitted),
           ~suppressWarnings(as.numeric(.x))),
    source = "Automated extraction"
  ) |>
  filter(!is.na(zone), !zone %in% c("Total"))

poe_ours <- read_csv(
  here::here("outputs", "master_poe_counts.csv"),
  col_types      = cols(.default = "c"),
  show_col_types = FALSE
) |>
  mutate(
    date = dmy(date),
    across(c(total_passed, total_screened, total_handwashing, total_sensitised),
           ~suppressWarnings(as.numeric(.x))),
    source = "Automated extraction"
  ) |>
  select(date, total_passed, total_screened, total_handwashing, total_sensitised, source)
```

### Data quality notes — response indicators

Two response-indicator discrepancies were identified. For Rwampara on 21 May, SitRep 007 presents three unexplained sub-columns; summing rules are ambiguous, so both series should be treated cautiously at this point. For 20-21 May, earlier table layouts likely led to a Kraemer column-mapping error in new admissions; concordance improves from 22 May onward when the layout is standardised.

```{r}
#| label: fig-comparison-response
#| fig-cap: "Cross-source comparison of zone-level response indicators. Solid lines: automated extraction; dashed lines: Kraemer lab. The 21 May Rwampara values are unreliable in both series: the sitrep presents three unexplained sub-columns whose relationship is ambiguous. For 20–21 May, Kraemer's new-admissions values reflect a column-mapping error in the earlier table format."
#| fig-width: 12
#| fig-height: 9
PAL_RESP <- c("Automated extraction" = "#0072B2", "Kraemer lab" = "#E69F00")
LTY_RESP <- c("Automated extraction" = "solid",   "Kraemer lab" = "dashed")

resp_comp <- bind_rows(
  resp_ours |>
    transmute(date, zone,
              hospitalised        = total_isolated,
              in_bed_previous_day,
              total_admissions,
              source),
  kraemer_response |>
    transmute(date, zone = factor(zone, levels = ZONE_LEVELS),
              hospitalised,
              in_bed_previous_day,
              total_admissions,
              source)
) |>
  pivot_longer(c(hospitalised, in_bed_previous_day, total_admissions),
               names_to = "metric", values_to = "count") |>
  mutate(
    metric = recode(metric,
      "hospitalised"        = "Currently hospitalised",
      "in_bed_previous_day" = "In bed (prev. day)",
      "total_admissions"    = "New admissions (total)"
    ),
    metric = factor(metric, levels = c("Currently hospitalised",
                                        "In bed (prev. day)",
                                        "New admissions (total)")),
    source = factor(source, levels = c("Automated extraction", "Kraemer lab"))
  ) |>
  filter(!is.na(count), !is.na(zone))

p <- ggplot(resp_comp, aes(x = date, y = count, colour = source, linetype = source)) +
  geom_line(linewidth = 0.9, na.rm = TRUE) +
  geom_point(size = 2.0, na.rm = TRUE) +
  facet_nested(metric ~ zone, scales = "free_y") +
  scale_colour_manual(values = PAL_RESP) +
  scale_linetype_manual(values = LTY_RESP) +
  scale_x_date(date_labels = "%d %b", date_breaks = "2 days") +
  scale_y_continuous(breaks = int_breaks, limits = c(0, NA),
                     expand = expansion(mult = c(0, 0.15))) +
  labs(
    title    = "MVE DRC \u2014 Response indicators — cross-source validation",
    subtitle = "Automated extraction vs Kraemer lab (manual, same source PDFs)",
    x = NULL, y = "Count", colour = NULL, linetype = NULL,
    caption  = "Sources: INSP DRC situation reports; github.com/kraemer-lab/Ebola_DRC_2026"
  ) +
  theme_sitrep(base_size = 9) +
  theme(
    axis.text.x  = element_text(angle = 45, hjust = 1, size = 7),
    panel.border = element_rect(colour = "grey70", fill = NA, linewidth = 0.4),
    panel.spacing = unit(0.5, "lines")
  )
ggsave(here("outputs", "plots", "fig-comparison-response.png"), plot = p,
       width = 12, height = 9, dpi = 300)
p
```

```{r}
#| label: fig-comparison-poe
#| fig-cap: "Cross-source comparison of daily Point of Entry (POE) screening totals. Automated extraction reports a single aggregate per sitrep date; Kraemer lab data are summed across individual checkpoints to produce a daily total."
#| fig-width: 9
#| fig-height: 6
poe_comp <- bind_rows(poe_ours, kraemer_poe) |>
  pivot_longer(c(total_passed, total_screened, total_handwashing, total_sensitised),
               names_to = "metric", values_to = "count") |>
  mutate(
    metric = recode(metric,
      "total_passed"      = "Total passed",
      "total_screened"    = "Screened",
      "total_handwashing" = "Handwashing stations",
      "total_sensitised"  = "Sensitised"
    ),
    metric = factor(metric, levels = c("Total passed", "Screened",
                                        "Handwashing stations", "Sensitised")),
    source = factor(source, levels = c("Automated extraction", "Kraemer lab"))
  ) |>
  filter(!is.na(count))

p <- ggplot(poe_comp, aes(x = date, y = count, colour = source, linetype = source)) +
  geom_line(linewidth = 0.9, na.rm = TRUE) +
  geom_point(size = 2.5, na.rm = TRUE) +
  facet_wrap(~ metric, scales = "free_y", ncol = 2) +
  scale_colour_manual(values = PAL_RESP) +
  scale_linetype_manual(values = LTY_RESP) +
  scale_x_date(date_labels = "%d %b", date_breaks = "2 days") +
  scale_y_continuous(breaks = int_breaks, limits = c(0, NA),
                     expand = expansion(mult = c(0, 0.12))) +
  labs(
    title    = "MVE DRC \u2014 POE screening — cross-source validation",
    subtitle = "Daily aggregate totals: automated extraction vs Kraemer lab (sum across checkpoints)",
    x = NULL, y = "Count", colour = NULL, linetype = NULL,
    caption  = "Sources: INSP DRC situation reports; github.com/kraemer-lab/Ebola_DRC_2026"
  ) +
  theme_sitrep(base_size = 11) +
  theme(
    axis.text.x  = element_text(angle = 45, hjust = 1, size = 8),
    panel.border = element_rect(colour = "grey70", fill = NA, linewidth = 0.4),
    panel.spacing = unit(0.6, "lines")
  )
ggsave(here("outputs", "plots", "fig-comparison-poe.png"), plot = p,
       width = 9, height = 6, dpi = 300)
p
```

```{r}
#| label: fig-comparison-contacts
#| fig-cap: "Cross-source comparison of cumulative contacts traced by health zone. Solid lines: automated extraction (cummax applied); dashed lines: Kraemer lab (manual extraction). ND values excluded."
#| fig-width: 11
#| fig-height: 7
contacts_ours <- df |>
  filter(
    count_type == "Cumules",
    zone %in% ZONE_LEVELS,
    !is.na(contacts), contacts != ""
  ) |>
  mutate(contacts = suppressWarnings(as.numeric(contacts)),
         zone = factor(zone, levels = ZONE_LEVELS)) |>
  filter(!is.na(contacts)) |>
  arrange(zone, date) |>
  group_by(zone) |>
  mutate(contacts = cummax(contacts)) |>
  ungroup() |>
  transmute(date, zone, contacts, source = "Automated extraction")

contacts_kraemer <- kraemer_contacts |>
  filter(zone %in% ZONE_LEVELS) |>
  transmute(date, zone = factor(zone, levels = ZONE_LEVELS),
            contacts = contacts_traced, source = "Kraemer lab")

contacts_comp <- bind_rows(contacts_ours, contacts_kraemer) |>
  filter(!is.na(contacts)) |>
  left_join(zone_province_lut, by = "zone") |>
  mutate(
    source   = factor(source, levels = c("Automated extraction", "Kraemer lab")),
    province = factor(province, levels = c("Ituri", "Nord-Kivu")),
    zone     = factor(zone, levels = ZONE_LEVELS)
  )

p <- ggplot(contacts_comp,
            aes(x = date, y = contacts, colour = source, linetype = source)) +
  geom_line(linewidth = 0.9, na.rm = TRUE) +
  geom_point(size = 2.0, na.rm = TRUE) +
  facet_nested(~ province + zone, scales = "free_y") +
  scale_colour_manual(values = PAL_COMP[c("Automated extraction","Kraemer lab")]) +
  scale_linetype_manual(values = LTY_COMP[c("Automated extraction","Kraemer lab")]) +
  scale_x_date(date_labels = "%d %b", date_breaks = "2 days") +
  scale_y_continuous(breaks = int_breaks, limits = c(0, NA),
                     expand = expansion(mult = c(0, 0.15))) +
  labs(
    title    = "MVE DRC \u2014 Cumulative contacts traced — cross-source validation",
    subtitle = "Automated extraction (cummax) vs Kraemer lab (manual extraction, same source PDFs)",
    x = NULL, y = "Cumulative contacts traced", colour = NULL, linetype = NULL,
    caption  = "Sources: INSP DRC situation reports; github.com/kraemer-lab/Ebola_DRC_2026"
  ) +
  theme_sitrep(base_size = 9) +
  theme(
    axis.text.x  = element_text(angle = 45, hjust = 1, size = 7),
    panel.border = element_rect(colour = "grey70", fill = NA, linewidth = 0.4),
    panel.spacing = unit(0.5, "lines")
  )
ggsave(here("outputs", "plots", "fig-comparison-contacts.png"), plot = p,
       width = 11, height = 7, dpi = 300)
p
```

## Discussion

The outbreak trajectory shows rising cumulative burden over the reporting period, with confirmed cases remaining a minority of total reported cases. Transmission spans multiple zones across Ituri and Nord-Kivu, supporting the need for coordinated multi-province response.

Interpretation should account for data-structure limitations and reporting artefacts. Early reports include limited zone-level detail, response indicators are not consistently structured across sitreps, and cumulative-series enforcement (cummax) can conceal true downward revisions. Because this pipeline and the Kraemer series use the same MD5-verified PDFs, observed divergences primarily reflect extraction and interpretation differences, especially around ambiguous table layouts and early aggregate-only reporting.
