---
title: "Maladie à Virus Ebola — Situation Report"
subtitle: "Democratic Republic of Congo"
date: today
date-format: "DD MMMM YYYY"
format:
  html:
    toc: true
    toc-depth: 2
    toc-title: "Contents"
    theme: flatly
    embed-resources: true
    fig-align: center
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")
pacman::p_load(tidyverse, lubridate, here, scales, ggrepel, flextable, ggh4x)

set_flextable_defaults(
  font.family = "Arial",
  font.size   = 10,
  border.color = "#CCCCCC"
)
```

```{r}
#| label: load-data
df_raw <- read_csv(
  here::here("outputs", "master_combined_counts.csv"),
  col_types       = cols(.default = "c"),
  show_col_types  = FALSE
)

df <- df_raw |>
  mutate(
    date        = dmy(count_end_date),
    suspect     = as.numeric(cases_suspect),
    probable    = as.numeric(cases_probable),
    confirmed   = as.numeric(cases_confirmed),
    deaths      = coalesce(as.numeric(deaths_suspected), as.numeric(deaths_probable)),
    is_agg      = (is_aggregate == "TRUE"),
    # Only sum to zero when at least one component is present; otherwise NA
    # (avoids a false zero for zones that simply weren't reported in that sitrep)
    unconfirmed = if_else(
      !is.na(suspect) | !is.na(probable),
      coalesce(suspect, 0L) + coalesce(probable, 0L),
      NA_real_
    )
  )

ZONE_LEVELS <- c(
  "Mongbwalu", "Rwampara", "Bunia", "Nyankunde", "Bambu",
  "Butembo", "Katwa", "Goma"
)

# Colorblind-safe palette — Wong (2011)
PAL <- c(
  "Suspect/probable cases" = "#56B4E9",
  "Confirmed cases"        = "#D55E00",
  "Outbreak deaths"        = "#404040"
)

int_breaks <- function(x) unique(floor(pretty(x)))

theme_sitrep <- function(base_size = 11) {
  theme_minimal(base_size = base_size) %+replace%
    theme(
      plot.title       = element_text(face = "bold", size = base_size + 2, hjust = 0),
      plot.subtitle    = element_text(size = base_size - 1, colour = "grey45", hjust = 0,
                                      margin = margin(b = 8)),
      plot.caption     = element_text(size = base_size - 3, colour = "grey55", hjust = 1,
                                      margin = margin(t = 8)),
      plot.margin      = margin(12, 16, 8, 12),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(colour = "grey92"),
      legend.position  = "bottom",
      legend.key.size  = unit(0.75, "lines"),
      legend.text      = element_text(size = base_size - 1),
      strip.text       = element_text(face = "bold", size = base_size - 1,
                                      margin = margin(b = 4)),
      axis.title       = element_text(size = base_size - 1, colour = "grey35"),
      axis.text        = element_text(size = base_size - 2, colour = "grey25")
    )
}

# ── Derived subsets ────────────────────────────────────────────────────────────
totals <- df |>
  filter(count_type == "Cumules", is_agg, zone == "Total") |>
  group_by(date) |>
  slice_max(order_by = unconfirmed, n = 1, with_ties = FALSE) |>
  ungroup() |>
  arrange(date) |>
  mutate(deaths = cummax(replace_na(deaths, 0L)))

zone_ts <- df |>
  filter(count_type == "Cumules", !is_agg, zone %in% ZONE_LEVELS) |>
  group_by(date, zone, province) |>
  slice_max(order_by = unconfirmed, n = 1, with_ties = FALSE) |>
  ungroup() |>
  mutate(zone = factor(zone, levels = ZONE_LEVELS)) |>
  arrange(zone, date) |>
  group_by(zone) |>
  mutate(
    # Flag raw downward revisions before enforcing monotonicity
    confirmed_raw   = confirmed,
    unconfirmed_raw = unconfirmed,
    rev_confirmed   = confirmed   < lag(cummax(replace_na(confirmed,   0L)), default = 0L),
    rev_unconfirmed = unconfirmed < lag(cummax(replace_na(unconfirmed, 0L)), default = 0L),
    # Apply cummax only over observed values; leave NAs before the first
    # real observation so geom_line breaks rather than drawing from zero
    unconfirmed = {x <- unconfirmed; i <- !is.na(x); if (any(i)) x[i] <- cummax(x[i]); x},
    confirmed   = {x <- confirmed;   i <- !is.na(x); if (any(i)) x[i] <- cummax(x[i]); x},
    deaths      = {x <- deaths;      i <- !is.na(x); if (any(i)) x[i] <- cummax(x[i]); x}
  ) |>
  ungroup()

# Warn about any downward revisions that cummax suppressed
revisions <- zone_ts |>
  filter(rev_confirmed | rev_unconfirmed) |>
  transmute(
    zone, date,
    confirmed_raw, confirmed,
    unconfirmed_raw, unconfirmed
  )
if (nrow(revisions) > 0) {
  message("NOTE: cummax suppressed ", nrow(revisions),
          " downward revision(s) in cumulative counts:\n",
          paste(capture.output(print(revisions)), collapse = "\n"))
}

zone_new <- df |>
  filter(count_type == "Nouveaux", !is_agg, zone %in% ZONE_LEVELS) |>
  group_by(date, zone, province) |>
  slice_max(order_by = unconfirmed, n = 1, with_ties = FALSE) |>
  ungroup() |>
  mutate(zone = factor(zone, levels = ZONE_LEVELS)) |>
  arrange(zone, date)

daily_new <- df |>
  filter(count_type == "Nouveaux", is_agg, zone == "Total") |>
  group_by(date) |>
  slice_max(order_by = unconfirmed, n = 1, with_ties = FALSE) |>
  ungroup() |>
  arrange(date)

make_zone_labels <- function(dat) {
  dat |>
    mutate(
      zone_label = paste0(zone, " (", province, ")"),
      zone_label = factor(zone_label,
        levels = dat |>
          distinct(zone, province) |>
          arrange(zone) |>
          mutate(lbl = paste0(zone, " (", province, ")")) |>
          pull(lbl)
      )
    )
}

# ── Inline summary statistics ──────────────────────────────────────────────────
latest_total <- totals |> slice_max(date, n = 1)
latest_date  <- format(max(totals$date, na.rm = TRUE), "%d %B %Y")
n_unconf     <- latest_total$unconfirmed
n_conf       <- latest_total$confirmed
n_deaths     <- coalesce(latest_total$deaths, 0L)
n_total      <- n_unconf + n_conf
cfr_pct      <- if (n_total > 0) round(n_deaths / n_total * 100, 1) else NA_real_
n_zones      <- n_distinct(zone_ts$zone)
n_provinces  <- n_distinct(zone_ts$province)
```

```{r}
#| label: load-kraemer
# Kraemer lab canonical name → this extraction zone name
KRAEMER_NAME_MAP <- c("Mongbalu" = "Mongbwalu", "Nyakunde" = "Nyankunde")

read_kraemer_long <- function(filename) {
  read_csv(
    here::here("Ebola_DRC_2026", "build", "long", filename),
    col_types      = cols(.default = "c"),
    show_col_types = FALSE
  )
}

kraemer <- read_kraemer_long("insp_sitrep__cumulative_confirmed_cases.csv") |>
  rename(confirmed = cumulative_confirmed_cases) |>
  full_join(
    read_kraemer_long("insp_sitrep__cumulative_suspected_cases.csv") |>
      rename(suspected = cumulative_suspected_cases),
    by = c("nom", "date")
  ) |>
  full_join(
    read_kraemer_long("insp_sitrep__cumulative_suspected_deaths.csv") |>
      rename(deaths = cumulative_suspected_deaths),
    by = c("nom", "date")
  ) |>
  mutate(
    zone      = recode(nom, !!!KRAEMER_NAME_MAP),
    date      = as.Date(date),
    confirmed = suppressWarnings(as.numeric(confirmed)),
    suspected = suppressWarnings(as.numeric(suspected)),
    deaths    = suppressWarnings(as.numeric(deaths)),
    source    = "Kraemer lab"
  ) |>
  select(zone, date, confirmed, suspected, deaths, source) |>
  filter(zone %in% ZONE_LEVELS)

kraemer_hosp <- read_kraemer_long("insp_sitrep__hospitalised.csv") |>
  mutate(
    zone         = recode(nom, !!!KRAEMER_NAME_MAP),
    date         = as.Date(date),
    hospitalised = suppressWarnings(as.numeric(hospitalised))
  )

kraemer_contacts <- read_kraemer_long("insp_sitrep__cumulative_contacts_traced.csv") |>
  mutate(
    zone            = recode(nom, !!!KRAEMER_NAME_MAP),
    date            = as.Date(date),
    contacts_traced = suppressWarnings(as.numeric(cumulative_contacts_traced))
  ) |>
  filter(!is.na(contacts_traced))
```

## Introduction

The Democratic Republic of Congo is experiencing an outbreak of Maladie à Virus Ebola (MVE) with cases reported across health zones in Ituri and Nord-Kivu provinces. This report summarises the epidemiological situation based on data extracted from official INSP DRC situation reports. It is generated automatically and reflects the most recent available data.

## Methods

Case count data were extracted from PDF situation reports published by the Institut National de Santé Publique (INSP) DRC using an automated extraction pipeline. Suspect and probable cases are combined into a single unconfirmed category, as individual situation reports differ in which field they populate. The reported death count combines deaths among suspect cases and deaths among probable cases, as situation reports are inconsistent in which column they use for the total — cumulative maxima are applied at both the zonal and aggregate level to enforce monotonicity across reporting dates. Cumulative counts are enforced to be monotonically non-decreasing using the cumulative maximum across successive reporting dates.

This report and the underlying extraction pipeline were developed with the assistance of GitHub Copilot (Microsoft) and the Claude API (Anthropic, model: claude-sonnet-4-6).

## Results

As of `r latest_date`, the DRC MVE outbreak had recorded a cumulative total of `r prettyNum(n_total, big.mark = ",")` cases (`r prettyNum(n_conf, big.mark = ",")` confirmed, `r prettyNum(n_unconf, big.mark = ",")` suspect or probable) and `r prettyNum(n_deaths, big.mark = ",")` deaths, corresponding to a case fatality ratio of `r cfr_pct`%. Cases have been reported across `r n_zones` health zones in `r n_provinces` provinces.

### Cumulative epidemic curve

```{r}
#| label: fig-cumulative
#| fig-cap: "Cumulative MVE cases and deaths, DRC. Suspect and probable cases are combined. End-of-line labels show the latest recorded value. Source: INSP DRC."
#| fig-width: 8
#| fig-height: 5
p1_dat <- totals |>
  select(date,
         `Suspect/probable cases` = unconfirmed,
         `Confirmed cases`        = confirmed,
         `Outbreak deaths`        = deaths) |>
  pivot_longer(-date, names_to = "category", values_to = "count") |>
  mutate(category = factor(category, levels = names(PAL)))

p1_last <- p1_dat |>
  filter(!is.na(count)) |>
  group_by(category) |>
  slice_max(date, n = 1) |>
  ungroup()

p <- ggplot(p1_dat, aes(x = date, y = count, colour = category, group = category)) +
  geom_line(linewidth = 1.1, na.rm = TRUE) +
  geom_point(size = 2.8, na.rm = TRUE) +
  geom_label_repel(
    data           = p1_last,
    aes(label      = count),
    size           = 3.5,
    nudge_x        = 1.5,
    direction      = "y",
    hjust          = 0,
    segment.colour = "grey70",
    segment.size   = 0.3,
    show.legend    = FALSE,
    na.rm          = TRUE
  ) +
  scale_colour_manual(values = PAL) +
  scale_x_date(
    date_labels = "%d %b",
    date_breaks = "2 days",
    expand      = expansion(add = c(0, 4))
  ) +
  scale_y_continuous(breaks = int_breaks, expand = expansion(mult = c(0, 0.08))) +
  labs(
    title    = "MVE DRC \u2014 Cumulative epidemic curve",
    subtitle = "Total counts across all health zones",
    x = NULL, y = "Cumulative count", colour = NULL,
    caption  = "Source: INSP DRC situation reports"
  ) +
  theme_sitrep(base_size = 12)
ggsave(here("outputs", "plots", "fig-cumulative.png"), plot = p, width = 8, height = 5, dpi = 300)
p
```

### Daily new cases

```{r}
#| label: fig-new-cases
#| fig-cap: "New cases reported per situation report across all health zones, DRC. Source: INSP DRC."
#| fig-width: 7
#| fig-height: 4
p <- daily_new |>
  select(date,
         `Suspect/probable cases` = unconfirmed,
         `Confirmed cases`        = confirmed) |>
  pivot_longer(-date, names_to = "category", values_to = "count") |>
  mutate(category = factor(category, levels = c("Suspect/probable cases", "Confirmed cases"))) |>
  ggplot(aes(x = date, y = count, fill = category)) +
  geom_col(position = "dodge", width = 0.3, na.rm = TRUE) +
  scale_fill_manual(values = PAL) +
  scale_x_date(date_labels = "%d %b", date_breaks = "1 day") +
  scale_y_continuous(breaks = int_breaks, expand = expansion(mult = c(0, 0.1))) +
  labs(
    title    = "MVE DRC \u2014 Daily new cases",
    subtitle = "Total reported per situation report",
    x = NULL, y = "New cases", fill = NULL,
    caption  = "Source: INSP DRC situation reports"
  ) +
  theme_sitrep(base_size = 12)
ggsave(here("outputs", "plots", "fig-new-cases.png"), plot = p, width = 7, height = 4, dpi = 300)
p
```

### Geographic distribution — cumulative counts

```{r}
#| label: fig-zone-cumulative
#| fig-cap: "Cumulative cases by health zone. Y-axis is free across panels. Source: INSP DRC."
#| fig-width: 10
#| fig-height: 11
p <- zone_ts |>
  make_zone_labels() |>
  select(date, zone_label,
         `Suspect/probable cases` = unconfirmed,
         `Confirmed cases`        = confirmed,
         `Outbreak deaths`        = deaths) |>
  pivot_longer(
    cols      = c(`Suspect/probable cases`, `Confirmed cases`, `Outbreak deaths`),
    names_to  = "category",
    values_to = "count"
  ) |>
  mutate(category = factor(category, levels = names(PAL))) |>
  ggplot(aes(x = date, y = count, colour = category, group = category)) +
  geom_line(linewidth = 0.9, na.rm = TRUE) +
  geom_point(size = 2.2, na.rm = TRUE) +
  scale_colour_manual(values = PAL) +
  scale_x_date(date_labels = "%d %b", date_breaks = "1 day") +
  scale_y_continuous(breaks = int_breaks, expand = expansion(mult = c(0, 0.15))) +
  facet_wrap(~ zone_label, scales = "free_y", ncol = 2) +
  labs(
    title    = "MVE DRC \u2014 Cumulative cases by health zone",
    subtitle = "Each panel shows one health zone; y-axis is free.",
    x = NULL, y = "Cumulative count", colour = NULL,
    caption  = "Source: INSP DRC situation reports"
  ) +
  theme_sitrep(base_size = 10) +
  theme(axis.text.x = element_text(size = 8, angle = 45, hjust = 1))
ggsave(here("outputs", "plots", "fig-zone-cumulative.png"), plot = p, width = 10, height = 11, dpi = 300)
p
```

### Geographic distribution — new cases

```{r}
#| label: fig-zone-new
#| fig-cap: "New cases per situation report by health zone. Y-axis is free across panels. Source: INSP DRC."
#| fig-width: 10
#| fig-height: 11
p <- zone_new |>
  make_zone_labels() |>
  select(date, zone_label,
         `Suspect/probable cases` = unconfirmed,
         `Confirmed cases`        = confirmed) |>
  pivot_longer(
    cols      = c(`Suspect/probable cases`, `Confirmed cases`),
    names_to  = "category",
    values_to = "count"
  ) |>
  mutate(category = factor(category, levels = c("Suspect/probable cases", "Confirmed cases"))) |>
  ggplot(aes(x = date, y = count, fill = category)) +
  geom_col(position = "dodge", width = 0.4, na.rm = TRUE) +
  scale_fill_manual(values = PAL) +
  scale_x_date(date_labels = "%d %b", date_breaks = "1 day") +
  scale_y_continuous(breaks = int_breaks, expand = expansion(mult = c(0, 0.15))) +
  facet_wrap(~ zone_label, scales = "free_y", ncol = 2) +
  labs(
    title    = "MVE DRC \u2014 Daily new cases by health zone",
    subtitle = "Per situation report; y-axis is free.",
    x = NULL, y = "New cases", fill = NULL,
    caption  = "Source: INSP DRC situation reports"
  ) +
  theme_sitrep(base_size = 10) +
  theme(axis.text.x = element_text(size = 8, angle = 45, hjust = 1))
ggsave(here("outputs", "plots", "fig-zone-new.png"), plot = p, width = 10, height = 11, dpi = 300)
p
```

### Current counts by health zone

Table 1 presents the most recent cumulative case counts disaggregated by health zone, drawn from the latest available situation report for each zone.

```{r}
#| label: tbl-zones
#| tbl-cap: "Table 1. Latest cumulative case counts by health zone."
zone_latest <- zone_ts |>
  group_by(zone, province) |>
  slice_max(date, n = 1) |>
  ungroup() |>
  arrange(zone) |>
  transmute(
    `Health zone`      = as.character(zone),
    Province           = province,
    `Latest sitrep`    = format(date, "%d %b %Y"),
    `Suspect/probable` = unconfirmed,
    `Confirmed`        = confirmed,
    `Deaths`           = deaths
  )

total_row <- tibble(
  `Health zone`      = "Total",
  Province           = "\u2014",
  `Latest sitrep`    = format(max(totals$date, na.rm = TRUE), "%d %b %Y"),
  `Suspect/probable` = sum(zone_latest$`Suspect/probable`, na.rm = TRUE),
  `Confirmed`        = sum(zone_latest$`Confirmed`,        na.rm = TRUE),
  `Deaths`           = sum(zone_latest$`Deaths`,           na.rm = TRUE)
)

bind_rows(zone_latest, total_row) |>
  flextable() |>
  bold(part = "header") |>
  bold(i = nrow(zone_latest) + 1L) |>
  bg(part = "header", bg = "#1A3A5C") |>
  color(part = "header", color = "white") |>
  bg(i = nrow(zone_latest) + 1L, bg = "#E8EFF7") |>
  hline(i = nrow(zone_latest), part = "body") |>
  align(j = 4:6, align = "right", part = "all") |>
  colformat_int(j = 4:6) |>
  autofit()
```

### Cross-source validation

Cumulative zone-level counts from this pipeline are compared against the [Kraemer lab Ebola DRC 2026](https://github.com/kraemer-lab/Ebola_DRC_2026) repository, which codes INSP DRC situation reports into a standardised long-format by manual human extraction. National INSP SitReps N°001, N°002, and N°004, extracted separately from PDFs obtained via the Kraemer lab repository, are shown as a third series. Zone names differ between sources at two sites (Mongbwalu/Mongbalu and Nyankunde/Nyakunde); counts are matched on the harmonised name. `ND` values in the Kraemer lab series are treated as missing.

```{r}
#| label: fig-comparison
#| fig-cap: "Cross-source comparison of cumulative confirmed cases, suspect/probable cases, and outbreak deaths by health zone. Solid lines: automated extraction (provincial docs for early sitreps); dotted lines: national INSP SitReps N°001/002/004 (this pipeline); dashed lines: Kraemer lab."
#| fig-width: 12
#| fig-height: 11
PAL_COMP <- c(
  "Automated extraction" = "#0072B2",
  "National INSP (early)" = "#009E73",
  "Kraemer lab"           = "#E69F00"
)
LTY_COMP <- c(
  "Automated extraction"  = "solid",
  "National INSP (early)" = "dotted",
  "Kraemer lab"           = "dashed"
)

our_for_comp <- zone_ts |>
  transmute(zone, date, confirmed, suspected = unconfirmed, deaths,
            source = "Automated extraction")

national_for_comp <- df |>
  filter(
    count_type == "Cumules", !is_agg, zone %in% ZONE_LEVELS,
    sitrep_source %in% c(
      "MVE_SitRep_001_national_2026-05-14",
      "MVE_SitRep_002_national_2026-05-15",
      "MVE_SitRep_004_national_2026-05-18"
    )
  ) |>
  group_by(date, zone) |>
  slice_max(order_by = unconfirmed, n = 1, with_ties = FALSE) |>
  ungroup() |>
  transmute(zone, date, confirmed, suspected = unconfirmed, deaths,
            source = "National INSP (early)")

comp_long <- bind_rows(our_for_comp, national_for_comp, kraemer) |>
  pivot_longer(c(confirmed, suspected, deaths), names_to = "metric", values_to = "count") |>
  mutate(
    metric = recode(metric,
      "confirmed" = "Confirmed cases",
      "suspected" = "Suspect/probable cases",
      "deaths"    = "Outbreak deaths"
    ),
    source = factor(source, levels = c("Automated extraction", "National INSP (early)", "Kraemer lab"))
  ) |>
  filter(!is.na(count))

p <- ggplot(comp_long, aes(x = date, y = count, colour = source, linetype = source)) +
  geom_line(linewidth = 0.9, na.rm = TRUE) +
  geom_point(size = 2.0, na.rm = TRUE) +
  facet_grid(metric ~ zone, scales = "free_y") +
  scale_colour_manual(values = PAL_COMP) +
  scale_linetype_manual(values = LTY_COMP) +
  scale_x_date(date_labels = "%d %b", date_breaks = "2 days") +
  scale_y_continuous(breaks = int_breaks, limits = c(0, NA), expand = expansion(mult = c(0, 0.15))) +
  labs(
    title    = "MVE DRC \u2014 Cross-source validation",
    subtitle = "Cumulative counts: automated extraction vs national INSP (early) vs Kraemer lab",
    x = NULL, y = "Cumulative count", colour = NULL, linetype = NULL,
    caption  = "Sources: INSP DRC situation reports; github.com/kraemer-lab/Ebola_DRC_2026"
  ) +
  theme_sitrep(base_size = 9) +
  theme(
    axis.text.x     = element_text(angle = 45, hjust = 1, size = 7),
    panel.border    = element_rect(colour = "grey70", fill = NA, linewidth = 0.4),
    panel.spacing   = unit(0.6, "lines")
  )
ggsave(here("outputs", "plots", "fig-comparison.png"), plot = p, width = 12, height = 11, dpi = 300)
p
```

Table 2 presents numerical differences at the most recent date available in each source for each zone.

```{r}
#| label: tbl-comparison
#| tbl-cap: "Table 2. Cumulative case counts at the latest available date per zone: automated extraction vs Kraemer lab. Delta = ours minus Kraemer. Dashes indicate no data in that source."
ours_latest <- zone_ts |>
  group_by(zone) |>
  slice_max(date, n = 1) |>
  ungroup() |>
  transmute(zone, conf_ours = confirmed, susp_ours = unconfirmed, dead_ours = deaths)

kraemer_latest <- kraemer |>
  group_by(zone) |>
  slice_max(date, n = 1) |>
  ungroup() |>
  transmute(zone, conf_them = confirmed, susp_them = suspected, dead_them = deaths)

full_join(ours_latest, kraemer_latest, by = "zone") |>
  mutate(
    d_conf = conf_ours - conf_them,
    d_susp = susp_ours - susp_them,
    d_dead = dead_ours - dead_them
  ) |>
  arrange(zone) |>
  transmute(
    zone,
    conf_ours, conf_them, d_conf,
    susp_ours, susp_them, d_susp,
    dead_ours, dead_them, d_dead
  ) |>
  setNames(c(
    "Health zone",
    "Confirmed (ours)", "Confirmed (Kraemer)", "\u0394 confirmed",
    "Suspect/prob. (ours)", "Suspect/prob. (Kraemer)", "\u0394 suspect/prob.",
    "Deaths (ours)", "Deaths (Kraemer)", "\u0394 deaths"
  )) |>
  flextable() |>
  bold(part = "header") |>
  bg(part = "header", bg = "#1A3A5C") |>
  color(part = "header", color = "white") |>
  align(j = 2:10, align = "right", part = "all") |>
  colformat_int(j = 2:10, na_str = "\u2014") |>
  autofit()
```

### Early sitrep source comparison

For SitReps N°001, N°002, and N°004, the INSP DRC website hosted Ituri DPS provincial documents rather than the national INSP reports. The national documents were obtained from the Kraemer lab repository and extracted separately. Table 3 presents cumulative suspect/probable cases, confirmed cases, and deaths by health zone from each pair of documents. Provincial figures use the *probabilités* (probable cases) column; national figures use *cas suspects* (combined suspect and probable). Zones absent from the provincial documents appear with missing values.

```{r}
#| label: tbl-early-comp
#| tbl-cap: "Table 3. Ituri DPS provincial documents vs national INSP documents for SitReps N°001, N°002 and N°004: cumulative suspect/probable cases, confirmed cases, and deaths by health zone. Delta = national minus provincial."
PROV_SOURCES <- c(
  "Draft-MVE-Ituri-ActuelB-1",
  "Sitrep_MVE-Ituri_N\u00b0-002_18-mai-2026",
  "Sitrep_MVE-Ituri_N\u00b0004-au-20-mai-2026"
)

NAT_SOURCES <- c(
  "MVE_SitRep_001_national_2026-05-14",
  "MVE_SitRep_002_national_2026-05-15",
  "MVE_SitRep_004_national_2026-05-18"
)

SLOT_MAP <- c(
  "Draft-MVE-Ituri-ActuelB-1"                  = "N\u00b0001",
  "Sitrep_MVE-Ituri_N\u00b0-002_18-mai-2026"   = "N\u00b0002",
  "Sitrep_MVE-Ituri_N\u00b0004-au-20-mai-2026" = "N\u00b0004",
  "MVE_SitRep_001_national_2026-05-14"          = "N\u00b0001",
  "MVE_SitRep_002_national_2026-05-15"          = "N\u00b0002",
  "MVE_SitRep_004_national_2026-05-18"          = "N\u00b0004"
)

early_prov <- df |>
  filter(count_type == "Cumules", !is_agg, sitrep_source %in% PROV_SOURCES) |>
  mutate(slot = recode(sitrep_source, !!!SLOT_MAP)) |>
  group_by(slot, zone) |>
  slice_max(date, n = 1, with_ties = FALSE) |>
  ungroup() |>
  transmute(
    slot, zone,
    prov_date      = date,
    prov_suspect   = unconfirmed,
    prov_confirmed = confirmed,
    prov_deaths    = deaths
  )

early_nat <- df |>
  filter(count_type == "Cumules", !is_agg, sitrep_source %in% NAT_SOURCES) |>
  mutate(slot = recode(sitrep_source, !!!SLOT_MAP)) |>
  group_by(slot, zone) |>
  slice_max(date, n = 1, with_ties = FALSE) |>
  ungroup() |>
  transmute(
    slot, zone,
    nat_date      = date,
    nat_suspect   = unconfirmed,
    nat_confirmed = confirmed,
    nat_deaths    = deaths
  )

full_join(early_nat, early_prov, by = c("slot", "zone")) |>
  mutate(d_suspect = nat_suspect - prov_suspect) |>
  arrange(slot, zone) |>
  transmute(
    `SitRep`         = slot,
    `Health zone`    = as.character(zone),
    `Prov. date`     = format(prov_date, "%d %b"),
    `Prov. suspect`  = prov_suspect,
    `Prov. confirm.` = prov_confirmed,
    `Prov. deaths`   = prov_deaths,
    `Natl. date`     = format(nat_date, "%d %b"),
    `Natl. suspect`  = nat_suspect,
    `Natl. confirm.` = nat_confirmed,
    `Natl. deaths`   = nat_deaths,
    `Delta suspect`  = d_suspect
  ) |>
  flextable() |>
  bold(part = "header") |>
  bg(part = "header", bg = "#1A3A5C") |>
  color(part = "header", color = "white") |>
  merge_v(j = "SitRep") |>
  valign(j = "SitRep", valign = "top", part = "body") |>
  align(j = 3:11, align = "right", part = "all") |>
  colformat_int(j = 4:11, na_str = "\u2014") |>
  autofit()
```

### Response indicators

```{r}
#| label: fig-hospitalised
#| fig-cap: "Patients currently hospitalised by health zone, 20–23 May 2026. Source: Kraemer lab / INSP DRC situation reports."
#| fig-width: 8
#| fig-height: 5
ZONE_COLOURS <- c(
  "Mongbwalu" = "#56B4E9",
  "Rwampara"  = "#E69F00",
  "Bunia"     = "#009E73",
  "Nyankunde" = "#D55E00",
  "Aru"       = "#CC79A7"
)

kraemer_hosp |>
  filter(!is.na(hospitalised)) |>
  mutate(zone = factor(zone, levels = names(ZONE_COLOURS))) |>
  ggplot(aes(x = date, y = hospitalised, colour = zone, group = zone)) +
  geom_line(linewidth = 1.1, na.rm = TRUE) +
  geom_point(size = 2.8, na.rm = TRUE) +
  scale_colour_manual(values = ZONE_COLOURS, na.translate = FALSE) +
  scale_x_date(date_labels = "%d %b", date_breaks = "1 day") +
  scale_y_continuous(breaks = int_breaks, limits = c(0, NA),
                     expand = expansion(mult = c(0, 0.12))) +
  labs(
    title    = "MVE DRC \u2014 Hospitalised patients",
    subtitle = "Patients currently in hospital per reporting day, by health zone",
    x = NULL, y = "Patients hospitalised", colour = NULL,
    caption  = "Source: Kraemer lab / INSP DRC situation reports"
  ) +
  theme_sitrep(base_size = 12)
```

```{r}
#| label: fig-contacts
#| fig-cap: "Cumulative contacts traced by health zone, 14–23 May 2026. ND values (not reported) are excluded. Source: Kraemer lab / INSP DRC situation reports."
#| fig-width: 8
#| fig-height: 5
CONTACTS_PAL <- c(
  "Mongbwalu" = "#56B4E9", "Rwampara"  = "#E69F00",
  "Bunia"     = "#009E73", "Nyankunde" = "#D55E00",
  "Bambu"     = "#88CCEE", "Aru"       = "#CC79A7",
  "Komanda"   = "#0072B2", "Mangala"   = "#999999",
  "Nizi"      = "#117733", "Kilo"      = "#882255"
)

kraemer_contacts |>
  mutate(zone = factor(zone, levels = names(CONTACTS_PAL))) |>
  ggplot(aes(x = date, y = contacts_traced, colour = zone, group = zone)) +
  geom_line(linewidth = 1.1, na.rm = TRUE) +
  geom_point(size = 2.8, na.rm = TRUE) +
  scale_colour_manual(values = CONTACTS_PAL, na.translate = FALSE) +
  scale_x_date(date_labels = "%d %b", date_breaks = "2 days") +
  scale_y_continuous(breaks = int_breaks, limits = c(0, NA),
                     expand = expansion(mult = c(0, 0.12))) +
  labs(
    title    = "MVE DRC \u2014 Cumulative contacts traced",
    subtitle = "Cumulative contacts traced per reporting day, by health zone",
    x = NULL, y = "Contacts traced (cumulative)", colour = NULL,
    caption  = "Source: Kraemer lab / INSP DRC situation reports"
  ) +
  theme_sitrep(base_size = 12)
```

## Discussion

The outbreak trajectory shows a consistent increase in suspect and probable case counts across the reporting period. Confirmed cases remain a minority of the total, reflecting the challenges of laboratory confirmation in the field setting. The geographic spread across both Ituri and Nord-Kivu provinces underscores the need for multi-province coordination of the response.

Per-zone new case data are only available from 20 May 2026 onwards, as earlier situation reports did not disaggregate new cases by health zone. National INSP SitReps N°001, N°002, and N°004 were not available on the INSP website and were obtained from the Kraemer lab repository; Table 3 documents the systematic differences between these national documents and the Ituri DPS provincial documents that occupied the same URL slots. Response indicator data (hospitalisation, contacts traced) are sourced from the Kraemer lab repository as these metrics are not captured in the automated extraction pipeline. Results should be interpreted in the context of reporting delays and the possibility of incomplete situation report coverage for some health zones and dates. The case fatality ratio reported here should be treated as a provisional estimate, as delays in death reporting may affect the numerator, and the deaths column captures the first available of deaths-among-suspects and deaths-among-probables, which are not consistently labelled across situation reports.

Inspection of the raw PDF extractions and the Kraemer lab source documents identified four distinct reasons for divergence in the cross-source comparison. The most substantive concerns the source documents used for the first three sitrep slots (N°001–003). The INSP DRC website hosted Ituri DPS provincial sitreps at these URL positions rather than the national INSP documents. These provincial documents cover only the three main Ituri health zones and use the column heading *probabilités* (probable cases) at zone level, whereas the national INSP format uses *cas suspects* — a combined suspect and probable tally. As a result, zone-level unconfirmed case counts disagree substantially for dates prior to 19 May 2026: for example, Mongbwalu on 15 May is recorded as 57 probable cases in the Ituri DPS SitRep N°002, compared with 246 suspected cases in the Kraemer lab national SitRep N°002 — a figure coinciding with the aggregate suspect total across all Ituri zones in the provincial document. From SitRep N°005 (19 May 2026) onwards, both pipelines use the same nationally issued INSP documents, and confirmed case counts, suspected case counts, and deaths are near-identical between sources for this period.

A second source of divergence is that three of the PDFs processed here carry explicit draft labels in their filenames (`Draft-MVE-Ituri-ActuelB-1`, `Draft_SitRep_MVE_RDC_20260520_vf1`, `Draft_SitRep_MVE_RDC_2022vf2_IM`), whereas the Kraemer lab holds the corresponding final published versions; provisional counts in drafts may differ from the figures retained in the final document. Third, both pipelines assign the data reference date stated in the cumulative table title (for example, *au 21 mai 2026*) rather than the publication date, so date alignment between sources is consistent from SitRep N°005 onwards and is not itself a source of discrepancy for that period. Fourth, this pipeline enforces monotonic non-decrease via cumulative maximum, masking genuine downward revisions that are visible in the Kraemer series: Mongbwalu confirmed cases fall from 14 (SitRep N°007, 21 May) to 10 (SitRep N°008, 22 May) in the Kraemer data, most likely reflecting reclassification of four previously confirmed cases; this pipeline retains the cummax value of 14 at 22 May. Finally, the Kraemer lab data extend to SitRep N°008 (22 May), whereas this pipeline has additionally processed SitReps N°009 and N°010 (through 25 May), and several health zones present in the Kraemer series (Aru, Kilo, Nizi, Miti-Murhesa) are excluded from this pipeline's zone filter.
