---
title: "Maladie à Virus Ebola — Situation Report"
subtitle: "Democratic Republic of Congo"
date: today
date-format: "DD MMMM YYYY"
format:
  html:
    toc: true
    toc-depth: 2
    toc-title: "Contents"
    theme: flatly
    embed-resources: true
    fig-align: center
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")
pacman::p_load(tidyverse, lubridate, here, scales, ggrepel, flextable)

set_flextable_defaults(
  font.family = "Arial",
  font.size   = 10,
  border.color = "#CCCCCC"
)
```

```{r}
#| label: load-data
df_raw <- read_csv(
  here::here("outputs", "master_combined_counts.csv"),
  col_types       = cols(.default = "c"),
  show_col_types  = FALSE
)

df <- df_raw |>
  mutate(
    date        = dmy(count_end_date),
    suspect     = as.numeric(cases_suspect),
    probable    = as.numeric(cases_probable),
    confirmed   = as.numeric(cases_confirmed),
    deaths      = as.numeric(deaths_suspected),
    is_agg      = (is_aggregate == "TRUE"),
    unconfirmed = coalesce(suspect, 0L) + coalesce(probable, 0L)
  )

ZONE_LEVELS <- c(
  "Mongbwalu", "Rwampara", "Bunia", "Nyankunde", "Bambu",
  "Butembo", "Katwa", "Goma"
)

# Colorblind-safe palette — Wong (2011)
PAL <- c(
  "Suspect/probable cases" = "#56B4E9",
  "Confirmed cases"        = "#D55E00",
  "Suspected deaths"       = "#404040"
)

int_breaks <- function(x) unique(floor(pretty(x)))

theme_sitrep <- function(base_size = 11) {
  theme_minimal(base_size = base_size) %+replace%
    theme(
      plot.title       = element_text(face = "bold", size = base_size + 2, hjust = 0),
      plot.subtitle    = element_text(size = base_size - 1, colour = "grey45", hjust = 0,
                                      margin = margin(b = 8)),
      plot.caption     = element_text(size = base_size - 3, colour = "grey55", hjust = 1,
                                      margin = margin(t = 8)),
      plot.margin      = margin(12, 16, 8, 12),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(colour = "grey92"),
      legend.position  = "bottom",
      legend.key.size  = unit(0.75, "lines"),
      legend.text      = element_text(size = base_size - 1),
      strip.text       = element_text(face = "bold", size = base_size - 1,
                                      margin = margin(b = 4)),
      axis.title       = element_text(size = base_size - 1, colour = "grey35"),
      axis.text        = element_text(size = base_size - 2, colour = "grey25")
    )
}

# ── Derived subsets ────────────────────────────────────────────────────────────
totals <- df |>
  filter(count_type == "Cumules", is_agg, zone == "Total") |>
  group_by(date) |>
  slice_max(order_by = unconfirmed, n = 1, with_ties = FALSE) |>
  ungroup() |>
  arrange(date)

zone_ts <- df |>
  filter(count_type == "Cumules", !is_agg, zone %in% ZONE_LEVELS) |>
  group_by(date, zone, province) |>
  slice_max(order_by = unconfirmed, n = 1, with_ties = FALSE) |>
  ungroup() |>
  mutate(zone = factor(zone, levels = ZONE_LEVELS)) |>
  arrange(zone, date) |>
  group_by(zone) |>
  mutate(
    unconfirmed = cummax(replace_na(unconfirmed, 0L)),
    confirmed   = cummax(replace_na(confirmed,   0L)),
    deaths      = cummax(replace_na(deaths,      0L))
  ) |>
  ungroup()

zone_new <- df |>
  filter(count_type == "Nouveaux", !is_agg, zone %in% ZONE_LEVELS) |>
  group_by(date, zone, province) |>
  slice_max(order_by = unconfirmed, n = 1, with_ties = FALSE) |>
  ungroup() |>
  mutate(zone = factor(zone, levels = ZONE_LEVELS)) |>
  arrange(zone, date)

daily_new <- df |>
  filter(count_type == "Nouveaux", is_agg, zone == "Total") |>
  group_by(date) |>
  slice_max(order_by = unconfirmed, n = 1, with_ties = FALSE) |>
  ungroup() |>
  arrange(date)

make_zone_labels <- function(dat) {
  dat |>
    mutate(
      zone_label = paste0(zone, " (", province, ")"),
      zone_label = factor(zone_label,
        levels = dat |>
          distinct(zone, province) |>
          arrange(zone) |>
          mutate(lbl = paste0(zone, " (", province, ")")) |>
          pull(lbl)
      )
    )
}

# ── Inline summary statistics ──────────────────────────────────────────────────
latest_total <- totals |> slice_max(date, n = 1)
latest_date  <- format(max(totals$date, na.rm = TRUE), "%d %B %Y")
n_unconf     <- latest_total$unconfirmed
n_conf       <- latest_total$confirmed
n_deaths     <- coalesce(latest_total$deaths, 0L)
n_total      <- n_unconf + n_conf
cfr_pct      <- if (n_total > 0) round(n_deaths / n_total * 100, 1) else NA_real_
n_zones      <- n_distinct(zone_ts$zone)
n_provinces  <- n_distinct(zone_ts$province)
```

## Introduction

The Democratic Republic of Congo is experiencing an outbreak of Maladie à Virus Ebola (MVE) with cases reported across health zones in Ituri and Nord-Kivu provinces. This report summarises the epidemiological situation based on data extracted from official INSP DRC situation reports. It is generated automatically and reflects the most recent available data.

## Methods

Case count data were extracted from PDF situation reports published by the Institut National de Santé Publique (INSP) DRC using an automated extraction pipeline. Suspect and probable cases are combined into a single unconfirmed category, as individual situation reports differ in which field they populate. Cumulative counts are enforced to be monotonically non-decreasing using the cumulative maximum across successive reporting dates.

## Results

As of `r latest_date`, the DRC MVE outbreak had recorded a cumulative total of `r prettyNum(n_total, big.mark = ",")` cases (`r prettyNum(n_conf, big.mark = ",")` confirmed, `r prettyNum(n_unconf, big.mark = ",")` suspect or probable) and `r prettyNum(n_deaths, big.mark = ",")` deaths, corresponding to a case fatality ratio of `r cfr_pct`%. Cases have been reported across `r n_zones` health zones in `r n_provinces` provinces.

### Cumulative epidemic curve

```{r}
#| label: fig-cumulative
#| fig-cap: "Cumulative MVE cases and deaths, DRC. Suspect and probable cases are combined. End-of-line labels show the latest recorded value. Source: INSP DRC."
#| fig-width: 8
#| fig-height: 5
p1_dat <- totals |>
  select(date,
         `Suspect/probable cases` = unconfirmed,
         `Confirmed cases`        = confirmed,
         `Suspected deaths`       = deaths) |>
  pivot_longer(-date, names_to = "category", values_to = "count") |>
  mutate(category = factor(category, levels = names(PAL)))

p1_last <- p1_dat |>
  filter(!is.na(count)) |>
  group_by(category) |>
  slice_max(date, n = 1) |>
  ungroup()

ggplot(p1_dat, aes(x = date, y = count, colour = category, group = category)) +
  geom_line(linewidth = 1.1, na.rm = TRUE) +
  geom_point(size = 2.8, na.rm = TRUE) +
  geom_label_repel(
    data           = p1_last,
    aes(label      = count),
    size           = 3.5,
    nudge_x        = 1.5,
    direction      = "y",
    hjust          = 0,
    segment.colour = "grey70",
    segment.size   = 0.3,
    show.legend    = FALSE,
    na.rm          = TRUE
  ) +
  scale_colour_manual(values = PAL) +
  scale_x_date(
    date_labels = "%d %b",
    date_breaks = "2 days",
    expand      = expansion(add = c(0, 4))
  ) +
  scale_y_continuous(breaks = int_breaks, expand = expansion(mult = c(0, 0.08))) +
  labs(
    title    = "MVE DRC \u2014 Cumulative epidemic curve",
    subtitle = "Total counts across all health zones",
    x = NULL, y = "Cumulative count", colour = NULL,
    caption  = "Source: INSP DRC situation reports"
  ) +
  theme_sitrep(base_size = 12)
```

### Daily new cases

```{r}
#| label: fig-new-cases
#| fig-cap: "New cases reported per situation report across all health zones, DRC. Source: INSP DRC."
#| fig-width: 7
#| fig-height: 4
daily_new |>
  select(date,
         `Suspect/probable cases` = unconfirmed,
         `Confirmed cases`        = confirmed) |>
  pivot_longer(-date, names_to = "category", values_to = "count") |>
  mutate(category = factor(category, levels = c("Suspect/probable cases", "Confirmed cases"))) |>
  ggplot(aes(x = date, y = count, fill = category)) +
  geom_col(position = "dodge", width = 0.3, na.rm = TRUE) +
  scale_fill_manual(values = PAL) +
  scale_x_date(date_labels = "%d %b", date_breaks = "1 day") +
  scale_y_continuous(breaks = int_breaks, expand = expansion(mult = c(0, 0.1))) +
  labs(
    title    = "MVE DRC \u2014 Daily new cases",
    subtitle = "Total reported per situation report",
    x = NULL, y = "New cases", fill = NULL,
    caption  = "Source: INSP DRC situation reports"
  ) +
  theme_sitrep(base_size = 12)
```

### Geographic distribution — cumulative counts

```{r}
#| label: fig-zone-cumulative
#| fig-cap: "Cumulative cases by health zone. Y-axis is free across panels. Source: INSP DRC."
#| fig-width: 10
#| fig-height: 11
zone_ts |>
  make_zone_labels() |>
  select(date, zone_label,
         `Suspect/probable cases` = unconfirmed,
         `Confirmed cases`        = confirmed,
         `Suspected deaths`       = deaths) |>
  pivot_longer(
    cols      = c(`Suspect/probable cases`, `Confirmed cases`, `Suspected deaths`),
    names_to  = "category",
    values_to = "count"
  ) |>
  mutate(category = factor(category, levels = names(PAL))) |>
  ggplot(aes(x = date, y = count, colour = category, group = category)) +
  geom_line(linewidth = 0.9, na.rm = TRUE) +
  geom_point(size = 2.2, na.rm = TRUE) +
  scale_colour_manual(values = PAL) +
  scale_x_date(date_labels = "%d %b", date_breaks = "1 day") +
  scale_y_continuous(breaks = int_breaks, expand = expansion(mult = c(0, 0.15))) +
  facet_wrap(~ zone_label, scales = "free_y", ncol = 2) +
  labs(
    title    = "MVE DRC \u2014 Cumulative cases by health zone",
    subtitle = "Each panel shows one health zone; y-axis is free.",
    x = NULL, y = "Cumulative count", colour = NULL,
    caption  = "Source: INSP DRC situation reports"
  ) +
  theme_sitrep(base_size = 10) +
  theme(axis.text.x = element_text(size = 8, angle = 45, hjust = 1))
```

### Geographic distribution — new cases

```{r}
#| label: fig-zone-new
#| fig-cap: "New cases per situation report by health zone. Y-axis is free across panels. Source: INSP DRC."
#| fig-width: 10
#| fig-height: 11
zone_new |>
  make_zone_labels() |>
  select(date, zone_label,
         `Suspect/probable cases` = unconfirmed,
         `Confirmed cases`        = confirmed) |>
  pivot_longer(
    cols      = c(`Suspect/probable cases`, `Confirmed cases`),
    names_to  = "category",
    values_to = "count"
  ) |>
  mutate(category = factor(category, levels = c("Suspect/probable cases", "Confirmed cases"))) |>
  ggplot(aes(x = date, y = count, fill = category)) +
  geom_col(position = "dodge", width = 0.4, na.rm = TRUE) +
  scale_fill_manual(values = PAL) +
  scale_x_date(date_labels = "%d %b", date_breaks = "1 day") +
  scale_y_continuous(breaks = int_breaks, expand = expansion(mult = c(0, 0.15))) +
  facet_wrap(~ zone_label, scales = "free_y", ncol = 2) +
  labs(
    title    = "MVE DRC \u2014 Daily new cases by health zone",
    subtitle = "Per situation report; y-axis is free.",
    x = NULL, y = "New cases", fill = NULL,
    caption  = "Source: INSP DRC situation reports"
  ) +
  theme_sitrep(base_size = 10) +
  theme(axis.text.x = element_text(size = 8, angle = 45, hjust = 1))
```

### Current counts by health zone

Table 1 presents the most recent cumulative case counts disaggregated by health zone, drawn from the latest available situation report for each zone.

```{r}
#| label: tbl-zones
#| tbl-cap: "Table 1. Latest cumulative case counts by health zone."
zone_latest <- zone_ts |>
  group_by(zone, province) |>
  slice_max(date, n = 1) |>
  ungroup() |>
  arrange(zone) |>
  transmute(
    `Health zone`      = as.character(zone),
    Province           = province,
    `Latest sitrep`    = format(date, "%d %b %Y"),
    `Suspect/probable` = unconfirmed,
    `Confirmed`        = confirmed,
    `Deaths`           = deaths
  )

total_row <- tibble(
  `Health zone`      = "Total",
  Province           = "\u2014",
  `Latest sitrep`    = format(max(totals$date, na.rm = TRUE), "%d %b %Y"),
  `Suspect/probable` = sum(zone_latest$`Suspect/probable`, na.rm = TRUE),
  `Confirmed`        = sum(zone_latest$`Confirmed`,        na.rm = TRUE),
  `Deaths`           = sum(zone_latest$`Deaths`,           na.rm = TRUE)
)

bind_rows(zone_latest, total_row) |>
  flextable() |>
  bold(part = "header") |>
  bold(i = nrow(zone_latest) + 1L) |>
  bg(part = "header", bg = "#1A3A5C") |>
  color(part = "header", color = "white") |>
  bg(i = nrow(zone_latest) + 1L, bg = "#E8EFF7") |>
  hline(i = nrow(zone_latest), part = "body") |>
  align(j = 4:6, align = "right", part = "all") |>
  colformat_int(j = 4:6) |>
  autofit()
```

## Discussion

The outbreak trajectory shows a consistent increase in suspect and probable case counts across the reporting period. Confirmed cases remain a minority of the total, reflecting the challenges of laboratory confirmation in the field setting. The geographic spread across both Ituri and Nord-Kivu provinces underscores the need for multi-province coordination of the response.

Per-zone new case data are only available from 20 May 2026 onwards, as earlier situation reports did not disaggregate new cases by health zone. Results should be interpreted in the context of reporting delays and the possibility of incomplete situation report coverage for some health zones and dates. The case fatality ratio reported here should be treated as a provisional estimate, as delays in death reporting may affect the numerator.
