---
title: "Maladie à Virus Ebola — Situation Report"
subtitle: "Democratic Republic of Congo"
date: today
date-format: "DD MMMM YYYY"
format:
  html:
    toc: true
    toc-depth: 2
    toc-title: "Contents"
    theme: flatly
    embed-resources: true
    fig-align: center
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")
pacman::p_load(tidyverse, lubridate, here, scales, ggrepel, flextable, ggh4x)

set_flextable_defaults(
  font.family = "Arial",
  font.size   = 10,
  border.color = "#CCCCCC"
)
```

```{r}
#| label: load-data
df_raw <- read_csv(
  here::here("outputs", "master_combined_counts.csv"),
  col_types       = cols(.default = "c"),
  show_col_types  = FALSE
)

df <- df_raw |>
  mutate(
    date        = dmy(count_end_date),
    suspect     = as.numeric(cases_suspect),
    probable    = as.numeric(cases_probable),
    confirmed   = as.numeric(cases_confirmed),
    deaths      = coalesce(as.numeric(deaths_suspected), as.numeric(deaths_probable)),
    is_agg      = (is_aggregate == "TRUE"),
    # Only sum to zero when at least one component is present; otherwise NA
    # (avoids a false zero for zones that simply weren't reported in that sitrep)
    unconfirmed = if_else(
      !is.na(suspect) | !is.na(probable),
      coalesce(suspect, 0L) + coalesce(probable, 0L),
      NA_real_
    )
  )

ZONE_LEVELS <- c(
  "Mongbwalu", "Rwampara", "Bunia", "Nyankunde", "Bambu",
  "Butembo", "Katwa", "Goma"
)

# Colorblind-safe palette — Wong (2011)
PAL <- c(
  "Suspect/probable cases" = "#56B4E9",
  "Confirmed cases"        = "#D55E00",
  "Outbreak deaths"        = "#404040"
)

int_breaks <- function(x) unique(floor(pretty(x)))

theme_sitrep <- function(base_size = 11) {
  theme_minimal(base_size = base_size) %+replace%
    theme(
      plot.title       = element_text(face = "bold", size = base_size + 2, hjust = 0),
      plot.subtitle    = element_text(size = base_size - 1, colour = "grey45", hjust = 0,
                                      margin = margin(b = 8)),
      plot.caption     = element_text(size = base_size - 3, colour = "grey55", hjust = 1,
                                      margin = margin(t = 8)),
      plot.margin      = margin(12, 16, 8, 12),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(colour = "grey92"),
      legend.position  = "bottom",
      legend.key.size  = unit(0.75, "lines"),
      legend.text      = element_text(size = base_size - 1),
      strip.text       = element_text(face = "bold", size = base_size - 1,
                                      margin = margin(b = 4)),
      axis.title       = element_text(size = base_size - 1, colour = "grey35"),
      axis.text        = element_text(size = base_size - 2, colour = "grey25")
    )
}

# ── Derived subsets ────────────────────────────────────────────────────────────
totals <- df |>
  filter(count_type == "Cumules", is_agg, zone == "Total") |>
  group_by(date) |>
  slice_max(order_by = unconfirmed, n = 1, with_ties = FALSE) |>
  ungroup() |>
  arrange(date) |>
  mutate(deaths = cummax(replace_na(deaths, 0L)))

zone_ts <- df |>
  filter(count_type == "Cumules", !is_agg, zone %in% ZONE_LEVELS) |>
  group_by(date, zone, province) |>
  slice_max(order_by = unconfirmed, n = 1, with_ties = FALSE) |>
  ungroup() |>
  mutate(zone = factor(zone, levels = ZONE_LEVELS)) |>
  arrange(zone, date) |>
  group_by(zone) |>
  mutate(
    # Flag raw downward revisions before enforcing monotonicity
    confirmed_raw   = confirmed,
    unconfirmed_raw = unconfirmed,
    rev_confirmed   = confirmed   < lag(cummax(replace_na(confirmed,   0L)), default = 0L),
    rev_unconfirmed = unconfirmed < lag(cummax(replace_na(unconfirmed, 0L)), default = 0L),
    # Apply cummax only over observed values; leave NAs before the first
    # real observation so geom_line breaks rather than drawing from zero
    unconfirmed = {x <- unconfirmed; i <- !is.na(x); if (any(i)) x[i] <- cummax(x[i]); x},
    confirmed   = {x <- confirmed;   i <- !is.na(x); if (any(i)) x[i] <- cummax(x[i]); x},
    deaths      = {x <- deaths;      i <- !is.na(x); if (any(i)) x[i] <- cummax(x[i]); x}
  ) |>
  ungroup()

# Warn about any downward revisions that cummax suppressed
revisions <- zone_ts |>
  filter(rev_confirmed | rev_unconfirmed) |>
  transmute(
    zone, date,
    confirmed_raw, confirmed,
    unconfirmed_raw, unconfirmed
  )
if (nrow(revisions) > 0) {
  message("NOTE: cummax suppressed ", nrow(revisions),
          " downward revision(s) in cumulative counts:\n",
          paste(capture.output(print(revisions)), collapse = "\n"))
}

zone_new <- df |>
  filter(count_type == "Nouveaux", !is_agg, zone %in% ZONE_LEVELS) |>
  group_by(date, zone, province) |>
  slice_max(order_by = unconfirmed, n = 1, with_ties = FALSE) |>
  ungroup() |>
  mutate(zone = factor(zone, levels = ZONE_LEVELS)) |>
  arrange(zone, date)

daily_new <- df |>
  filter(count_type == "Nouveaux", is_agg, zone == "Total") |>
  group_by(date) |>
  slice_max(order_by = unconfirmed, n = 1, with_ties = FALSE) |>
  ungroup() |>
  arrange(date)

make_zone_labels <- function(dat) {
  dat |>
    mutate(
      zone_label = paste0(zone, " (", province, ")"),
      zone_label = factor(zone_label,
        levels = dat |>
          distinct(zone, province) |>
          arrange(zone) |>
          mutate(lbl = paste0(zone, " (", province, ")")) |>
          pull(lbl)
      )
    )
}

# ── Inline summary statistics ──────────────────────────────────────────────────
latest_total <- totals |> slice_max(date, n = 1)
latest_date  <- format(max(totals$date, na.rm = TRUE), "%d %B %Y")
n_unconf     <- latest_total$unconfirmed
n_conf       <- latest_total$confirmed
n_deaths     <- coalesce(latest_total$deaths, 0L)
n_total      <- n_unconf + n_conf
cfr_pct      <- if (n_total > 0) round(n_deaths / n_total * 100, 1) else NA_real_
n_zones      <- n_distinct(zone_ts$zone)
n_provinces  <- n_distinct(zone_ts$province)
```

```{r}
#| label: load-kraemer
# Kraemer lab canonical name → this extraction zone name
KRAEMER_NAME_MAP <- c("Mongbalu" = "Mongbwalu", "Nyakunde" = "Nyankunde")

read_kraemer_long <- function(filename) {
  read_csv(
    here::here("Ebola_DRC_2026", "build", "long", filename),
    col_types      = cols(.default = "c"),
    show_col_types = FALSE
  )
}

kraemer <- read_kraemer_long("insp_sitrep__cumulative_confirmed_cases.csv") |>
  rename(confirmed = cumulative_confirmed_cases) |>
  full_join(
    read_kraemer_long("insp_sitrep__cumulative_suspected_cases.csv") |>
      rename(suspected = cumulative_suspected_cases),
    by = c("nom", "date")
  ) |>
  full_join(
    read_kraemer_long("insp_sitrep__cumulative_suspected_deaths.csv") |>
      rename(deaths = cumulative_suspected_deaths),
    by = c("nom", "date")
  ) |>
  mutate(
    zone      = recode(nom, !!!KRAEMER_NAME_MAP),
    date      = as.Date(date),
    confirmed = suppressWarnings(as.numeric(confirmed)),
    suspected = suppressWarnings(as.numeric(suspected)),
    deaths    = suppressWarnings(as.numeric(deaths)),
    source    = "Kraemer lab"
  ) |>
  select(zone, date, confirmed, suspected, deaths, source) |>
  filter(zone %in% ZONE_LEVELS)
```

## Introduction

The Democratic Republic of Congo is experiencing an outbreak of Maladie à Virus Ebola (MVE) with cases reported across health zones in Ituri and Nord-Kivu provinces. This report summarises the epidemiological situation based on data extracted from official INSP DRC situation reports. It is generated automatically and reflects the most recent available data.

## Methods

Case count data were extracted from PDF situation reports published by the Institut National de Santé Publique (INSP) DRC using an automated extraction pipeline. Suspect and probable cases are combined into a single unconfirmed category, as individual situation reports differ in which field they populate. The reported death count combines deaths among suspect cases and deaths among probable cases, as situation reports are inconsistent in which column they use for the total — cumulative maxima are applied at both the zonal and aggregate level to enforce monotonicity across reporting dates. Cumulative counts are enforced to be monotonically non-decreasing using the cumulative maximum across successive reporting dates.

## Results

As of `r latest_date`, the DRC MVE outbreak had recorded a cumulative total of `r prettyNum(n_total, big.mark = ",")` cases (`r prettyNum(n_conf, big.mark = ",")` confirmed, `r prettyNum(n_unconf, big.mark = ",")` suspect or probable) and `r prettyNum(n_deaths, big.mark = ",")` deaths, corresponding to a case fatality ratio of `r cfr_pct`%. Cases have been reported across `r n_zones` health zones in `r n_provinces` provinces.

### Cumulative epidemic curve

```{r}
#| label: fig-cumulative
#| fig-cap: "Cumulative MVE cases and deaths, DRC. Suspect and probable cases are combined. End-of-line labels show the latest recorded value. Source: INSP DRC."
#| fig-width: 8
#| fig-height: 5
p1_dat <- totals |>
  select(date,
         `Suspect/probable cases` = unconfirmed,
         `Confirmed cases`        = confirmed,
         `Outbreak deaths`        = deaths) |>
  pivot_longer(-date, names_to = "category", values_to = "count") |>
  mutate(category = factor(category, levels = names(PAL)))

p1_last <- p1_dat |>
  filter(!is.na(count)) |>
  group_by(category) |>
  slice_max(date, n = 1) |>
  ungroup()

p <- ggplot(p1_dat, aes(x = date, y = count, colour = category, group = category)) +
  geom_line(linewidth = 1.1, na.rm = TRUE) +
  geom_point(size = 2.8, na.rm = TRUE) +
  geom_label_repel(
    data           = p1_last,
    aes(label      = count),
    size           = 3.5,
    nudge_x        = 1.5,
    direction      = "y",
    hjust          = 0,
    segment.colour = "grey70",
    segment.size   = 0.3,
    show.legend    = FALSE,
    na.rm          = TRUE
  ) +
  scale_colour_manual(values = PAL) +
  scale_x_date(
    date_labels = "%d %b",
    date_breaks = "2 days",
    expand      = expansion(add = c(0, 4))
  ) +
  scale_y_continuous(breaks = int_breaks, expand = expansion(mult = c(0, 0.08))) +
  labs(
    title    = "MVE DRC \u2014 Cumulative epidemic curve",
    subtitle = "Total counts across all health zones",
    x = NULL, y = "Cumulative count", colour = NULL,
    caption  = "Source: INSP DRC situation reports"
  ) +
  theme_sitrep(base_size = 12)
ggsave(here("outputs", "plots", "fig-cumulative.png"), plot = p, width = 8, height = 5, dpi = 300)
p
```

### Daily new cases

```{r}
#| label: fig-new-cases
#| fig-cap: "New cases reported per situation report across all health zones, DRC. Source: INSP DRC."
#| fig-width: 7
#| fig-height: 4
p <- daily_new |>
  select(date,
         `Suspect/probable cases` = unconfirmed,
         `Confirmed cases`        = confirmed) |>
  pivot_longer(-date, names_to = "category", values_to = "count") |>
  mutate(category = factor(category, levels = c("Suspect/probable cases", "Confirmed cases"))) |>
  ggplot(aes(x = date, y = count, fill = category)) +
  geom_col(position = "dodge", width = 0.3, na.rm = TRUE) +
  scale_fill_manual(values = PAL) +
  scale_x_date(date_labels = "%d %b", date_breaks = "1 day") +
  scale_y_continuous(breaks = int_breaks, expand = expansion(mult = c(0, 0.1))) +
  labs(
    title    = "MVE DRC \u2014 Daily new cases",
    subtitle = "Total reported per situation report",
    x = NULL, y = "New cases", fill = NULL,
    caption  = "Source: INSP DRC situation reports"
  ) +
  theme_sitrep(base_size = 12)
ggsave(here("outputs", "plots", "fig-new-cases.png"), plot = p, width = 7, height = 4, dpi = 300)
p
```

### Geographic distribution — cumulative counts

```{r}
#| label: fig-zone-cumulative
#| fig-cap: "Cumulative cases by health zone. Y-axis is free across panels. Source: INSP DRC."
#| fig-width: 10
#| fig-height: 11
p <- zone_ts |>
  make_zone_labels() |>
  select(date, zone_label,
         `Suspect/probable cases` = unconfirmed,
         `Confirmed cases`        = confirmed,
         `Outbreak deaths`        = deaths) |>
  pivot_longer(
    cols      = c(`Suspect/probable cases`, `Confirmed cases`, `Outbreak deaths`),
    names_to  = "category",
    values_to = "count"
  ) |>
  mutate(category = factor(category, levels = names(PAL))) |>
  ggplot(aes(x = date, y = count, colour = category, group = category)) +
  geom_line(linewidth = 0.9, na.rm = TRUE) +
  geom_point(size = 2.2, na.rm = TRUE) +
  scale_colour_manual(values = PAL) +
  scale_x_date(date_labels = "%d %b", date_breaks = "1 day") +
  scale_y_continuous(breaks = int_breaks, expand = expansion(mult = c(0, 0.15))) +
  facet_wrap(~ zone_label, scales = "free_y", ncol = 2) +
  labs(
    title    = "MVE DRC \u2014 Cumulative cases by health zone",
    subtitle = "Each panel shows one health zone; y-axis is free.",
    x = NULL, y = "Cumulative count", colour = NULL,
    caption  = "Source: INSP DRC situation reports"
  ) +
  theme_sitrep(base_size = 10) +
  theme(axis.text.x = element_text(size = 8, angle = 45, hjust = 1))
ggsave(here("outputs", "plots", "fig-zone-cumulative.png"), plot = p, width = 10, height = 11, dpi = 300)
p
```

### Geographic distribution — new cases

```{r}
#| label: fig-zone-new
#| fig-cap: "New cases per situation report by health zone. Y-axis is free across panels. Source: INSP DRC."
#| fig-width: 10
#| fig-height: 11
p <- zone_new |>
  make_zone_labels() |>
  select(date, zone_label,
         `Suspect/probable cases` = unconfirmed,
         `Confirmed cases`        = confirmed) |>
  pivot_longer(
    cols      = c(`Suspect/probable cases`, `Confirmed cases`),
    names_to  = "category",
    values_to = "count"
  ) |>
  mutate(category = factor(category, levels = c("Suspect/probable cases", "Confirmed cases"))) |>
  ggplot(aes(x = date, y = count, fill = category)) +
  geom_col(position = "dodge", width = 0.4, na.rm = TRUE) +
  scale_fill_manual(values = PAL) +
  scale_x_date(date_labels = "%d %b", date_breaks = "1 day") +
  scale_y_continuous(breaks = int_breaks, expand = expansion(mult = c(0, 0.15))) +
  facet_wrap(~ zone_label, scales = "free_y", ncol = 2) +
  labs(
    title    = "MVE DRC \u2014 Daily new cases by health zone",
    subtitle = "Per situation report; y-axis is free.",
    x = NULL, y = "New cases", fill = NULL,
    caption  = "Source: INSP DRC situation reports"
  ) +
  theme_sitrep(base_size = 10) +
  theme(axis.text.x = element_text(size = 8, angle = 45, hjust = 1))
ggsave(here("outputs", "plots", "fig-zone-new.png"), plot = p, width = 10, height = 11, dpi = 300)
p
```

### Current counts by health zone

Table 1 presents the most recent cumulative case counts disaggregated by health zone, drawn from the latest available situation report for each zone.

```{r}
#| label: tbl-zones
#| tbl-cap: "Table 1. Latest cumulative case counts by health zone."
zone_latest <- zone_ts |>
  group_by(zone, province) |>
  slice_max(date, n = 1) |>
  ungroup() |>
  arrange(zone) |>
  transmute(
    `Health zone`      = as.character(zone),
    Province           = province,
    `Latest sitrep`    = format(date, "%d %b %Y"),
    `Suspect/probable` = unconfirmed,
    `Confirmed`        = confirmed,
    `Deaths`           = deaths
  )

total_row <- tibble(
  `Health zone`      = "Total",
  Province           = "\u2014",
  `Latest sitrep`    = format(max(totals$date, na.rm = TRUE), "%d %b %Y"),
  `Suspect/probable` = sum(zone_latest$`Suspect/probable`, na.rm = TRUE),
  `Confirmed`        = sum(zone_latest$`Confirmed`,        na.rm = TRUE),
  `Deaths`           = sum(zone_latest$`Deaths`,           na.rm = TRUE)
)

bind_rows(zone_latest, total_row) |>
  flextable() |>
  bold(part = "header") |>
  bold(i = nrow(zone_latest) + 1L) |>
  bg(part = "header", bg = "#1A3A5C") |>
  color(part = "header", color = "white") |>
  bg(i = nrow(zone_latest) + 1L, bg = "#E8EFF7") |>
  hline(i = nrow(zone_latest), part = "body") |>
  align(j = 4:6, align = "right", part = "all") |>
  colformat_int(j = 4:6) |>
  autofit()
```

### Cross-source validation

Cumulative zone-level counts from this pipeline are compared against the [Kraemer lab Ebola DRC 2026](https://github.com/kraemer-lab/Ebola_DRC_2026) repository, which independently processes the same INSP SitRep PDFs into a standardised long-format. Zone names differ between sources at two sites (Mongbwalu/Mongbalu and Nyankunde/Nyakunde); counts are matched on the harmonised name. `ND` values in the Kraemer lab series (metric not reported on that date) are treated as missing.

```{r}
#| label: fig-comparison
#| fig-cap: "Cross-source comparison of cumulative confirmed cases, suspect/probable cases, and outbreak deaths by health zone. Solid lines: automated extraction; dashed lines: Kraemer lab. Sources: INSP DRC situation reports; github.com/kraemer-lab/Ebola_DRC_2026."
#| fig-width: 12
#| fig-height: 11
PAL_COMP <- c("Automated extraction" = "#0072B2", "Kraemer lab" = "#E69F00")
LTY_COMP <- c("Automated extraction" = "solid",   "Kraemer lab" = "dashed")

our_for_comp <- zone_ts |>
  transmute(zone, date, confirmed, suspected = unconfirmed, deaths, source = "Automated extraction")

comp_long <- bind_rows(our_for_comp, kraemer) |>
  pivot_longer(c(confirmed, suspected, deaths), names_to = "metric", values_to = "count") |>
  mutate(
    metric = recode(metric,
      "confirmed" = "Confirmed cases",
      "suspected" = "Suspect/probable cases",
      "deaths"    = "Outbreak deaths"
    ),
    source = factor(source, levels = c("Automated extraction", "Kraemer lab"))
  ) |>
  filter(!is.na(count))

p <- ggplot(comp_long, aes(x = date, y = count, colour = source, linetype = source)) +
  geom_line(linewidth = 0.9, na.rm = TRUE) +
  geom_point(size = 2.0, na.rm = TRUE) +
  facet_grid(metric ~ zone, scales = "free_y") +
  scale_colour_manual(values = PAL_COMP) +
  scale_linetype_manual(values = LTY_COMP) +
  scale_x_date(date_labels = "%d %b", date_breaks = "2 days") +
  scale_y_continuous(breaks = int_breaks, limits = c(0, NA), expand = expansion(mult = c(0, 0.15))) +
  labs(
    title    = "MVE DRC \u2014 Cross-source validation",
    subtitle = "Cumulative counts: automated extraction vs Kraemer lab (kraemer-lab/Ebola_DRC_2026)",
    x = NULL, y = "Cumulative count", colour = NULL, linetype = NULL,
    caption  = "Sources: INSP DRC situation reports; github.com/kraemer-lab/Ebola_DRC_2026"
  ) +
  theme_sitrep(base_size = 9) +
  theme(
    axis.text.x     = element_text(angle = 45, hjust = 1, size = 7),
    panel.border    = element_rect(colour = "grey70", fill = NA, linewidth = 0.4),
    panel.spacing   = unit(0.6, "lines")
  )
ggsave(here("outputs", "plots", "fig-comparison.png"), plot = p, width = 12, height = 11, dpi = 300)
p
```

Table 2 presents numerical differences at the most recent date available in each source for each zone.

```{r}
#| label: tbl-comparison
#| tbl-cap: "Table 2. Cumulative case counts at the latest available date per zone: automated extraction vs Kraemer lab. Delta = ours minus Kraemer. Dashes indicate no data in that source."
ours_latest <- zone_ts |>
  group_by(zone) |>
  slice_max(date, n = 1) |>
  ungroup() |>
  transmute(zone, conf_ours = confirmed, susp_ours = unconfirmed, dead_ours = deaths)

kraemer_latest <- kraemer |>
  group_by(zone) |>
  slice_max(date, n = 1) |>
  ungroup() |>
  transmute(zone, conf_them = confirmed, susp_them = suspected, dead_them = deaths)

full_join(ours_latest, kraemer_latest, by = "zone") |>
  mutate(
    d_conf = conf_ours - conf_them,
    d_susp = susp_ours - susp_them,
    d_dead = dead_ours - dead_them
  ) |>
  arrange(zone) |>
  transmute(
    zone,
    conf_ours, conf_them, d_conf,
    susp_ours, susp_them, d_susp,
    dead_ours, dead_them, d_dead
  ) |>
  setNames(c(
    "Health zone",
    "Confirmed (ours)", "Confirmed (Kraemer)", "\u0394 confirmed",
    "Suspect/prob. (ours)", "Suspect/prob. (Kraemer)", "\u0394 suspect/prob.",
    "Deaths (ours)", "Deaths (Kraemer)", "\u0394 deaths"
  )) |>
  flextable() |>
  bold(part = "header") |>
  bg(part = "header", bg = "#1A3A5C") |>
  color(part = "header", color = "white") |>
  align(j = 2:10, align = "right", part = "all") |>
  colformat_int(j = 2:10, na_str = "\u2014") |>
  autofit()
```

## Discussion

The outbreak trajectory shows a consistent increase in suspect and probable case counts across the reporting period. Confirmed cases remain a minority of the total, reflecting the challenges of laboratory confirmation in the field setting. The geographic spread across both Ituri and Nord-Kivu provinces underscores the need for multi-province coordination of the response.

Per-zone new case data are only available from 20 May 2026 onwards, as earlier situation reports did not disaggregate new cases by health zone. No official SitRep 004 has been identified in either the INSP website archive or the Kraemer lab repository, representing a gap in the series between SitRep 003 (18 May) and SitRep 005 (19 May). Results should be interpreted in the context of reporting delays and the possibility of incomplete situation report coverage for some health zones and dates. The case fatality ratio reported here should be treated as a provisional estimate, as delays in death reporting may affect the numerator, and the deaths column captures the first available of deaths-among-suspects and deaths-among-probables, which are not consistently labelled across situation reports. Where the two independently extracted series diverge in the cross-source comparison, differences reflect the date-attribution convention adopted by each pipeline: the Kraemer lab assigns the filing date of the situation report, whereas this pipeline uses the data reference date stated in the cumulative table title. Once dates are aligned by document rather than by calendar date, confirmed case counts agree closely between sources.
