---
title: "Maladie à Virus Ebola — Situation Report"
subtitle: "Democratic Republic of Congo"
date: today
date-format: "DD MMMM YYYY"
format:
  html:
    toc: true
    toc-depth: 2
    toc-title: "Contents"
    theme: flatly
    embed-resources: true
    fig-align: center
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")
pacman::p_load(tidyverse, lubridate, here, scales, ggrepel, flextable, ggh4x)

set_flextable_defaults(
  font.family = "Arial",
  font.size   = 10,
  border.color = "#CCCCCC"
)
```

```{r}
#| label: load-data
df_raw <- read_csv(
  here::here("outputs", "master_combined_counts.csv"),
  col_types       = cols(.default = "c"),
  show_col_types  = FALSE
)

df <- df_raw |>
  mutate(
    date        = dmy(count_end_date),
    suspect     = as.numeric(cases_suspect),
    probable    = as.numeric(cases_probable),
    confirmed   = as.numeric(cases_confirmed),
    deaths      = coalesce(as.numeric(deaths_suspected), as.numeric(deaths_probable), as.numeric(deaths_confirmed)),
    is_agg      = (is_aggregate == "TRUE"),
    # Only sum to zero when at least one component is present; otherwise NA
    # (avoids a false zero for zones that simply weren't reported in that sitrep)
    unconfirmed = if_else(
      !is.na(suspect) | !is.na(probable),
      coalesce(suspect, 0L) + coalesce(probable, 0L),
      NA_real_
    )
  )

ZONE_LEVELS <- c(
  "Mongbwalu", "Rwampara", "Bunia", "Nyankunde", "Bambu",
  "Butembo", "Katwa", "Goma"
)

# Colorblind-safe palette — Wong (2011)
PAL <- c(
  "Suspect/probable cases" = "#56B4E9",
  "Confirmed cases"        = "#D55E00",
  "Outbreak deaths"        = "#404040"
)

int_breaks <- function(x) unique(floor(pretty(x)))

theme_sitrep <- function(base_size = 11) {
  theme_minimal(base_size = base_size) %+replace%
    theme(
      plot.title       = element_text(face = "bold", size = base_size + 2, hjust = 0),
      plot.subtitle    = element_text(size = base_size - 1, colour = "grey45", hjust = 0,
                                      margin = margin(b = 8)),
      plot.caption     = element_text(size = base_size - 3, colour = "grey55", hjust = 1,
                                      margin = margin(t = 8)),
      plot.margin      = margin(12, 16, 8, 12),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(colour = "grey92"),
      legend.position  = "bottom",
      legend.key.size  = unit(0.75, "lines"),
      legend.text      = element_text(size = base_size - 1),
      strip.text       = element_text(face = "bold", size = base_size - 1,
                                      margin = margin(b = 4)),
      axis.title       = element_text(size = base_size - 1, colour = "grey35"),
      axis.text        = element_text(size = base_size - 2, colour = "grey25")
    )
}

# ── Derived subsets ────────────────────────────────────────────────────────────
totals <- df |>
  filter(count_type == "Cumules", is_agg, zone == "Total") |>
  group_by(date) |>
  slice_max(order_by = unconfirmed, n = 1, with_ties = FALSE) |>
  ungroup() |>
  arrange(date) |>
  mutate(deaths = cummax(replace_na(deaths, 0L)))

zone_ts <- df |>
  filter(count_type == "Cumules", !is_agg, zone %in% ZONE_LEVELS) |>
  group_by(date, zone, province) |>
  slice_max(order_by = unconfirmed, n = 1, with_ties = FALSE) |>
  ungroup() |>
  mutate(zone = factor(zone, levels = ZONE_LEVELS)) |>
  arrange(zone, date) |>
  group_by(zone) |>
  mutate(
    # Flag raw downward revisions before enforcing monotonicity
    confirmed_raw   = confirmed,
    unconfirmed_raw = unconfirmed,
    rev_confirmed   = confirmed   < lag(cummax(replace_na(confirmed,   0L)), default = 0L),
    rev_unconfirmed = unconfirmed < lag(cummax(replace_na(unconfirmed, 0L)), default = 0L),
    # Apply cummax only over observed values; leave NAs before the first
    # real observation so geom_line breaks rather than drawing from zero
    unconfirmed = {x <- unconfirmed; i <- !is.na(x); if (any(i)) x[i] <- cummax(x[i]); x},
    confirmed   = {x <- confirmed;   i <- !is.na(x); if (any(i)) x[i] <- cummax(x[i]); x},
    deaths      = {x <- deaths;      i <- !is.na(x); if (any(i)) x[i] <- cummax(x[i]); x}
  ) |>
  ungroup()

# Warn about any downward revisions that cummax suppressed
revisions <- zone_ts |>
  filter(rev_confirmed | rev_unconfirmed) |>
  transmute(
    zone, date,
    confirmed_raw, confirmed,
    unconfirmed_raw, unconfirmed
  )
if (nrow(revisions) > 0) {
  message("NOTE: cummax suppressed ", nrow(revisions),
          " downward revision(s) in cumulative counts:\n",
          paste(capture.output(print(revisions)), collapse = "\n"))
}

zone_new <- df |>
  filter(count_type == "Nouveaux", !is_agg, zone %in% ZONE_LEVELS) |>
  group_by(date, zone, province) |>
  slice_max(order_by = unconfirmed, n = 1, with_ties = FALSE) |>
  ungroup() |>
  mutate(zone = factor(zone, levels = ZONE_LEVELS)) |>
  arrange(zone, date)

daily_new <- df |>
  filter(count_type == "Nouveaux", is_agg, zone == "Total") |>
  group_by(date) |>
  slice_max(order_by = unconfirmed, n = 1, with_ties = FALSE) |>
  ungroup() |>
  arrange(date)

make_zone_labels <- function(dat) {
  dat |>
    mutate(
      zone_label = paste0(zone, " (", province, ")"),
      zone_label = factor(zone_label,
        levels = dat |>
          distinct(zone, province) |>
          arrange(zone) |>
          mutate(lbl = paste0(zone, " (", province, ")")) |>
          pull(lbl)
      )
    )
}

# ── Inline summary statistics ──────────────────────────────────────────────────
latest_total <- totals |> slice_max(date, n = 1)
latest_date  <- format(max(totals$date, na.rm = TRUE), "%d %B %Y")
n_unconf     <- latest_total$unconfirmed
n_conf       <- latest_total$confirmed
n_deaths     <- coalesce(latest_total$deaths, 0L)
n_total      <- n_unconf + n_conf
cfr_pct      <- if (n_total > 0) round(n_deaths / n_total * 100, 1) else NA_real_
n_zones      <- n_distinct(zone_ts$zone)
n_provinces  <- n_distinct(zone_ts$province)
```

```{r}
#| label: load-kraemer
# Kraemer lab canonical name → this extraction zone name
KRAEMER_NAME_MAP <- c("Mongbalu" = "Mongbwalu", "Nyakunde" = "Nyankunde")

read_kraemer_long <- function(filename) {
  read_csv(
    here::here("Ebola_DRC_2026", "build", "long", filename),
    col_types      = cols(.default = "c"),
    show_col_types = FALSE
  )
}

kraemer <- read_kraemer_long("insp_sitrep__cumulative_confirmed_cases.csv") |>
  rename(confirmed = cumulative_confirmed_cases) |>
  full_join(
    read_kraemer_long("insp_sitrep__cumulative_suspected_cases.csv") |>
      rename(suspected = cumulative_suspected_cases),
    by = c("nom", "date")
  ) |>
  full_join(
    read_kraemer_long("insp_sitrep__cumulative_suspected_deaths.csv") |>
      rename(deaths = cumulative_suspected_deaths),
    by = c("nom", "date")
  ) |>
  mutate(
    zone      = recode(nom, !!!KRAEMER_NAME_MAP),
    date      = as.Date(date),
    confirmed = suppressWarnings(as.numeric(confirmed)),
    suspected = suppressWarnings(as.numeric(suspected)),
    deaths    = suppressWarnings(as.numeric(deaths)),
    source    = "Kraemer lab"
  ) |>
  select(zone, date, confirmed, suspected, deaths, source) |>
  filter(zone %in% ZONE_LEVELS)

kraemer_hosp <- read_kraemer_long("insp_sitrep__hospitalised.csv") |>
  mutate(
    zone         = recode(nom, !!!KRAEMER_NAME_MAP),
    date         = as.Date(date),
    hospitalised = suppressWarnings(as.numeric(hospitalised))
  )

kraemer_contacts <- read_kraemer_long("insp_sitrep__cumulative_contacts_traced.csv") |>
  mutate(
    zone            = recode(nom, !!!KRAEMER_NAME_MAP),
    date            = as.Date(date),
    contacts_traced = suppressWarnings(as.numeric(cumulative_contacts_traced))
  ) |>
  filter(!is.na(contacts_traced))
```

## Introduction

The Democratic Republic of Congo is experiencing an outbreak of Maladie à Virus Ebola (MVE) with cases reported across health zones in Ituri and Nord-Kivu provinces. This report summarises the epidemiological situation based on data extracted from official INSP DRC situation reports. It is generated automatically and reflects the most recent available data.

## Methods

### Data sources

Epidemiological data were obtained from PDF situation reports published by the Institut National de Santé Publique (INSP) DRC during the 2026 MVE outbreak in Ituri Province and neighbouring zones. The automated fetcher (`fetch_sitreps.py`) scrapes two INSP web pages (<https://insp.cd/ebola/> and <https://insp.cd/category/sitrep/>) for PDF hyperlinks and downloads any document whose URL contains at least one of the tokens *MVE*, *SitRep*, *Ebola*, or *Marburg* and whose WordPress upload path indicates a date of May 2026 or later. A JSON manifest records each downloaded file with its source URL, file size, and download timestamp, ensuring idempotent incremental runs. Eleven PDFs were processed in total, all obtained directly from the INSP DRC website. MD5 verification confirmed that the PDFs downloaded from the INSP website are byte-for-byte identical to those archived in the Kraemer lab repository, establishing that both pipelines operate on the same source documents.

For cross-source validation, pre-harmonised long-format case-count data were obtained from the same Kraemer lab repository (`build/long/`), which represents independent manual human extraction of INSP DRC situation reports into a standardised schema.

### Automated PDF extraction pipeline

Each PDF was passed to the Anthropic Claude claude-sonnet-4-6 large language model via the Messages API as a base64-encoded document, enabling visual (pixel-level) reading of the PDF rather than relying on text-layer extraction. A temperature of 0 was used throughout to maximise response reproducibility.

The extraction pipeline makes two sequential API calls per document. The first call uses a focused structured-extraction prompt (up to 8,192 output tokens) and returns a single JSON object with two keys: `new_cases` (the table of incident cases for the most recent reporting day, *nouvelles données* or equivalent) and `cumulative` (the cumulative summary table, labelled *Tableau 3* or *données cumulatives* in most documents). Each table object carries table-level metadata — the exact table title string, a reporting date, a period end date, a period start date, and any footnote text — as well as an array of row objects. Every row is normalised to eight fields: `province`, `zone_de_sante`, `cas_suspects`, `cas_probables`, `cas_confirmes`, `deces_suspects`, `deces_probables`, and `deces_confirmes`. Empty strings are used for fields absent from the source; genuine zero counts are preserved as `"0"`. Where a numeric field is absent from the table itself but is explicitly and unambiguously stated in the document body text or *Points Saillants* section for the same reporting unit and the same reporting period, the prompt instructs the model to populate that field from the narrative; values are never inferred, estimated, or calculated. The prompt further instructs the model to transpose any pivoted table in which health zones appear as column headers rather than row labels. The second API call (up to 48,000 output tokens) requests a verbatim full-text transcription of the document and is stored in `raw_extraction.json` alongside the structured data, providing a complete audit trail for each PDF. If the second call fails, the pipeline continues without the transcription rather than aborting.

If the Claude response is not valid JSON, the `json_repair` library is applied before parsing.

### Post-extraction validation

Three successive validation checks guard against spurious data being introduced into the output. First, a title-based check examines the `table_title` field of the cumulative table object; if the title contains any of a defined set of French alert-investigation keywords (*alertes*, *reçues*, *validées*, *investiguées*, *enquêtées*, *notifiées*, *investigation*, *suivi des alertes*) and the extracted rows contain no case-count data, the cumulative rows are discarded and a warning is emitted. This check rejects the *Tableau de suivi des alertes* format used in some early Ituri DPS provincial documents. Second, a row-level heuristic discards the cumulative table if, in any extracted row, the probable-case count exceeds three times the suspect-case count — a pattern consistent with an alert-investigation table in which the *probable* column records investigated alerts rather than probable cases. Third, if the cumulative table has been discarded by either check and the new-cases table has no identifiable table title, new-cases rows are also suppressed on the grounds that they were likely synthesised by the model from body text rather than read from a dedicated table.

### Date resolution

The end date assigned to cumulative rows follows a defined priority cascade: (1) a date parsed from the cumulative table title using a French-language date pattern (*DD mois YYYY*); (2) the `reporting_date` field returned by the model; (3) the `period_end_date` field; (4) a date parsed from the table footnotes; (5) a date parsed from the PDF filename, which is attempted first as a compact numeric pattern (*YYYYMMDD*) and then as a French text date (*DD-mois-YYYY*). For new-cases rows, the filename date is used preferentially as the end date, falling back to the model-reported fields. An empty string is assigned only if no date is recoverable by any method.

### Data harmonisation

Zone names are normalised to a canonical spelling prior to analysis: *Mungbwalu* is corrected to *Mongbwalu*. Rows are classified as aggregate (province- or outbreak-level summaries) if the health zone field contains the word *total* or *sous-total*, is empty while a province name is present, or contains multiple zone names separated by commas or parenthetical zone counts (*e.g.* Mongbwalu, Rwampara et Bunia (3 ZS)). Sentinel non-data values (*ND*, *N/A*, dashes, and em-dashes) are treated as missing.

Suspect and probable case counts are combined into a single unconfirmed category, because individual situation reports differ in which of the two fields they populate: earlier Ituri DPS provincial documents use the *probabilités* column for zone-level counts, whereas national INSP documents use *cas suspects*. The unconfirmed count for a row is the sum of suspects and probables, with each coerced to zero when absent but the other is present; when both are absent the unconfirmed value is left as missing rather than set to zero, to prevent false zero observations for zones not reported in a given sitrep. Deaths are derived as the first non-missing value from `deaths_suspected`, `deaths_probable`, and `deaths_confirmed`, in that order of priority. The three-way coalesce is necessary because early Ituri DPS provincial documents record deaths exclusively in the `deces_confirmes` column, while later national INSP documents use `deces_suspects`; the column labelling is not consistent across situation reports.

To enforce temporal consistency, cumulative counts are constrained to be monotonically non-decreasing within each health zone using the running cumulative maximum across successive reporting dates. This operation is applied separately to confirmed cases, unconfirmed cases, and deaths. Where multiple source documents provide data for the same health zone and the same reporting date, the row with the highest unconfirmed count is retained.

### Output structure

Per-PDF outputs (new cases table, cumulative table, combined linelist, and raw JSON extraction) are written to a dedicated subfolder under `outputs/sitreps/`. The canonical folder name is derived from the PDF's entry in a manifest file where available, or from the PDF filename stem otherwise. All processed PDFs are appended to a single master linelist (`outputs/master_combined_counts.csv`) sorted chronologically by reporting date. A registry (`outputs/processed.json`) records each processed PDF with its processing timestamp and row count, enabling incremental updates without reprocessing previously analysed documents.

### Statistical analysis

All statistical analyses and data visualisations were produced in R (version `r R.version$major`.`r R.version$minor`) using the `tidyverse` suite of packages [@tidyverse], with figures produced using `ggplot2` [@ggplot2] and tables formatted with `flextable` [@flextable]. The report was rendered with Quarto. This report and the underlying extraction pipeline were developed with the assistance of GitHub Copilot (Microsoft) and the Claude API (Anthropic, model: claude-sonnet-4-6).

## Results

As of `r latest_date`, the DRC MVE outbreak had recorded a cumulative total of `r prettyNum(n_total, big.mark = ",")` cases (`r prettyNum(n_conf, big.mark = ",")` confirmed, `r prettyNum(n_unconf, big.mark = ",")` suspect or probable) and `r prettyNum(n_deaths, big.mark = ",")` deaths, corresponding to a case fatality ratio of `r cfr_pct`%. Cases have been reported across `r n_zones` health zones in `r n_provinces` provinces.

### Cumulative epidemic curve

```{r}
#| label: fig-cumulative
#| fig-cap: "Cumulative MVE cases and deaths, DRC. Suspect and probable cases are combined. End-of-line labels show the latest recorded value. Source: INSP DRC."
#| fig-width: 8
#| fig-height: 5
p1_dat <- totals |>
  select(date,
         `Suspect/probable cases` = unconfirmed,
         `Confirmed cases`        = confirmed,
         `Outbreak deaths`        = deaths) |>
  pivot_longer(-date, names_to = "category", values_to = "count") |>
  mutate(category = factor(category, levels = names(PAL)))

p1_last <- p1_dat |>
  filter(!is.na(count)) |>
  group_by(category) |>
  slice_max(date, n = 1) |>
  ungroup()

p <- ggplot(p1_dat, aes(x = date, y = count, colour = category, group = category)) +
  geom_line(linewidth = 1.1, na.rm = TRUE) +
  geom_point(size = 2.8, na.rm = TRUE) +
  geom_label_repel(
    data           = p1_last,
    aes(label      = count),
    size           = 3.5,
    nudge_x        = 1.5,
    direction      = "y",
    hjust          = 0,
    segment.colour = "grey70",
    segment.size   = 0.3,
    show.legend    = FALSE,
    na.rm          = TRUE
  ) +
  scale_colour_manual(values = PAL) +
  scale_x_date(
    date_labels = "%d %b",
    date_breaks = "2 days",
    expand      = expansion(add = c(0, 4))
  ) +
  scale_y_continuous(breaks = int_breaks, expand = expansion(mult = c(0, 0.08))) +
  labs(
    title    = "MVE DRC \u2014 Cumulative epidemic curve",
    subtitle = "Total counts across all health zones",
    x = NULL, y = "Cumulative count", colour = NULL,
    caption  = "Source: INSP DRC situation reports"
  ) +
  theme_sitrep(base_size = 12)
ggsave(here("outputs", "plots", "fig-cumulative.png"), plot = p, width = 8, height = 5, dpi = 300)
p
```

### Daily new cases

```{r}
#| label: fig-new-cases
#| fig-cap: "New cases reported per situation report across all health zones, DRC. Source: INSP DRC."
#| fig-width: 7
#| fig-height: 4
p <- daily_new |>
  select(date,
         `Suspect/probable cases` = unconfirmed,
         `Confirmed cases`        = confirmed) |>
  pivot_longer(-date, names_to = "category", values_to = "count") |>
  mutate(category = factor(category, levels = c("Suspect/probable cases", "Confirmed cases"))) |>
  ggplot(aes(x = date, y = count, fill = category)) +
  geom_col(position = "dodge", width = 0.3, na.rm = TRUE) +
  scale_fill_manual(values = PAL) +
  scale_x_date(date_labels = "%d %b", date_breaks = "1 day") +
  scale_y_continuous(breaks = int_breaks, expand = expansion(mult = c(0, 0.1))) +
  labs(
    title    = "MVE DRC \u2014 Daily new cases",
    subtitle = "Total reported per situation report",
    x = NULL, y = "New cases", fill = NULL,
    caption  = "Source: INSP DRC situation reports"
  ) +
  theme_sitrep(base_size = 12)
ggsave(here("outputs", "plots", "fig-new-cases.png"), plot = p, width = 7, height = 4, dpi = 300)
p
```

### Geographic distribution — cumulative counts

```{r}
#| label: fig-zone-cumulative
#| fig-cap: "Cumulative cases by health zone. Y-axis is free across panels. Source: INSP DRC."
#| fig-width: 10
#| fig-height: 11
p <- zone_ts |>
  make_zone_labels() |>
  select(date, zone_label,
         `Suspect/probable cases` = unconfirmed,
         `Confirmed cases`        = confirmed,
         `Outbreak deaths`        = deaths) |>
  pivot_longer(
    cols      = c(`Suspect/probable cases`, `Confirmed cases`, `Outbreak deaths`),
    names_to  = "category",
    values_to = "count"
  ) |>
  mutate(category = factor(category, levels = names(PAL))) |>
  ggplot(aes(x = date, y = count, colour = category, group = category)) +
  geom_line(linewidth = 0.9, na.rm = TRUE) +
  geom_point(size = 2.2, na.rm = TRUE) +
  scale_colour_manual(values = PAL) +
  scale_x_date(date_labels = "%d %b", date_breaks = "1 day") +
  scale_y_continuous(breaks = int_breaks, expand = expansion(mult = c(0, 0.15))) +
  facet_wrap(~ zone_label, scales = "free_y", ncol = 2) +
  labs(
    title    = "MVE DRC \u2014 Cumulative cases by health zone",
    subtitle = "Each panel shows one health zone; y-axis is free.",
    x = NULL, y = "Cumulative count", colour = NULL,
    caption  = "Source: INSP DRC situation reports"
  ) +
  theme_sitrep(base_size = 10) +
  theme(axis.text.x = element_text(size = 8, angle = 45, hjust = 1))
ggsave(here("outputs", "plots", "fig-zone-cumulative.png"), plot = p, width = 10, height = 11, dpi = 300)
p
```

### Geographic distribution — new cases

```{r}
#| label: fig-zone-new
#| fig-cap: "New cases per situation report by health zone. Y-axis is free across panels. Source: INSP DRC."
#| fig-width: 10
#| fig-height: 11
p <- zone_new |>
  make_zone_labels() |>
  select(date, zone_label,
         `Suspect/probable cases` = unconfirmed,
         `Confirmed cases`        = confirmed) |>
  pivot_longer(
    cols      = c(`Suspect/probable cases`, `Confirmed cases`),
    names_to  = "category",
    values_to = "count"
  ) |>
  mutate(category = factor(category, levels = c("Suspect/probable cases", "Confirmed cases"))) |>
  ggplot(aes(x = date, y = count, fill = category)) +
  geom_col(position = "dodge", width = 0.4, na.rm = TRUE) +
  scale_fill_manual(values = PAL) +
  scale_x_date(date_labels = "%d %b", date_breaks = "1 day") +
  scale_y_continuous(breaks = int_breaks, expand = expansion(mult = c(0, 0.15))) +
  facet_wrap(~ zone_label, scales = "free_y", ncol = 2) +
  labs(
    title    = "MVE DRC \u2014 Daily new cases by health zone",
    subtitle = "Per situation report; y-axis is free.",
    x = NULL, y = "New cases", fill = NULL,
    caption  = "Source: INSP DRC situation reports"
  ) +
  theme_sitrep(base_size = 10) +
  theme(axis.text.x = element_text(size = 8, angle = 45, hjust = 1))
ggsave(here("outputs", "plots", "fig-zone-new.png"), plot = p, width = 10, height = 11, dpi = 300)
p
```

### Current counts by health zone

Table 1 presents the most recent cumulative case counts disaggregated by health zone, drawn from the latest available situation report for each zone.

```{r}
#| label: tbl-zones
#| tbl-cap: "Table 1. Latest cumulative case counts by health zone."
zone_latest <- zone_ts |>
  group_by(zone, province) |>
  slice_max(date, n = 1) |>
  ungroup() |>
  arrange(zone) |>
  transmute(
    `Health zone`      = as.character(zone),
    Province           = province,
    `Latest sitrep`    = format(date, "%d %b %Y"),
    `Suspect/probable` = unconfirmed,
    `Confirmed`        = confirmed,
    `Deaths`           = deaths
  )

total_row <- tibble(
  `Health zone`      = "Total",
  Province           = "\u2014",
  `Latest sitrep`    = format(max(totals$date, na.rm = TRUE), "%d %b %Y"),
  `Suspect/probable` = sum(zone_latest$`Suspect/probable`, na.rm = TRUE),
  `Confirmed`        = sum(zone_latest$`Confirmed`,        na.rm = TRUE),
  `Deaths`           = sum(zone_latest$`Deaths`,           na.rm = TRUE)
)

bind_rows(zone_latest, total_row) |>
  flextable() |>
  bold(part = "header") |>
  bold(i = nrow(zone_latest) + 1L) |>
  bg(part = "header", bg = "#1A3A5C") |>
  color(part = "header", color = "white") |>
  bg(i = nrow(zone_latest) + 1L, bg = "#E8EFF7") |>
  hline(i = nrow(zone_latest), part = "body") |>
  align(j = 4:6, align = "right", part = "all") |>
  colformat_int(j = 4:6) |>
  autofit()
```

### Cross-source validation

Cumulative zone-level counts from this pipeline are compared against the [Kraemer lab Ebola DRC 2026](https://github.com/kraemer-lab/Ebola_DRC_2026) repository, which codes INSP DRC situation reports into a standardised long-format by manual human extraction. Both pipelines operate on the same source PDFs (MD5-verified). Zone names differ between sources at two sites (Mongbwalu/Mongbalu and Nyankunde/Nyakunde); counts are matched on the harmonised name. `ND` values in the Kraemer lab series are treated as missing. Any divergence between the two series therefore reflects differences in extraction method — automated LLM-based visual extraction versus human manual coding — rather than differences in the source documents.

```{r}
#| label: fig-comparison
#| fig-cap: "Cross-source comparison of cumulative confirmed cases, suspect/probable cases, and outbreak deaths by health zone. Solid lines: automated extraction (this pipeline); dashed lines: Kraemer lab (manual extraction). Both sources use identical PDFs; divergence reflects extraction method differences."
#| fig-width: 12
#| fig-height: 11
PAL_COMP <- c(
  "Automated extraction" = "#0072B2",
  "Kraemer lab"          = "#E69F00"
)
LTY_COMP <- c(
  "Automated extraction" = "solid",
  "Kraemer lab"          = "dashed"
)

our_for_comp <- zone_ts |>
  transmute(zone, date, confirmed, suspected = unconfirmed, deaths,
            source = "Automated extraction")

comp_long <- bind_rows(our_for_comp, kraemer) |>
  pivot_longer(c(confirmed, suspected, deaths), names_to = "metric", values_to = "count") |>
  mutate(
    metric = recode(metric,
      "confirmed" = "Confirmed cases",
      "suspected" = "Suspect/probable cases",
      "deaths"    = "Outbreak deaths"
    ),
    source = factor(source, levels = c("Automated extraction", "Kraemer lab"))
  ) |>
  filter(!is.na(count))

p <- ggplot(comp_long, aes(x = date, y = count, colour = source, linetype = source)) +
  geom_line(linewidth = 0.9, na.rm = TRUE) +
  geom_point(size = 2.0, na.rm = TRUE) +
  facet_grid(metric ~ zone, scales = "free_y") +
  scale_colour_manual(values = PAL_COMP) +
  scale_linetype_manual(values = LTY_COMP) +
  scale_x_date(date_labels = "%d %b", date_breaks = "2 days") +
  scale_y_continuous(breaks = int_breaks, limits = c(0, NA), expand = expansion(mult = c(0, 0.15))) +
  labs(
    title    = "MVE DRC \u2014 Cross-source validation",
    subtitle = "Cumulative counts: automated extraction vs Kraemer lab (manual extraction, same source PDFs)",
    x = NULL, y = "Cumulative count", colour = NULL, linetype = NULL,
    caption  = "Sources: INSP DRC situation reports; github.com/kraemer-lab/Ebola_DRC_2026"
  ) +
  theme_sitrep(base_size = 9) +
  theme(
    axis.text.x     = element_text(angle = 45, hjust = 1, size = 7),
    panel.border    = element_rect(colour = "grey70", fill = NA, linewidth = 0.4),
    panel.spacing   = unit(0.6, "lines")
  )
ggsave(here("outputs", "plots", "fig-comparison.png"), plot = p, width = 12, height = 11, dpi = 300)
p
```

Table 2 presents numerical differences at the most recent date available in each source for each zone.

```{r}
#| label: tbl-comparison
#| tbl-cap: "Table 2. Cumulative case counts at the latest available date per zone: automated extraction vs Kraemer lab. Delta = ours minus Kraemer. Dashes indicate no data in that source."
ours_latest <- zone_ts |>
  group_by(zone) |>
  slice_max(date, n = 1) |>
  ungroup() |>
  transmute(zone, conf_ours = confirmed, susp_ours = unconfirmed, dead_ours = deaths)

kraemer_latest <- kraemer |>
  group_by(zone) |>
  slice_max(date, n = 1) |>
  ungroup() |>
  transmute(zone, conf_them = confirmed, susp_them = suspected, dead_them = deaths)

full_join(ours_latest, kraemer_latest, by = "zone") |>
  mutate(
    d_conf = conf_ours - conf_them,
    d_susp = susp_ours - susp_them,
    d_dead = dead_ours - dead_them
  ) |>
  arrange(zone) |>
  transmute(
    zone,
    conf_ours, conf_them, d_conf,
    susp_ours, susp_them, d_susp,
    dead_ours, dead_them, d_dead
  ) |>
  setNames(c(
    "Health zone",
    "Confirmed (ours)", "Confirmed (Kraemer)", "\u0394 confirmed",
    "Suspect/prob. (ours)", "Suspect/prob. (Kraemer)", "\u0394 suspect/prob.",
    "Deaths (ours)", "Deaths (Kraemer)", "\u0394 deaths"
  )) |>
  flextable() |>
  bold(part = "header") |>
  bg(part = "header", bg = "#1A3A5C") |>
  color(part = "header", color = "white") |>
  align(j = 2:10, align = "right", part = "all") |>
  colformat_int(j = 2:10, na_str = "\u2014") |>
  autofit()
```

### Response indicators

```{r}
#| label: fig-hospitalised
#| fig-cap: "Patients currently hospitalised by health zone, 20–23 May 2026. Source: Kraemer lab / INSP DRC situation reports."
#| fig-width: 8
#| fig-height: 5
ZONE_COLOURS <- c(
  "Mongbwalu" = "#56B4E9",
  "Rwampara"  = "#E69F00",
  "Bunia"     = "#009E73",
  "Nyankunde" = "#D55E00",
  "Aru"       = "#CC79A7"
)

kraemer_hosp |>
  filter(!is.na(hospitalised)) |>
  mutate(zone = factor(zone, levels = names(ZONE_COLOURS))) |>
  ggplot(aes(x = date, y = hospitalised, colour = zone, group = zone)) +
  geom_line(linewidth = 1.1, na.rm = TRUE) +
  geom_point(size = 2.8, na.rm = TRUE) +
  scale_colour_manual(values = ZONE_COLOURS, na.translate = FALSE) +
  scale_x_date(date_labels = "%d %b", date_breaks = "1 day") +
  scale_y_continuous(breaks = int_breaks, limits = c(0, NA),
                     expand = expansion(mult = c(0, 0.12))) +
  labs(
    title    = "MVE DRC \u2014 Hospitalised patients",
    subtitle = "Patients currently in hospital per reporting day, by health zone",
    x = NULL, y = "Patients hospitalised", colour = NULL,
    caption  = "Source: Kraemer lab / INSP DRC situation reports"
  ) +
  theme_sitrep(base_size = 12)
```

```{r}
#| label: fig-contacts
#| fig-cap: "Cumulative contacts traced by health zone, 14–23 May 2026. ND values (not reported) are excluded. Source: Kraemer lab / INSP DRC situation reports."
#| fig-width: 8
#| fig-height: 5
CONTACTS_PAL <- c(
  "Mongbwalu" = "#56B4E9", "Rwampara"  = "#E69F00",
  "Bunia"     = "#009E73", "Nyankunde" = "#D55E00",
  "Bambu"     = "#88CCEE", "Aru"       = "#CC79A7",
  "Komanda"   = "#0072B2", "Mangala"   = "#999999",
  "Nizi"      = "#117733", "Kilo"      = "#882255"
)

kraemer_contacts |>
  mutate(zone = factor(zone, levels = names(CONTACTS_PAL))) |>
  ggplot(aes(x = date, y = contacts_traced, colour = zone, group = zone)) +
  geom_line(linewidth = 1.1, na.rm = TRUE) +
  geom_point(size = 2.8, na.rm = TRUE) +
  scale_colour_manual(values = CONTACTS_PAL, na.translate = FALSE) +
  scale_x_date(date_labels = "%d %b", date_breaks = "2 days") +
  scale_y_continuous(breaks = int_breaks, limits = c(0, NA),
                     expand = expansion(mult = c(0, 0.12))) +
  labs(
    title    = "MVE DRC \u2014 Cumulative contacts traced",
    subtitle = "Cumulative contacts traced per reporting day, by health zone",
    x = NULL, y = "Contacts traced (cumulative)", colour = NULL,
    caption  = "Source: Kraemer lab / INSP DRC situation reports"
  ) +
  theme_sitrep(base_size = 12)
```

## Discussion

The outbreak trajectory shows a consistent increase in suspect and probable case counts across the reporting period. Confirmed cases remain a minority of the total, reflecting the challenges of laboratory confirmation in the field setting. The geographic spread across both Ituri and Nord-Kivu provinces underscores the need for multi-province coordination of the response.

Per-zone new case data are only available from 20 May 2026 onwards, as earlier situation reports did not disaggregate new cases by health zone. Response indicator data (hospitalisation, contacts traced) are sourced from the Kraemer lab repository as these metrics are not captured in the automated extraction pipeline. Results should be interpreted in the context of reporting delays and the possibility of incomplete situation report coverage for some health zones and dates. The case fatality ratio reported here should be treated as a provisional estimate, as delays in death reporting may affect the numerator, and the deaths column captures the first available of deaths-among-suspects, deaths-among-probables, and deaths-among-confirmed, which are not consistently labelled across situation reports.

MD5 checksum verification confirmed that the PDFs processed by this pipeline are byte-for-byte identical to those archived in the Kraemer lab repository. Divergence in the cross-source comparison therefore reflects either differences in how each extraction method interprets ambiguous table layouts, or outright reading errors in one of the two series.

Four sources of divergence were identified. First, for SitReps N°001 and N°002 (14–15 May), the INSP documents present only an outbreak-wide aggregate without a standardised zone-level cumulative tableau. The N°001 PDF text explicitly states that its 246-case suspect total covers "Zones de Santé (ZS) de Mongbwalu, Rwampara et Bunia" collectively; Kraemer incorrectly attributed this aggregate to Mongbwalu alone. This pipeline extracted the zone-level probable case figures from Tableau 1 in N°002 (57 for Mongbwalu, 27 for Rwampara, 3 for Bunia) and has no zone-specific entry for 14 May, where no zone-level breakdown exists. Second, direct comparison of extracted values against the PDF text reveals reading errors in the Kraemer series for two zone–date combinations: Katwa confirmed cases are recorded as 1 and 0 by Kraemer for 21 and 22 May respectively, whereas both the published PDF text and this pipeline's extraction agree on 2 and 3; Butembo confirmed cases are recorded as 2 by Kraemer for 21 May, whereas the PDF states 1 and this pipeline extracted 1. A persistent one-case difference in Goma suspected cases (0 in this pipeline versus 1 in Kraemer across 20–23 May) could not be attributed to a specific textual feature and may reflect a cell-border ambiguity. Third, this pipeline enforces monotonic non-decrease via cumulative maximum, masking genuine downward revisions visible in the Kraemer series: Mongbwalu confirmed cases fall from 14 (21 May) to 10 (22 May) in the Kraemer data, most likely reflecting reclassification of four previously confirmed cases; this pipeline retains 14 on 22 May before both series converge at 19 on 23 May. Nyankunde confirmed cases similarly fall from 11 to 9 between 21 and 22 May in the Kraemer data; the cummax retains 11 across 22–23 May. Fourth, three of the PDFs processed here carry explicit draft labels in their filenames (`Draft-MVE-Ituri-ActuelB-1`, `Draft_SitRep_MVE_RDC_20260520_vf1`, `Draft_SitRep_MVE_RDC_2022vf2_IM`); the Kraemer lab may hold the corresponding final published versions, and provisional counts in drafts may differ from figures in the final document. This pipeline has additionally processed SitReps N°009 and N°010 (through 25 May), which extend beyond the current Kraemer series, and several health zones present in the Kraemer series (Aru, Kilo, Nizi, Miti-Murhesa) are excluded from this pipeline's zone filter.
